---
title: "Cell size, dilution, and Rac polarity"
subtitle: "Reproducing Buttenschön, Liu & Edelstein-Keshet (2020)"
output:
  html_document:
    toc: true
    toc_float: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = FALSE,
  message = FALSE,
  warning = FALSE,
  fig.align = "center"
)

library(ggplot2)
library(deSolve)

plot_theme <- theme_minimal(base_size = 11) +
  theme(
    legend.position = "bottom",
    panel.grid.minor = element_blank()
  )
```

The exact data used in the plots are exported to `docs/data/` when this report
is rendered inside the project repository.

# Question

How can changing cell size alter a Rac-like polarity system even if the
underlying biochemical reaction rates do not change?

This report reimplements the model in:

> Buttenschön A, Liu Y, Edelstein-Keshet L. **Cell Size, Mechanical Tension,
> and GTPase Signaling in the Single Cell.** *Bulletin of Mathematical
> Biology* 82, 28 (2020). DOI: 10.1007/s11538-020-00702-5.

The paper extends a mass-conserved wave-pinning GTPase model to a deforming
one-dimensional cell. It is unusually relevant to cell-size scaling because
size enters the signaling equations in two distinct ways:

1. the effective diffusion terms scale as \(1/R^2\) after mapping a cell of
   radius \(R(t)\) onto a fixed computational domain; and
2. expansion or contraction introduces an explicit dilution/concentration term
   \(-G\dot R/R\).

The paper then couples local GTPase activity to protrusive Rac-like or
contractile Rho-like forces at the cell ends.

The first two sections below reproduce model variants for which the manuscript
provides sufficiently complete parameter sets. The final section is a
**WGD-oriented extension**, not a result from the paper.

# 1. Published moving-domain GTPase model

## Reaction-diffusion system

The active, slowly diffusing membrane-associated GTPase is \(G\), while
\(G_i\) is the faster-diffusing inactive form. On the physical cell domain,

\[
\frac{\partial G}{\partial t}
=
D\nabla^2G + A(G)G_i - IG,
\]

\[
\frac{\partial G_i}{\partial t}
=
D_i\nabla^2G_i - A(G)G_i + IG,
\]

with

\[
A(G)=\beta+\gamma\frac{G^n}{1+G^n}.
\]

No-flux boundaries conserve GTPase material.

For a changing 1D cell, the authors map the physical interval to the fixed
coordinate \(\bar x\in[-1,1]\). If the cell radius is \(R(t)\), their Eq. (5)
becomes

\[
\frac{\partial G}{\partial t}
=
\frac{D}{R(t)^2}
\frac{\partial^2G}{\partial\bar x^2}
+
A(G)G_i-IG
-
G\frac{\dot R}{R},
\]

\[
\frac{\partial G_i}{\partial t}
=
\frac{D_i}{R(t)^2}
\frac{\partial^2G_i}{\partial\bar x^2}
-
A(G)G_i+IG
-
G_i\frac{\dot R}{R}.
\]

The conserved physical one-dimensional amount is

\[
M
=
R(t)\int_{-1}^{1}(G+G_i)\,d\bar x.
\]

This formulation makes the scaling mechanism explicit. Expansion both changes
the physical diffusion distance and dilutes the local concentrations.

## Numerical implementation

The paper used a cell-centred finite-difference method and a stiff
method-of-lines integrator. Here the same transformed PDE is discretized on a
cell-centred grid and integrated with `deSolve::lsodes`.

```{r model-functions}
laplacian_neumann <- function(u, dx) {
  n <- length(u)
  ans <- numeric(n)

  ans[1] <- (u[2] - u[1]) / dx^2
  ans[n] <- (u[n - 1] - u[n]) / dx^2

  if (n > 2) {
    ans[2:(n - 1)] <-
      (u[1:(n - 2)] - 2 * u[2:(n - 1)] + u[3:n]) / dx^2
  }

  ans
}

paper_chemistry <- list(
  beta = 0.1,
  gamma = 2,
  I = 1,
  n = 4,
  D = 0.01,
  Di = 1,
  L0_um = 55,
  k_s = 1e-3
)

simulate_prescribed_size <- function(
    mode = c("static", "growing", "shrinking"),
    mean_total_initial = 2.5,
    chemistry = paper_chemistry,
    n_cells = 81,
    duration_s = 2000,
    output_dt_s = 5,
    initial_state = NULL) {

  mode <- match.arg(mode)

  dx <- 2 / n_cells
  xbar <- -1 + (seq_len(n_cells) - 0.5) * dx
  R0 <- chemistry$L0_um / 2

  size_sign <- switch(
    mode,
    static = 0,
    growing = 1,
    shrinking = -1
  )

  radius_at <- function(time_s) {
    R0 * exp(size_sign * chemistry$k_s * time_s)
  }

  dilution_at <- function(time_s) {
    size_sign * chemistry$k_s
  }

  if (is.null(initial_state)) {
    # The manuscript does not numerically specify the initial profile used in
    # Fig. 4. We use an edge-biased step while preserving the requested mean
    # total concentration. This is an implementation choice, not a published
    # parameter.
    G0 <- ifelse(xbar <= -0.75, 1.5, 0.1)
    Gi0 <- rep(mean_total_initial - mean(G0), n_cells)

    if (any(Gi0 <= 0)) {
      stop("Initial inactive concentration is non-positive.")
    }

    initial_state <- c(G0, Gi0)
  }

  names(initial_state) <- c(
    sprintf("G_%03d", seq_len(n_cells)),
    sprintf("Gi_%03d", seq_len(n_cells))
  )

  rhs <- function(time_s, state, parms) {
    G <- state[seq_len(n_cells)]
    Gi <- state[n_cells + seq_len(n_cells)]

    R <- radius_at(time_s)
    dilution <- dilution_at(time_s)

    activation <-
      chemistry$beta +
      chemistry$gamma * G^chemistry$n / (1 + G^chemistry$n)

    reaction <- activation * Gi - chemistry$I * G

    dG <-
      chemistry$D / R^2 * laplacian_neumann(G, dx) +
      reaction -
      dilution * G

    dGi <-
      chemistry$Di / R^2 * laplacian_neumann(Gi, dx) -
      reaction -
      dilution * Gi

    list(c(dG, dGi))
  }

  times <- seq(0, duration_s, by = output_dt_s)

  out <- as.data.frame(
    ode(
      y = initial_state,
      times = times,
      func = rhs,
      parms = NULL,
      method = "lsodes",
      rtol = 2e-6,
      atol = 1e-8,
      maxsteps = 100000
    )
  )

  G_cols <- grep("^G_", names(out))
  Gi_cols <- grep("^Gi_", names(out))

  summary <- do.call(
    rbind,
    lapply(seq_len(nrow(out)), function(i) {
      G <- as.numeric(out[i, G_cols])
      Gi <- as.numeric(out[i, Gi_cols])
      R <- radius_at(out$time[i])

      physical_amount <-
        R * sum(G + Gi) * dx

      data.frame(
        time_s = out$time[i],
        radius_um = R,
        length_um = 2 * R,
        mean_total = physical_amount / (2 * R),
        physical_amount = physical_amount,
        active_min = min(G),
        active_max = max(G),
        polarity_score =
          (max(G) - min(G)) / (max(G) + min(G) + 1e-12)
      )
    })
  )

  profiles <- do.call(
    rbind,
    lapply(seq_len(nrow(out)), function(i) {
      G <- as.numeric(out[i, G_cols])
      R <- radius_at(out$time[i])

      data.frame(
        time_s = out$time[i],
        xbar = xbar,
        x_um = xbar * R,
        active_G = G
      )
    })
  )

  list(
    mode = mode,
    xbar = xbar,
    raw = out,
    summary = summary,
    profiles = profiles,
    final_state = as.numeric(out[nrow(out), -1])
  )
}
```

# 2. Reproduction: prescribed growth and shrinkage

Figure 4 of Buttenschön et al. uses:

```{r fig4-parameters}
fig4_parameters <- data.frame(
  quantity = c(
    "Initial cell length",
    "Hill coefficient n",
    "Inactive diffusion Di",
    "Active diffusion D",
    "Autoactivation gamma",
    "Basal activation beta",
    "Inactivation I",
    "Initial total level reported as GT",
    "Growth/shrinkage rate k"
  ),
  value = c(
    "55 um",
    "4",
    "1 um^2/s",
    "0.01 um^2/s",
    "2 /s",
    "0.1 /s",
    "1 /s",
    "2.5",
    "0.001 /s"
  )
)

knitr::kable(fig4_parameters, align = c("l", "l"))
```

There is one notation/reproducibility wrinkle. Earlier in the paper \(G_T\) is
defined as the **conserved total amount**, but the Figure 4 caption says that
\(G_T\) rises as the cell shrinks and falls as it grows. That statement can
only refer to a concentration-like total level rather than the conserved
material amount.

For numerical consistency with the paper's own mass-balance equations, this
report therefore interprets the Figure 4 value `2.5` as the **initial mean
total concentration**. The physical one-dimensional amount is then conserved
as the domain changes.

The numerical initial profile used for Figure 4 is not specified in the
manuscript. We use the same edge-biased profile for all three conditions. This
means the qualitative phenomenon is the appropriate reproduction target; the
exact transition time should not be treated as a pixel-for-pixel recreation of
Figure 4.

```{r prescribed-runs}
prescribed_runs <- lapply(
  c("shrinking", "static", "growing"),
  function(mode) {
    simulate_prescribed_size(
      mode = mode,
      mean_total_initial = 2.5,
      duration_s = 2000,
      output_dt_s = 5
    )
  }
)
names(prescribed_runs) <- c("Shrinking", "Static", "Growing")

prescribed_summary <- do.call(
  rbind,
  lapply(names(prescribed_runs), function(label) {
    x <- prescribed_runs[[label]]$summary
    x$condition <- label
    x
  })
)

prescribed_profiles <- do.call(
  rbind,
  lapply(names(prescribed_runs), function(label) {
    x <- prescribed_runs[[label]]$profiles
    x$condition <- label
    x
  })
)
```

```{r prescribed-kymographs, fig.width=10, fig.height=7}
ggplot(
  prescribed_profiles,
  aes(time_s, xbar, fill = active_G)
) +
  geom_raster() +
  facet_wrap(~ condition, ncol = 1) +
  scale_fill_viridis_c() +
  labs(
    title = "Moving-domain wave-pinning model",
    subtitle = "Active GTPase on the transformed cell coordinate",
    x = "Time (s)",
    y = "Normalized position across cell",
    fill = "Active G"
  ) +
  plot_theme
```

```{r prescribed-summary-plot, fig.width=9, fig.height=6}
prescribed_long <- rbind(
  data.frame(
    prescribed_summary[, c("time_s", "condition")],
    quantity = "Cell length (um)",
    value = prescribed_summary$length_um
  ),
  data.frame(
    prescribed_summary[, c("time_s", "condition")],
    quantity = "Mean total GTPase",
    value = prescribed_summary$mean_total
  ),
  data.frame(
    prescribed_summary[, c("time_s", "condition")],
    quantity = "Polarity score",
    value = prescribed_summary$polarity_score
  )
)

ggplot(
  prescribed_long,
  aes(time_s, value, colour = condition)
) +
  geom_line(linewidth = 0.8) +
  facet_wrap(~ quantity, scales = "free_y", ncol = 1) +
  labs(
    title = "Why changing size eventually changes the signaling state",
    x = "Time (s)",
    y = NULL,
    colour = "Domain"
  ) +
  plot_theme
```

The static cell retains a polarized active-GTPase profile. In the shrinking
cell, the conserved GTPase pool becomes progressively concentrated; in the
growing cell it becomes progressively diluted. In either direction, a large
enough size excursion can move the chemistry out of the wave-pinning regime.

This reproduces the paper's principal qualitative result: **both growth and
shrinkage can destroy polarity, but for different concentration states**.
The authors report an abrupt collapse at approximately 800 s in their Figure 4.
The precise timing in this implementation can differ because the paper does
not provide the numerical initial profile.

The conserved-amount check is:

```{r mass-check}
mass_check <- do.call(
  rbind,
  lapply(names(prescribed_runs), function(label) {
    x <- prescribed_runs[[label]]$summary
    data.frame(
      condition = label,
      relative_mass_error =
        max(abs(x$physical_amount - x$physical_amount[1])) /
        x$physical_amount[1]
    )
  })
)

knitr::kable(mass_check, digits = 3)
```

# 3. Reproduction: dilution-induced Rac/size oscillations

The paper next couples a spatially uniform Rac-like GTPase to protrusive
mechanics. With \(L\) denoting cell length,

\[
\frac{dL}{dt}
=
\frac{-2E(L-L_0)+F_T(G)}{\mu+2\eta},
\]

\[
F_T(G)
=
2F_{\max}
\frac{G^m}{G_c^m+G^m},
\]

and

\[
\frac{dG}{dt}
=
\left(
\beta+\gamma\frac{G^n}{1+G^n}
\right)
\left(
\frac{G_T}{L}-G
\right)
-
IG
-
G\frac{\dot L}{L}.
\]

The last term is again the dilution term. Rac activation pushes the cell
outward; expansion then dilutes Rac, providing a negative feedback around a
locally self-activating signaling system.

For the Rac bifurcation example in Figure 8, the manuscript gives the
dimensionless mechanical combinations directly:

\[
\frac{2E}{\mu+2\eta}=5,
\qquad
\frac{2F_{\max}}{\mu+2\eta}=55.
\]

It reports a Rac limit-cycle regime around \(G_T=65\) and
\(3.2<\gamma<3.6\). We use \(\gamma=3.4\), the midpoint of that reported
interval.

```{r uniform-rac-functions}
uniform_rac_parameters <- list(
  beta = 1,
  gamma = 3.4,
  I = 10,
  n = 8,
  m = 8,
  L0 = 5.5,
  elastic_rate = 5,
  force_rate = 55,
  Gc = 0.85,
  GT = 65
)

simulate_uniform_rac <- function(
    parameters = uniform_rac_parameters,
    duration_s = 100,
    output_dt_s = 0.05,
    initial_G = parameters$GT / parameters$L0) {

  rhs <- function(time_s, state, parms) {
    L <- state[["L"]]
    G <- state[["G"]]

    force_switch <-
      G^parameters$m /
      (parameters$Gc^parameters$m + G^parameters$m)

    dL <-
      -parameters$elastic_rate * (L - parameters$L0) +
      parameters$force_rate * force_switch

    activation <-
      parameters$beta +
      parameters$gamma * G^parameters$n /
      (1 + G^parameters$n)

    dG <-
      activation * (parameters$GT / L - G) -
      parameters$I * G -
      G * dL / L

    list(c(dL, dG))
  }

  times <- seq(0, duration_s, by = output_dt_s)

  as.data.frame(
    ode(
      y = c(L = parameters$L0, G = initial_G),
      times = times,
      func = rhs,
      parms = NULL,
      method = "lsoda",
      rtol = 1e-9,
      atol = 1e-11
    )
  )
}

uniform_rac <- simulate_uniform_rac()
```

```{r uniform-rac-plot, fig.width=9, fig.height=6}
uniform_rac_long <- rbind(
  data.frame(
    time_s = uniform_rac$time,
    quantity = "Cell length",
    value = uniform_rac$L
  ),
  data.frame(
    time_s = uniform_rac$time,
    quantity = "Active Rac-like GTPase",
    value = uniform_rac$G
  )
)

ggplot(uniform_rac_long, aes(time_s, value)) +
  geom_line(linewidth = 0.8) +
  facet_wrap(~ quantity, scales = "free_y", ncol = 1) +
  labs(
    title = "Dilution-induced Rac/size oscillation",
    subtitle = "Figure 8 Rac parameter regime, gamma = 3.4 and GT = 65",
    x = "Time",
    y = NULL
  ) +
  plot_theme
```

After transients, the system enters the behavior described by the paper:
Rac-driven expansion increases cell length, which lowers \(G_T/L\) and dilutes
active Rac; reduced Rac force then allows the elastic cell to retract, which
concentrates the signaling pool again. The result is a self-sustained
mechanochemical oscillation.

This is a stronger size-dependent mechanism than merely changing diffusion
distance. **Cell size feeds back directly onto the concentration of the
signaling machinery.**

# 4. Full spatial mechanochemical model

For completeness, the paper's full single-GTPase model combines the moving
reaction-diffusion system with endpoint mechanics:

\[
\dot x_c=\frac{F_1+F_2}{2\mu},
\]

\[
\dot R=
\frac{
F_2-F_1-2E(2R-L_0)
}{
2(\mu+2\eta)
}.
\]

For protrusive Rac, outward forces are

\[
F_1
=
-F_{\max}
\frac{G(-1)^m}{G_c^m+G(-1)^m},
\qquad
F_2
=
+F_{\max}
\frac{G(1)^m}{G_c^m+G(1)^m}.
\]

The code below implements Eqs. (12a-e) directly.

```{r full-spatial-rac-solver}
simulate_spatial_rac_mechanics <- function(
    chemistry,
    mechanics,
    mean_total_initial,
    n_cells = 81,
    duration_s = 1000,
    output_dt_s = 5,
    edge = c("right", "left")) {

  edge <- match.arg(edge)

  dx <- 2 / n_cells
  xbar <- -1 + (seq_len(n_cells) - 0.5) * dx

  G0 <- rep(0.1, n_cells)
  if (edge == "right") {
    G0[xbar >= 0.75] <- 1.5
  } else {
    G0[xbar <= -0.75] <- 1.5
  }

  Gi0 <- rep(mean_total_initial - mean(G0), n_cells)

  if (any(Gi0 <= 0)) {
    stop("Initial inactive concentration is non-positive.")
  }

  state0 <- c(
    G0,
    Gi0,
    xc = 0,
    R = mechanics$L0_um / 2
  )

  names(state0)[seq_len(n_cells)] <-
    sprintf("G_%03d", seq_len(n_cells))
  names(state0)[n_cells + seq_len(n_cells)] <-
    sprintf("Gi_%03d", seq_len(n_cells))

  rhs <- function(time_s, state, parms) {
    G <- state[seq_len(n_cells)]
    Gi <- state[n_cells + seq_len(n_cells)]
    xc <- state[2 * n_cells + 1]
    R <- state[2 * n_cells + 2]

    if (R <= 0) {
      stop("Cell radius became non-positive.")
    }

    hill_force <- function(g) {
      g^mechanics$m /
        (mechanics$Gc^mechanics$m + g^mechanics$m)
    }

    # Outward protrusive Rac force:
    F1 <- -mechanics$Fmax * hill_force(G[1])
    F2 <-  mechanics$Fmax * hill_force(G[n_cells])

    dxc <- (F1 + F2) / (2 * mechanics$mu)

    dR <-
      (
        F2 - F1 -
        2 * mechanics$E * (2 * R - mechanics$L0_um)
      ) /
      (2 * (mechanics$mu + 2 * mechanics$eta))

    dilution <- dR / R

    activation <-
      chemistry$beta +
      chemistry$gamma * G^chemistry$n /
      (1 + G^chemistry$n)

    reaction <- activation * Gi - chemistry$I * G

    dG <-
      chemistry$D / R^2 * laplacian_neumann(G, dx) +
      reaction -
      dilution * G

    dGi <-
      chemistry$Di / R^2 * laplacian_neumann(Gi, dx) -
      reaction -
      dilution * Gi

    list(c(dG, dGi, dxc, dR))
  }

  times <- seq(0, duration_s, by = output_dt_s)

  as.data.frame(
    ode(
      y = state0,
      times = times,
      func = rhs,
      parms = NULL,
      method = "lsodes",
      rtol = 2e-6,
      atol = 1e-8,
      maxsteps = 100000
    )
  )
}
```

The manuscript reports that a stiff Rac-driven cell can maintain a tight
polarity gradient and move, whereas a softer cell can stretch until the Rac
gradient collapses. However, the Figure 9 caption does **not** provide the full
numerical parameter set needed to recreate those panels uniquely: it identifies
the total GTPase levels and says stiffness \(E\) is varied, but the precise
mechanical parameter combination and numerical initial step are not all
specified.

Accordingly, this report includes the full solver but does not present an
arbitrarily tuned Figure 9 imitation as though it were an exact reproduction.

# 5. WGD-oriented extension: size versus dilution

The paper's dramatic size effects arise partly because the **total amount of
GTPase is conserved while cell size changes**. That is appropriate for an
individual cell undergoing relatively rapid expansion or contraction without
new protein synthesis.

For WGD-associated size differences between established cell states, however,
that assumption is not obviously the correct null. A larger cell may contain
more total Rac/Cdc42 machinery while maintaining an approximately similar
concentration.

We can separate these effects with a deliberately simple counterfactual.

For geometrically similar cells,

\[
L_{2V}=2^{1/3}L_{1V}\approx1.26L_{1V}.
\]

We compare static 1x and 2x-volume analogue domains under two one-dimensional
scaling rules:

- **fixed amount:** total GTPase material is unchanged, so mean concentration
  falls as \(1/L\);
- **concentration-preserving:** total 1D GTPase amount scales with \(L\), so
  mean concentration remains unchanged.

This is explicitly an extension of the paper. It is **not** a full 3D
membrane/cytosol scaling model.

```{r static-size-function}
simulate_static_domain <- function(
    length_um,
    mean_total,
    chemistry = paper_chemistry,
    n_cells = 81,
    duration_s = 1200,
    output_dt_s = 5) {

  chemistry_use <- chemistry
  chemistry_use$L0_um <- length_um

  simulate_prescribed_size(
    mode = "static",
    mean_total_initial = mean_total,
    chemistry = chemistry_use,
    n_cells = n_cells,
    duration_s = duration_s,
    output_dt_s = output_dt_s
  )
}

L1 <- paper_chemistry$L0_um
L2 <- L1 * 2^(1 / 3)
baseline_mean <- 2.5

wgd_conditions <- list(
  `1x | fixed amount` = list(
    length_um = L1,
    mean_total = baseline_mean,
    volume = "1x volume",
    rule = "Fixed amount"
  ),
  `2x | fixed amount` = list(
    length_um = L2,
    mean_total = baseline_mean * L1 / L2,
    volume = "2x volume",
    rule = "Fixed amount"
  ),
  `1x | concentration-preserving` = list(
    length_um = L1,
    mean_total = baseline_mean,
    volume = "1x volume",
    rule = "Concentration-preserving"
  ),
  `2x | concentration-preserving` = list(
    length_um = L2,
    mean_total = baseline_mean,
    volume = "2x volume",
    rule = "Concentration-preserving"
  )
)

wgd_runs <- lapply(
  wgd_conditions,
  function(z) {
    simulate_static_domain(
      length_um = z$length_um,
      mean_total = z$mean_total
    )
  }
)

wgd_final_profiles <- do.call(
  rbind,
  lapply(names(wgd_runs), function(label) {
    run <- wgd_runs[[label]]
    meta <- wgd_conditions[[label]]

    final_time <- max(run$profiles$time_s)
    x <- subset(run$profiles, time_s == final_time)

    data.frame(
      rule = meta$rule,
      volume = meta$volume,
      length_um = meta$length_um,
      mean_total_initial = meta$mean_total,
      x_fraction = (x$xbar + 1) / 2,
      x_um_from_left = (x$xbar + 1) * meta$length_um / 2,
      active_G = x$active_G
    )
  })
)

wgd_summary <- do.call(
  rbind,
  lapply(names(wgd_runs), function(label) {
    run <- wgd_runs[[label]]
    meta <- wgd_conditions[[label]]
    final <- run$summary[nrow(run$summary), ]

    data.frame(
      rule = meta$rule,
      volume = meta$volume,
      length_um = meta$length_um,
      initial_mean_total = meta$mean_total,
      final_active_min = final$active_min,
      final_active_max = final$active_max,
      final_polarity_score = final$polarity_score
    )
  })
)
```

```{r wgd-profiles, fig.width=9, fig.height=5}
ggplot(
  wgd_final_profiles,
  aes(
    x_fraction,
    active_G,
    colour = volume,
    linetype = volume
  )
) +
  geom_line(linewidth = 0.9) +
  facet_wrap(~ rule) +
  labs(
    title = "A 2x-volume analogue: geometry alone versus dilution",
    subtitle = "Same kinetics; static domains differing only in size and GTPase scaling rule",
    x = "Relative position across cell",
    y = "Final active GTPase",
    colour = "Cell size",
    linetype = "Cell size"
  ) +
  plot_theme
```

```{r wgd-summary-table}
knitr::kable(
  wgd_summary,
  digits = 3,
  align = c("l", "l", "r", "r", "r", "r", "r")
)
```

The comparison isolates an important point.

Under **fixed total amount**, making the one-dimensional cell 26% longer lowers
the mean GTPase concentration by about 21%. The polarized profile can therefore
change substantially even though no reaction rate has changed.

Under the **concentration-preserving** rule, much of that difference disappears:
the larger domain experiences a longer diffusion distance, but it no longer
pays the imposed dilution penalty.

So the model supports two separate statements:

1. **Size can matter directly** through transport distance and the
   reaction/diffusion length scale.
2. **Size can matter much more strongly when molecular abundance fails to scale
   with size**, because dilution moves the signaling network through its
   nonlinear state space.

For WGD, distinguishing these mechanisms is essential. A prediction obtained
by holding total Rac/Cdc42 abundance fixed should not automatically be
interpreted as a purely geometric consequence of increased cell size.

# Interpretation for the scaling sandbox

Buttenschön et al. provide a substantially more useful size-scaling example
than a static wave-pinning model because they make the **mass-balance
consequence of changing size explicit**.

The central scaling question is not merely:

> Does a longer cell polarize differently?

It is:

> When cell size changes, which quantities remain fixed -- molecular amount,
> concentration, membrane density, or some combination -- and does that scaling
> rule preserve the polarity regime?

That is directly relevant to a WGD scaling framework. However, the paper does
not answer the WGD question itself.

## Limits of the analogy

1. **The cell is one-dimensional.** It is a transect across a cell, not a 2D
   membrane coupled to a 3D cytosol.
2. **The single-GTPase model is intentionally minimal.** Mammalian front/rear
   polarity involves Rac, Cdc42, Rho, GEFs, GAPs, phosphoinositides,
   cytoskeletal feedback, adhesions, and mechanics.
3. **The moving-domain model assumes cytosol and membrane compartment sizes
   change proportionately in 1D.** The authors explicitly discuss this as a
   simplifying assumption.
4. **The WGD comparison above changes a 1D length by \(2^{1/3}\).** A true WGD
   bulk/surface model would separately represent \(V\), membrane area, and the
   cytosolic reservoir.
5. **No parameters here are fitted to WGD cancer cells.** This is a mechanistic
   scaling demonstration, not a quantitative prediction of a migration
   phenotype.
6. **Figure 9 is not uniquely reproducible from the manuscript alone.** The
   full equations are specified, but not every numerical choice used for those
   spatial/mechanical panels is given.

# Export plot data

```{r export-data, include=FALSE}
data_dir <- "../docs/data"

if (dir.exists("../docs")) {
  dir.create(data_dir, recursive = TRUE, showWarnings = FALSE)

  write.csv(
    prescribed_profiles,
    file.path(data_dir, "buttenschoen_prescribed_size_profiles.csv"),
    row.names = FALSE
  )

  write.csv(
    prescribed_summary,
    file.path(data_dir, "buttenschoen_prescribed_size_summary.csv"),
    row.names = FALSE
  )

  write.csv(
    uniform_rac,
    file.path(data_dir, "buttenschoen_uniform_rac_oscillation.csv"),
    row.names = FALSE
  )

  write.csv(
    wgd_final_profiles,
    file.path(data_dir, "buttenschoen_wgd_scaling_profiles.csv"),
    row.names = FALSE
  )

  write.csv(
    wgd_summary,
    file.path(data_dir, "buttenschoen_wgd_scaling_summary.csv"),
    row.names = FALSE
  )
}
```

# Reproducibility notes

The authors used a cell-centred finite-difference discretization with a stiff
ROWMAP integrator. This report uses the same fixed-domain equations and
cell-centred spatial discretization, but integrates the resulting ODE system
with `deSolve::lsodes`. Numerical trajectories therefore need not match the
authors' implementation to machine precision.

The manuscript is the source of truth for equations and published parameter
sets. Where a numerical choice is not supplied by the manuscript -- most
notably the exact Figure 4 initial profile and the complete Figure 9 parameter
combination -- the report says so rather than silently treating an inferred
choice as published.

# References

Buttenschön A, Liu Y, Edelstein-Keshet L. **Cell Size, Mechanical Tension,
and GTPase Signaling in the Single Cell.** *Bulletin of Mathematical Biology*
82, 28 (2020). DOI: 10.1007/s11538-020-00702-5. Preprint:
arXiv:1908.10840.

Mori Y, Jilkine A, Edelstein-Keshet L. **Wave-Pinning and Cell Polarity from
a Bistable Reaction-Diffusion System.** *Biophysical Journal* 94:3684-3697
(2008). DOI: 10.1529/biophysj.107.120824.

Mori Y, Jilkine A, Edelstein-Keshet L. **Asymptotic and Bifurcation Analysis
of Wave-Pinning in a Reaction-Diffusion Model for Cell Polarization.**
*SIAM Journal on Applied Mathematics* 71:1401-1427 (2011).
DOI: 10.1137/10079118X.
