---
title: "Membrane-to-nucleus signalling: competition between distributed and localized sinks"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
library(ggplot2)
source("../R/scaling.R")
source("../R/diffusion.R")
source("../scenarios/membrane_to_nucleus.R")

scenario <- membrane_to_nucleus_scenario()
volume_colours <- c(`1x volume` = "#0072B2", `2x volume` = "#D55E00")
diffusion_colours <- c(`3` = "#CC79A7", `30` = "#E69F00", `300` = "#009E73")
plot_theme <- theme_minimal(base_size = 11) + theme(legend.position = "bottom")

trapezoid <- function(x, y) {
  sum(diff(x) * (head(y, -1) + tail(y, -1)) / 2)
}

run_delivery_model <- function(volume_multiplier, diffusion_um2_s,
                               pde_multiplier = 1,
                               capture_permeability_um_s =
                                 scenario$boundary$nuclear_capture_permeability_um_s,
                               source_scaling = "fixed flux density",
                               duration_s = scenario$parameters$duration_s,
                               n_cells = 80L) {
  length_scale <- volume_multiplier ^ (1 / 3)
  cell_radius <- scenario$geometry$cell_radius_um * length_scale
  nucleus_radius <- scenario$geometry$nucleus_radius_um * length_scale
  grid <- radial_grid(cell_radius, n_cells, nucleus_radius)
  membrane_flux <- scenario$boundary$membrane_flux_uM_um_s
  if (identical(source_scaling, "total production proportional to volume")) {
    membrane_flux <- membrane_flux * length_scale
  }
  pde_rate <- function(concentration) {
    pde_multiplier * scenario$parameters$pde_vmax_uM_s * concentration /
      (scenario$parameters$pde_km_uM + concentration)
  }
  trajectory <- simulate_radial_diffusion(
    grid = grid,
    diffusion_um2_s = diffusion_um2_s,
    initial = scenario$parameters$initial_concentration_uM,
    duration_s = duration_s,
    reaction = function(time_s, concentration) pde_rate(concentration),
    inner_boundary = list(type = "reactive_sink",
                          value = capture_permeability_um_s),
    outer_boundary = list(type = "fixed_inward_flux", value = membrane_flux),
    save_times = seq(0, duration_s, by = 1)
  )
  concentrations <- trajectory$concentration_uM
  nuclear_boundary <- reactive_inner_boundary(
    concentrations[, 1], grid, diffusion_um2_s, capture_permeability_um_s
  )
  nuclear_area <- 4 * pi * nucleus_radius ^ 2
  membrane_area <- 4 * pi * cell_radius ^ 2
  capture_rate <- nuclear_boundary$flux_uM_um_s * nuclear_area
  pde_total_rate <- rowSums(pde_rate(concentrations) *
                              matrix(grid$volumes_um3, nrow(concentrations),
                                     length(grid$volumes_um3), byrow = TRUE))
  production_rate <- membrane_flux * membrane_area
  produced <- production_rate * duration_s
  captured <- trapezoid(trajectory$time_s, capture_rate)
  hydrolyzed <- trapezoid(trajectory$time_s, pde_total_rate)
  inventory <- as.numeric(concentrations %*% grid$volumes_um3)
  accumulated <- tail(inventory, 1) - inventory[1]
  fractions <- c(
    `Nuclear capture` = captured / produced,
    `PDE hydrolysis` = hydrolyzed / produced,
    `Net accumulation` = accumulated / produced
  )
  list(
    grid = grid,
    trajectory = trajectory,
    summary = data.frame(
      volume_multiplier = volume_multiplier,
      diffusion_um2_s = diffusion_um2_s,
      pde_multiplier = pde_multiplier,
      capture_permeability_um_s = capture_permeability_um_s,
      source_scaling = source_scaling,
      nuclear_capture_fraction = unname(fractions["Nuclear capture"]),
      pde_fraction = unname(fractions["PDE hydrolysis"]),
      accumulation_fraction = unname(fractions["Net accumulation"]),
      mass_balance_error = sum(fractions) - 1,
      cell_radius_um = cell_radius,
      nucleus_radius_um = nucleus_radius,
      membrane_flux_uM_um_s = membrane_flux
    )
  )
}
```

## Model and parameterization

A signal is produced uniformly at the plasma membrane, diffuses through a
spherical cytoplasmic shell, and faces two competing fates:

\[
\frac{\partial C}{\partial t}=D\nabla^2 C-
\frac{V_{\max}C}{K_m+C}.
\]

The outer boundary supplies a fixed inward flux density, (J_{mem}=\beta).
The nuclear envelope is a partially absorbing Robin boundary,
(J_{nuc}=P_{nuc}C(R_{nuc})). Nuclear capture is counted as productive;
cytoplasmic PDE hydrolysis is lost. The main outcome is the integrated nuclear
capture divided by integrated membrane production.

The baseline geometry and cAMP parameters come from published PMVEC models:
(R_{cell}=9.34\) um, (R_{nuc}=5.26\) um, (C_0=0.05\) uM,
(\beta=0.3611\) uM um/s, (V_{max}=0.295\) uM/s, (K_m=2\) uM, and
(D=300,30,3\) um2/s. The membrane flux was derived in that work from AC
activity and the spherical geometry. [Feinstein et al. (2012)](https://pmc.ncbi.nlm.nih.gov/articles/PMC3311237/)
and a later [finite-element reproduction](https://pmc.ncbi.nlm.nih.gov/articles/PMC8773462/)
provide the biological and numerical basis. The latter assumes no transport
between cytoplasm and nucleus, so (P_{nuc}) here is deliberately an abstract
productive-capture parameter, not an empirical cAMP permeability.

The baseline (P_{nuc}=0.03\) um/s is the geometric midpoint of a
0.003--0.3 um/s sensitivity range. Both radii scale with (V^{1/3}); the
1x-to-2x shell thickness therefore changes from 4.08 to 5.14 um.

At low concentration, (k_{eff}=V_{max}/K_m=0.1475\) /s and the survival
length (\sqrt{D/k_{eff}}) is 45.1, 14.3, or 4.51 um. This anticipates weak
transport limitation at high diffusion and a stronger opportunity for sink
competition as diffusion falls.

## Radial signal profiles at 300 seconds

```{r profiles, fig.width=9, fig.height=5}
base_runs <- list()
signal_profiles <- data.frame()
profile_boundaries <- data.frame()
signal_fate <- data.frame()

for (diffusion in scenario$parameters$diffusion_um2_s) {
  for (volume_multiplier in c(1, 2)) {
    key <- paste(diffusion, volume_multiplier, sep = "_")
    result <- run_delivery_model(volume_multiplier, diffusion)
    base_runs[[key]] <- result
    volume_label <- paste0(volume_multiplier, "x volume")
    signal_profiles <- rbind(signal_profiles, data.frame(
      diffusion_um2_s = diffusion,
      volume = volume_label,
      radius_um = result$grid$centers_um,
      concentration_uM = result$trajectory$concentration_uM[
        nrow(result$trajectory$concentration_uM), ]
    ))
    profile_boundaries <- rbind(profile_boundaries,
      data.frame(diffusion_um2_s = diffusion, volume = volume_label,
                 boundary = "Nuclear envelope",
                 radius_um = result$summary$nucleus_radius_um),
      data.frame(diffusion_um2_s = diffusion, volume = volume_label,
                 boundary = "Plasma membrane",
                 radius_um = result$summary$cell_radius_um)
    )
    signal_fate <- rbind(signal_fate,
      data.frame(diffusion_um2_s = diffusion, volume = volume_label,
                 fate = "Nuclear capture",
                 fraction = result$summary$nuclear_capture_fraction),
      data.frame(diffusion_um2_s = diffusion, volume = volume_label,
                 fate = "PDE hydrolysis", fraction = result$summary$pde_fraction),
      data.frame(diffusion_um2_s = diffusion, volume = volume_label,
                 fate = "Net accumulation",
                 fraction = result$summary$accumulation_fraction)
    )
  }
}

ggplot(signal_profiles,
       aes(radius_um, concentration_uM, colour = volume, linetype = volume)) +
  geom_line(linewidth = 0.8) +
  geom_vline(data = subset(profile_boundaries, boundary == "Nuclear envelope"),
             aes(xintercept = radius_um, colour = volume), linetype = "dotted",
             linewidth = 0.45, show.legend = FALSE) +
  geom_vline(data = subset(profile_boundaries, boundary == "Plasma membrane"),
             aes(xintercept = radius_um, colour = volume), linetype = "dashed",
             linewidth = 0.45, show.legend = FALSE) +
  facet_wrap(~ diffusion_um2_s, labeller = label_bquote(D == .(diffusion_um2_s))) +
  scale_colour_manual(values = volume_colours) +
  labs(title = "Membrane-to-nucleus signal profiles at 300 seconds",
       x = "Radius (um)", y = "Signal concentration (uM)", colour = "Cell size",
       linetype = "Cell size",
       caption = "Dotted: nuclear envelope. Dashed: plasma membrane.") +
  plot_theme
```

## Signal fate

```{r fate, fig.width=9, fig.height=5}
signal_fate$fate <- factor(signal_fate$fate,
                           levels = c("Nuclear capture", "PDE hydrolysis",
                                      "Net accumulation"))
ggplot(signal_fate, aes(volume, fraction, fill = fate)) +
  geom_col() +
  facet_wrap(~ diffusion_um2_s, labeller = label_bquote(D == .(diffusion_um2_s))) +
  scale_fill_manual(values = c("#0072B2", "#D55E00", "#999999")) +
  scale_y_continuous(labels = scales::percent_format(accuracy = 1)) +
  labs(title = "Fate of membrane-produced signal over 300 seconds",
       x = "Cell size", y = "Fraction of integrated production", fill = "Fate") +
  plot_theme
```

The three fractions close the mass balance: capture plus hydrolysis plus the
net change in cytoplasmic inventory equals membrane production. The largest
absolute numerical closure error across these runs is
`r format(max(abs(vapply(base_runs, function(x) x$summary$mass_balance_error, numeric(1)))), digits = 2)`.

## Nuclear delivery across cell sizes

```{r scaling, fig.width=8, fig.height=5}
delivery_scaling <- data.frame()
for (diffusion in scenario$parameters$diffusion_um2_s) {
  for (volume_multiplier in c(0.5, 1, 2, 4, 8)) {
    result <- if (volume_multiplier %in% c(1, 2)) {
      base_runs[[paste(diffusion, volume_multiplier, sep = "_")]]
    } else run_delivery_model(volume_multiplier, diffusion)
    delivery_scaling <- rbind(delivery_scaling, data.frame(
      diffusion_um2_s = factor(diffusion), volume_multiplier = volume_multiplier,
      nuclear_capture_fraction = result$summary$nuclear_capture_fraction
    ))
  }
}

ggplot(delivery_scaling,
       aes(volume_multiplier, nuclear_capture_fraction,
           colour = diffusion_um2_s, group = diffusion_um2_s)) +
  geom_line(linewidth = 0.8) + geom_point(size = 2) +
  scale_x_log10(breaks = c(0.5, 1, 2, 4, 8)) +
  scale_y_continuous(labels = scales::percent_format(accuracy = 1)) +
  scale_colour_manual(values = diffusion_colours) +
  labs(title = "Productive nuclear delivery declines with cell volume",
       x = "Cell-volume multiplier", y = "Nuclear capture / membrane production",
       colour = "D (um2/s)") + plot_theme
```

The high-diffusion curve still has a clear size effect. Fast mixing removes
most of the travel-distance penalty, but not the geometric mismatch between a
distributed PDE capacity that scales with cytoplasmic volume and a nuclear
capture boundary that scales with area. Lower diffusion adds a further spatial
penalty rather than creating the entire effect.

## Sensitivity to the abstract nuclear capture parameter

```{r capture-sensitivity, fig.width=8, fig.height=5}
capture_sensitivity <- data.frame()
for (diffusion in scenario$parameters$diffusion_um2_s) {
  for (permeability in scenario$parameters$nuclear_capture_sweep_um_s) {
    one <- run_delivery_model(1, diffusion,
                              capture_permeability_um_s = permeability)
    two <- run_delivery_model(2, diffusion,
                              capture_permeability_um_s = permeability)
    capture_sensitivity <- rbind(capture_sensitivity, data.frame(
      diffusion_um2_s = factor(diffusion),
      capture_permeability_um_s = permeability,
      fraction_1x = one$summary$nuclear_capture_fraction,
      fraction_2x = two$summary$nuclear_capture_fraction,
      log2_size_effect = log2(two$summary$nuclear_capture_fraction /
                               one$summary$nuclear_capture_fraction)
    ))
  }
}

ggplot(capture_sensitivity,
       aes(capture_permeability_um_s, log2_size_effect,
           colour = diffusion_um2_s, group = diffusion_um2_s)) +
  geom_hline(yintercept = 0, colour = "grey60", linetype = "dashed") +
  geom_line(linewidth = 0.8) + geom_point(size = 1.8) +
  scale_x_log10() + scale_colour_manual(values = diffusion_colours) +
  labs(title = "The size effect across nuclear-capture efficiencies",
       x = "Nuclear capture permeability (um/s)",
       y = "log2[nuclear delivery at 2V / delivery at V]",
       colour = "D (um2/s)") + plot_theme
```

## Parameter regimes in which a 2x-volume change matters

```{r regime-heatmap, fig.width=8, fig.height=6}
regime_heatmap <- data.frame()
diffusion_grid <- 10 ^ seq(log10(3), log10(300), length.out = 9)
pde_grid <- 2 ^ seq(-2, 2, length.out = 9)
for (diffusion in diffusion_grid) {
  for (pde_multiplier in pde_grid) {
    one <- run_delivery_model(1, diffusion, pde_multiplier = pde_multiplier)
    two <- run_delivery_model(2, diffusion, pde_multiplier = pde_multiplier)
    regime_heatmap <- rbind(regime_heatmap, data.frame(
      diffusion_um2_s = diffusion, pde_multiplier = pde_multiplier,
      fraction_1x = one$summary$nuclear_capture_fraction,
      fraction_2x = two$summary$nuclear_capture_fraction,
      log2_size_effect = log2(two$summary$nuclear_capture_fraction /
                               one$summary$nuclear_capture_fraction)
    ))
  }
}

ggplot(regime_heatmap,
       aes(diffusion_um2_s, pde_multiplier, fill = log2_size_effect)) +
  geom_tile() + scale_x_log10() + scale_y_log10() +
  scale_fill_gradient2(low = "#3B4CC0", mid = "white", high = "#B40426",
                       midpoint = 0) +
  labs(title = "Size-dependent nuclear delivery is a parameter-regime result",
       x = "Diffusion coefficient (um2/s)",
       y = "PDE activity multiplier",
       fill = "log2(2V / V)") + plot_theme
```

## Mechanistic control: remove the distributed sink

```{r pde-control, fig.width=9, fig.height=5}
pde_zero_control <- data.frame()
for (diffusion in scenario$parameters$diffusion_um2_s) {
  for (volume_multiplier in c(1, 2)) {
    result <- run_delivery_model(volume_multiplier, diffusion, pde_multiplier = 0)
    pde_zero_control <- rbind(pde_zero_control,
      data.frame(diffusion_um2_s = diffusion,
                 volume = paste0(volume_multiplier, "x volume"),
                 fate = "Nuclear capture",
                 fraction = result$summary$nuclear_capture_fraction),
      data.frame(diffusion_um2_s = diffusion,
                 volume = paste0(volume_multiplier, "x volume"),
                 fate = "Net accumulation",
                 fraction = result$summary$accumulation_fraction)
    )
  }
}

ggplot(pde_zero_control, aes(volume, fraction, fill = fate)) +
  geom_col() +
  facet_wrap(~ diffusion_um2_s, labeller = label_bquote(D == .(diffusion_um2_s))) +
  scale_fill_manual(values = c("#0072B2", "#999999")) +
  scale_y_continuous(labels = scales::percent_format(accuracy = 1)) +
  labs(title = "Without PDE, uncaptured signal remains available",
       subtitle = "At 300 seconds, size changes transient capture versus inventory, not irreversible loss",
       x = "Cell size", y = "Fraction of integrated production", fill = "Fate") +
  plot_theme
```

Removing PDE eliminates the competing irreversible fate. Any signal not yet
captured at 300 seconds remains in the cytoplasmic inventory and can be
captured later. This distinguishes persistent sink allocation from a simple
transport delay.

## Source-scaling control

```{r source-control, fig.width=9, fig.height=5}
source_scaling_control <- data.frame()
for (source_scaling in c("fixed flux density",
                         "total production proportional to volume")) {
  for (diffusion in scenario$parameters$diffusion_um2_s) {
    for (volume_multiplier in c(1, 2)) {
      result <- run_delivery_model(volume_multiplier, diffusion,
                                   source_scaling = source_scaling)
      source_scaling_control <- rbind(source_scaling_control, data.frame(
        source_scaling = source_scaling, diffusion_um2_s = factor(diffusion),
        volume = paste0(volume_multiplier, "x volume"),
        nuclear_capture_fraction = result$summary$nuclear_capture_fraction
      ))
    }
  }
}

ggplot(source_scaling_control,
       aes(volume, nuclear_capture_fraction, colour = diffusion_um2_s,
           group = diffusion_um2_s)) +
  geom_line(linewidth = 0.8) + geom_point(size = 2) +
  facet_wrap(~ source_scaling) +
  scale_y_continuous(labels = scales::percent_format(accuracy = 1)) +
  scale_colour_manual(values = diffusion_colours) +
  labs(title = "Source scaling changes loading but not the geometric question",
       x = "Cell size", y = "Nuclear capture / membrane production",
       colour = "D (um2/s)") + plot_theme
```

The primary model holds AC-like flux density fixed, so total production scales
with membrane area. The control increases flux density as (V^{1/3}), making
total production proportional to volume. Because PDE is saturable, this can
change its loading as well as the surface-to-volume relationship.

## Interpretation and limits

Increasing cell size need not merely slow intracellular transport. When a
membrane-generated signal can either be intercepted by a distributed
cytoplasmic sink or reach a localized nuclear target, changing cell scale can
alter the fraction of signalling flux assigned to those two fates.

This is a deliberately minimal radial model. It assumes concentric spherical
geometry, uniform membrane production, uniform PDE, and a phenomenological
nuclear capture boundary. It does not represent cAMP buffering, PDE/AC
nanodomains, nuclear pores, or downstream biochemical amplification. The
capture sweep is therefore more informative than any single fitted-looking
value of (P_{nuc}).

The same geometry could later support an ERK-like first-order model. Classic
[reaction--diffusion theory](https://pubmed.ncbi.nlm.nih.gov/10970807/) found
substantial diffusion control for membrane kinase opposed by cytoplasmic
phosphatase, and live-cell measurements place GFP-ERK1 diffusion near
6--8 um2/s ([live-cell FCS study](https://pmc.ncbi.nlm.nih.gov/articles/PMC2823437/)).
