---
title: "Cell enlargement shifts membrane-generated signal away from nuclear capture"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE)
library(ggplot2)
source("../R/scaling.R")
source("../R/diffusion.R")
source("../scenarios/membrane_to_nucleus.R")

scenario <- membrane_to_nucleus_scenario()
volume_colours <- c(`1x volume` = "#0072B2", `2x volume` = "#D55E00")
diffusion_colours <- c(`3` = "#CC79A7", `30` = "#E69F00", `300` = "#009E73")
plot_theme <- theme_minimal(base_size = 11) + theme(legend.position = "bottom")

trapezoid <- function(x, y) {
  sum(diff(x) * (head(y, -1) + tail(y, -1)) / 2)
}

run_delivery_model <- function(volume_multiplier, diffusion_um2_s,
                               pde_multiplier = 1,
                               capture_permeability_um_s =
                                 scenario$boundary$nuclear_capture_permeability_um_s,
                               source_scaling = "fixed flux density",
                               duration_s = scenario$parameters$duration_s,
                               n_cells = 80L) {
  length_scale <- volume_multiplier ^ (1 / 3)
  cell_radius <- scenario$geometry$cell_radius_um * length_scale
  nucleus_radius <- scenario$geometry$nucleus_radius_um * length_scale
  grid <- radial_grid(cell_radius, n_cells, nucleus_radius)
  membrane_flux <- scenario$boundary$membrane_flux_uM_um_s
  if (identical(source_scaling, "total production proportional to volume")) {
    membrane_flux <- membrane_flux * length_scale
  }
  pde_rate <- function(concentration) {
    pde_multiplier * scenario$parameters$pde_vmax_uM_s * concentration /
      (scenario$parameters$pde_km_uM + concentration)
  }
  trajectory <- simulate_radial_diffusion(
    grid = grid,
    diffusion_um2_s = diffusion_um2_s,
    initial = scenario$parameters$initial_concentration_uM,
    duration_s = duration_s,
    reaction = function(time_s, concentration) pde_rate(concentration),
    inner_boundary = list(type = "reactive_sink",
                          value = capture_permeability_um_s),
    outer_boundary = list(type = "fixed_inward_flux", value = membrane_flux),
    save_times = seq(0, duration_s, by = 1)
  )
  concentrations <- trajectory$concentration_uM
  nuclear_boundary <- reactive_inner_boundary(
    concentrations[, 1], grid, diffusion_um2_s, capture_permeability_um_s
  )
  nuclear_area <- 4 * pi * nucleus_radius ^ 2
  membrane_area <- 4 * pi * cell_radius ^ 2
  capture_rate <- nuclear_boundary$flux_uM_um_s * nuclear_area
  pde_total_rate <- rowSums(pde_rate(concentrations) *
                              matrix(grid$volumes_um3, nrow(concentrations),
                                     length(grid$volumes_um3), byrow = TRUE))
  production_rate <- membrane_flux * membrane_area
  produced <- production_rate * duration_s
  captured <- trapezoid(trajectory$time_s, capture_rate)
  hydrolyzed <- trapezoid(trajectory$time_s, pde_total_rate)
  inventory <- as.numeric(concentrations %*% grid$volumes_um3)
  accumulated <- tail(inventory, 1) - inventory[1]
  fractions <- c(
    `Nuclear capture` = captured / produced,
    `PDE hydrolysis` = hydrolyzed / produced,
    `Net accumulation` = accumulated / produced
  )
  list(
    grid = grid,
    trajectory = trajectory,
    summary = data.frame(
      volume_multiplier = volume_multiplier,
      diffusion_um2_s = diffusion_um2_s,
      pde_multiplier = pde_multiplier,
      capture_permeability_um_s = capture_permeability_um_s,
      source_scaling = source_scaling,
      nuclear_capture_fraction = unname(fractions["Nuclear capture"]),
      pde_fraction = unname(fractions["PDE hydrolysis"]),
      accumulation_fraction = unname(fractions["Net accumulation"]),
      mass_balance_error = sum(fractions) - 1,
      cell_radius_um = cell_radius,
      nucleus_radius_um = nucleus_radius,
      membrane_flux_uM_um_s = membrane_flux
    )
  )
}
```

The exact data used in every plot can be downloaded from the project landing
page, and the complete report and model source is available in the
[repository](https://github.com/Richard-Beck/scaling_sandbox).

## Model and parameterization

A signal is produced uniformly at the plasma membrane, diffuses through a
spherical cytoplasmic shell, and faces two competing fates:

\[
\frac{\partial C}{\partial t}=D\nabla^2 C-
\frac{V_{\max}C}{K_m+C}.
\]

The outer boundary supplies a fixed inward flux density, \(J_{\rm mem}=\beta\).
The nuclear envelope is a partially absorbing Robin boundary,
\(J_{\rm nuc}=P_{\rm nuc}C(R_{\rm nuc})\). Nuclear capture is counted as productive;
cytoplasmic PDE hydrolysis is lost. The main outcome is the integrated nuclear
capture divided by integrated membrane production.

The baseline geometry and cAMP parameters come from published PMVEC models:
\(R_{\rm cell}=9.34\) um, \(R_{\rm nuc}=5.26\) um, \(C_0=0.05\) uM,
\(\beta=0.3611\) uM um/s, \(V_{\max}=0.295\) uM/s, \(K_m=2\) uM, and
\(D=300,30,3\) um2/s. The membrane flux was derived in that work from AC
activity and the spherical geometry. [Feinstein et al. (2012)](https://pmc.ncbi.nlm.nih.gov/articles/PMC3311237/)
and a later [finite-element reproduction](https://pmc.ncbi.nlm.nih.gov/articles/PMC8773462/)
provide the biological and numerical basis. The latter assumes no transport
between cytoplasm and nucleus, so \(P_{\rm nuc}\) here is deliberately an abstract
productive-capture parameter, not an empirical cAMP permeability.

The baseline \(P_{\rm nuc}=0.03\) um/s is the geometric midpoint of a
0.003--0.3 um/s sensitivity range. Both radii scale with \(V^{1/3}\); the
1x-to-2x shell thickness therefore changes from 4.08 to 5.14 um.

At low concentration, \(k_{\rm eff}=V_{\max}/K_m=0.1475\) /s and the survival
length \(\sqrt{D/k_{\rm eff}}\) is 45.1, 14.3, or 4.51 um. This anticipates weak
transport limitation at high diffusion and a stronger opportunity for sink
competition as diffusion falls.

## Radial signal profiles at 300 seconds

```{r profiles, fig.width=9, fig.height=5}
base_runs <- list()
signal_profiles <- data.frame()
profile_boundaries <- data.frame()
signal_fate <- data.frame()

for (diffusion in scenario$parameters$diffusion_um2_s) {
  for (volume_multiplier in c(1, 2)) {
    key <- paste(diffusion, volume_multiplier, sep = "_")
    result <- run_delivery_model(volume_multiplier, diffusion)
    base_runs[[key]] <- result
    volume_label <- paste0(volume_multiplier, "x volume")
    signal_profiles <- rbind(signal_profiles, data.frame(
      diffusion_um2_s = diffusion,
      volume = volume_label,
      radius_um = result$grid$centers_um,
      concentration_uM = result$trajectory$concentration_uM[
        nrow(result$trajectory$concentration_uM), ]
    ))
    profile_boundaries <- rbind(profile_boundaries,
      data.frame(diffusion_um2_s = diffusion, volume = volume_label,
                 boundary = "Nuclear envelope",
                 radius_um = result$summary$nucleus_radius_um),
      data.frame(diffusion_um2_s = diffusion, volume = volume_label,
                 boundary = "Plasma membrane",
                 radius_um = result$summary$cell_radius_um)
    )
    signal_fate <- rbind(signal_fate,
      data.frame(diffusion_um2_s = diffusion, volume = volume_label,
                 fate = "Nuclear capture",
                 fraction = result$summary$nuclear_capture_fraction),
      data.frame(diffusion_um2_s = diffusion, volume = volume_label,
                 fate = "PDE hydrolysis", fraction = result$summary$pde_fraction),
      data.frame(diffusion_um2_s = diffusion, volume = volume_label,
                 fate = "Net accumulation",
                 fraction = result$summary$accumulation_fraction)
    )
  }
}

ggplot(signal_profiles,
       aes(radius_um, concentration_uM, colour = volume, linetype = volume)) +
  geom_line(linewidth = 0.8) +
  geom_vline(data = subset(profile_boundaries, boundary == "Nuclear envelope"),
             aes(xintercept = radius_um, colour = volume), linetype = "dotted",
             linewidth = 0.45, show.legend = FALSE) +
  geom_vline(data = subset(profile_boundaries, boundary == "Plasma membrane"),
             aes(xintercept = radius_um, colour = volume), linetype = "dashed",
             linewidth = 0.45, show.legend = FALSE) +
  facet_wrap(~ diffusion_um2_s, labeller = label_bquote(D == .(diffusion_um2_s))) +
  scale_colour_manual(values = volume_colours) +
  labs(title = "Membrane-to-nucleus signal profiles at 300 seconds",
       x = "Radius (um)", y = "Signal concentration (uM)", colour = "Cell size",
       linetype = "Cell size",
       caption = "Dotted: nuclear envelope. Dashed: plasma membrane.") +
  plot_theme
```

At \(D=300\) um2/s, the 1x profile spans about 1.685--1.688 uM and the 2x
profile 1.164--1.168 uM: both are nearly uniform. At \(D=3\) um2/s, the 1x
profile rises from about 1.559 uM at the nuclear side to 1.851 uM at the
membrane, while the 2x profile rises from 1.013 to 1.363 uM. Low diffusion thus
creates a visible transport gradient that is larger in the enlarged cell. The
high-diffusion result is especially informative: a strong size effect on
nuclear delivery survives even when cytoplasmic concentration is nearly
uniform. Long diffusion distance is therefore not its sole cause; lower
diffusion adds a genuine spatial transport penalty.

## Signal fate

```{r fate, fig.width=9, fig.height=5}
signal_fate$fate <- factor(signal_fate$fate,
                           levels = c("Nuclear capture", "PDE hydrolysis",
                                      "Net accumulation"))
ggplot(signal_fate, aes(volume, fraction, fill = fate)) +
  geom_col() +
  facet_wrap(~ diffusion_um2_s, labeller = label_bquote(D == .(diffusion_um2_s))) +
  scale_fill_manual(values = c("#0072B2", "#D55E00", "#999999")) +
  scale_y_continuous(labels = scales::percent_format(accuracy = 1)) +
  labs(title = "Fate of membrane-produced signal over 300 seconds",
       x = "Cell size", y = "Fraction of integrated production", fill = "Fate") +
  plot_theme
```

The three fractions close the mass balance: capture plus hydrolysis plus the
net change in cytoplasmic inventory equals membrane production. The largest
absolute numerical closure error across these runs is
`r format(max(abs(vapply(base_runs, function(x) x$summary$mass_balance_error, numeric(1)))), digits = 2)`.

At \(D=300\) um2/s, the 1x cell assigns 4.20% of production to nuclear capture,
91.94% to PDE hydrolysis, and 3.86% to net accumulation; the 2x values are
2.94%, 93.74%, and 3.32%. Doubling volume therefore reduces productive capture
by 30.1% relative. The corresponding reductions are 30.5% at \(D=30\) and
34.3% at \(D=3\) um2/s. Most of the size effect already exists in the
well-mixed limit: lowering diffusion by 100-fold increases the penalty only
from about 30% to 34%. In this parameterization, competition between the
area-localized nuclear target and volume-distributed PDE sink dominates, while
finite diffusion strengthens the effect. Mass-balance closure is better than
\(4\times10^{-5}\) in every baseline run.

## Nuclear delivery across cell sizes

```{r scaling, fig.width=8, fig.height=5}
delivery_scaling <- data.frame()
for (diffusion in scenario$parameters$diffusion_um2_s) {
  for (volume_multiplier in c(0.5, 1, 2, 4, 8)) {
    result <- if (volume_multiplier %in% c(1, 2)) {
      base_runs[[paste(diffusion, volume_multiplier, sep = "_")]]
    } else run_delivery_model(volume_multiplier, diffusion)
    delivery_scaling <- rbind(delivery_scaling, data.frame(
      diffusion_um2_s = factor(diffusion), volume_multiplier = volume_multiplier,
      nuclear_capture_fraction = result$summary$nuclear_capture_fraction
    ))
  }
}

ggplot(delivery_scaling,
       aes(volume_multiplier, nuclear_capture_fraction,
           colour = diffusion_um2_s, group = diffusion_um2_s)) +
  geom_line(linewidth = 0.8) + geom_point(size = 2) +
  scale_x_log10(breaks = c(0.5, 1, 2, 4, 8)) +
  scale_y_continuous(labels = scales::percent_format(accuracy = 1)) +
  scale_colour_manual(values = diffusion_colours) +
  labs(title = "Productive nuclear delivery declines with cell volume",
       x = "Cell-volume multiplier", y = "Nuclear capture / membrane production",
       colour = "D (um2/s)") + plot_theme
```

At \(D=300\) um2/s, nuclear capture declines monotonically from 6.26% to 4.20%,
2.94%, 2.12%, and 1.57% as volume increases from 0.5x through 1x, 2x, 4x, and
8x. The 8x cell therefore captures about 63% less signal per unit membrane
production than the 1x cell. At \(D=3\) um2/s, the 1x-to-8x decline is from
3.87% to 1.05%, about 73%. The effect is not peculiar to the 1x/2x comparison;
slower diffusion progressively steepens the scaling at large sizes.

## Sensitivity to the abstract nuclear capture parameter

```{r capture-sensitivity, fig.width=8, fig.height=5}
capture_sensitivity <- data.frame()
for (diffusion in scenario$parameters$diffusion_um2_s) {
  for (permeability in scenario$parameters$nuclear_capture_sweep_um_s) {
    one <- run_delivery_model(1, diffusion,
                              capture_permeability_um_s = permeability)
    two <- run_delivery_model(2, diffusion,
                              capture_permeability_um_s = permeability)
    capture_sensitivity <- rbind(capture_sensitivity, data.frame(
      diffusion_um2_s = factor(diffusion),
      capture_permeability_um_s = permeability,
      fraction_1x = one$summary$nuclear_capture_fraction,
      fraction_2x = two$summary$nuclear_capture_fraction,
      log2_size_effect = log2(two$summary$nuclear_capture_fraction /
                               one$summary$nuclear_capture_fraction)
    ))
  }
}

ggplot(capture_sensitivity,
       aes(capture_permeability_um_s, log2_size_effect,
           colour = diffusion_um2_s, group = diffusion_um2_s)) +
  geom_hline(yintercept = 0, colour = "grey60", linetype = "dashed") +
  geom_line(linewidth = 0.8) + geom_point(size = 1.8) +
  scale_x_log10() + scale_colour_manual(values = diffusion_colours) +
  labs(title = "The size effect across nuclear-capture efficiencies",
       x = "Nuclear capture permeability (um/s)",
       y = "log2[nuclear delivery at 2V / delivery at V]",
       colour = "D (um2/s)") + plot_theme
```

Across the full \(P_{\rm nuc}=0.003\)--0.3 um/s sweep, the 2x cell always has a
lower delivery fraction. At \(D=300\) um2/s, the penalty ranges from about 32%
at weak capture to 20% at strong capture; at \(D=3\), it ranges from about 36%
to 27%. Increasing capture efficiency raises absolute productive delivery and
partially buffers the size effect, but does not reverse it in the explored
range.

## Parameter regimes in which a 2x-volume change matters

```{r regime-heatmap, fig.width=8, fig.height=6}
regime_heatmap <- data.frame()
diffusion_grid <- 10 ^ seq(log10(3), log10(300), length.out = 9)
pde_grid <- 2 ^ seq(-2, 2, length.out = 9)
for (diffusion in diffusion_grid) {
  for (pde_multiplier in pde_grid) {
    one <- run_delivery_model(1, diffusion, pde_multiplier = pde_multiplier)
    two <- run_delivery_model(2, diffusion, pde_multiplier = pde_multiplier)
    regime_heatmap <- rbind(regime_heatmap, data.frame(
      diffusion_um2_s = diffusion, pde_multiplier = pde_multiplier,
      fraction_1x = one$summary$nuclear_capture_fraction,
      fraction_2x = two$summary$nuclear_capture_fraction,
      log2_size_effect = log2(two$summary$nuclear_capture_fraction /
                               one$summary$nuclear_capture_fraction)
    ))
  }
}

ggplot(regime_heatmap,
       aes(diffusion_um2_s, pde_multiplier, fill = log2_size_effect)) +
  geom_tile() + scale_x_log10() + scale_y_log10() +
  scale_fill_gradient2(low = "#3B4CC0", mid = "white", high = "#B40426",
                       midpoint = 0) +
  labs(title = "Size-dependent nuclear delivery is a parameter-regime result",
       x = "Diffusion coefficient (um2/s)",
       y = "PDE activity multiplier",
       fill = "log2(2V / V)") + plot_theme
```

All 81 combinations of \(D=3\)--300 um2/s and PDE multiplier 0.25--4 have a
negative \(\log_2(2V/V)\). The 2x capture fraction is 61--77% of the 1x value,
with no sign reversal in the grid. This supports robustness to the tested
diffusion and PDE activity region, not universality beyond it.

## Mechanistic control: remove the distributed sink

```{r pde-control, fig.width=9, fig.height=5}
pde_zero_control <- data.frame()
for (diffusion in scenario$parameters$diffusion_um2_s) {
  for (volume_multiplier in c(1, 2)) {
    result <- run_delivery_model(volume_multiplier, diffusion, pde_multiplier = 0)
    pde_zero_control <- rbind(pde_zero_control,
      data.frame(diffusion_um2_s = diffusion,
                 volume = paste0(volume_multiplier, "x volume"),
                 fate = "Nuclear capture",
                 fraction = result$summary$nuclear_capture_fraction),
      data.frame(diffusion_um2_s = diffusion,
                 volume = paste0(volume_multiplier, "x volume"),
                 fate = "Net accumulation",
                 fraction = result$summary$accumulation_fraction)
    )
  }
}

ggplot(pde_zero_control, aes(volume, fraction, fill = fate)) +
  geom_col() +
  facet_wrap(~ diffusion_um2_s, labeller = label_bquote(D == .(diffusion_um2_s))) +
  scale_fill_manual(values = c("#0072B2", "#999999")) +
  scale_y_continuous(labels = scales::percent_format(accuracy = 1)) +
  labs(title = "Without PDE, uncaptured signal remains intracellular",
       subtitle = "At 300 seconds, size changes capture versus inventory rather than irreversible loss",
       x = "Cell size", y = "Fraction of integrated production", fill = "Fate") +
  plot_theme
```

At \(D=300\) um2/s, nuclear capture by 300 s is 39.8% for 1x and 33.7% for 2x;
at \(D=3\), it is 39.3% and 33.1%. Removing PDE therefore does not eliminate
the finite-time size effect: the larger cell captures about 15--16% less by
300 s. The remainder is retained as cytoplasmic inventory rather than
hydrolyzed. With PDE present, enlargement changes signal fate as well as
delivery timing. The current 300-s control does not establish the subsequent
trajectory of that retained inventory.

## Source-scaling control

```{r source-control, fig.width=9, fig.height=5}
source_scaling_control <- data.frame()
for (source_scaling in c("fixed flux density",
                         "total production proportional to volume")) {
  for (diffusion in scenario$parameters$diffusion_um2_s) {
    for (volume_multiplier in c(1, 2)) {
      result <- run_delivery_model(volume_multiplier, diffusion,
                                   source_scaling = source_scaling)
      source_scaling_control <- rbind(source_scaling_control, data.frame(
        source_scaling = source_scaling, diffusion_um2_s = factor(diffusion),
        volume = paste0(volume_multiplier, "x volume"),
        nuclear_capture_fraction = result$summary$nuclear_capture_fraction
      ))
    }
  }
}

ggplot(source_scaling_control,
       aes(volume, nuclear_capture_fraction, colour = diffusion_um2_s,
           group = diffusion_um2_s)) +
  geom_line(linewidth = 0.8) + geom_point(size = 2) +
  facet_wrap(~ source_scaling) +
  scale_y_continuous(labels = scales::percent_format(accuracy = 1)) +
  scale_colour_manual(values = diffusion_colours) +
  labs(title = "Source scaling changes the magnitude of the size effect",
       x = "Cell size", y = "Nuclear capture / membrane production",
       colour = "D (um2/s)") + plot_theme
```

The primary model holds AC-like flux density fixed, so total production scales
with membrane area. The control increases flux density as \(V^{1/3}\), making
total production proportional to volume. With fixed flux density, the 2x
capture penalty is 30.1% at \(D=300\) and 34.3% at \(D=3\) um2/s. When total
production is proportional to volume, those penalties are 19.5% and 22.9%.
Source scaling substantially attenuates but does not eliminate the effect; its
magnitude depends materially on how membrane production scales.

## Interpretation and limits

Together, the simulations separate two mechanisms. Enlargement produces a
modest finite-time transport/capture disadvantage even without PDE, but the
larger effect arises when a target scaling with nuclear surface area competes
with an irreversible sink distributed throughout cytoplasmic volume. Doubling
volume reduces nuclear capture by about 30% even at \(D=300\) um2/s, where
spatial gradients are negligible; lowering diffusion raises that penalty to
about 34%. The effect remains negative throughout the tested capture, PDE, and
diffusion ranges and persists, although weakened, when production scales with
volume. The model therefore demonstrates a plausible mechanism by which cell
enlargement changes signalling allocation rather than merely diffusion time.

This is a deliberately minimal radial model. It assumes concentric spherical
geometry, uniform membrane production, uniform PDE, and a phenomenological
nuclear capture boundary. It does not represent cAMP buffering, PDE/AC
nanodomains, nuclear pores, or downstream biochemical amplification. The
capture sweep is therefore more informative than any single fitted-looking
value of \(P_{\rm nuc}\).

The same geometry could later support an ERK-like first-order model. Classic
[reaction--diffusion theory](https://pubmed.ncbi.nlm.nih.gov/10970807/) found
substantial diffusion control for membrane kinase opposed by cytoplasmic
phosphatase, and live-cell measurements place GFP-ERK1 diffusion near
6--8 um2/s ([live-cell FCS study](https://pmc.ncbi.nlm.nih.gov/articles/PMC2823437/)).
