---
title: "Cell enlargement shifts membrane-generated signal away from nuclear capture"
subtitle: "Published cAMP transport baseline with an abstract nuclear-capture extension"
output:
  html_document:
    toc: true
    toc_float: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE)
library(ggplot2)
source("../R/scaling.R")
source("../R/diffusion.R")
source("../scenarios/membrane_to_nucleus.R")

scenario <- membrane_to_nucleus_scenario()
volume_colours <- c(`1x volume` = "#0072B2", `2x volume` = "#D55E00")
diffusion_colours <- c(`3` = "#CC79A7", `30` = "#E69F00", `300` = "#009E73")
plot_theme <- theme_minimal(base_size = 11) + theme(legend.position = "bottom")

trapezoid <- function(x, y) {
  sum(diff(x) * (head(y, -1) + tail(y, -1)) / 2)
}

cumulative_trapezoid <- function(x, y) {
  c(0, cumsum(diff(x) * (head(y, -1) + tail(y, -1)) / 2))
}

run_delivery_model <- function(volume_multiplier, diffusion_um2_s,
                               pde_multiplier = 1,
                               capture_permeability_um_s =
                                 scenario$boundary$nuclear_capture_permeability_um_s,
                               source_scaling = "constant production per unit volume",
                               duration_s = scenario$parameters$duration_s,
                               save_interval_s = 1,
                               n_cells = 80L) {
  length_scale <- volume_multiplier ^ (1 / 3)
  cell_radius <- scenario$geometry$cell_radius_um * length_scale
  nucleus_radius <- scenario$geometry$nucleus_radius_um * length_scale
  grid <- radial_grid(cell_radius, n_cells, nucleus_radius)
  membrane_flux <- scenario$boundary$membrane_flux_uM_um_s
  if (identical(source_scaling, "constant production per unit volume")) {
    membrane_flux <- membrane_flux * length_scale
  }
  pde_rate <- function(concentration) {
    pde_multiplier * scenario$parameters$pde_vmax_uM_s * concentration /
      (scenario$parameters$pde_km_uM + concentration)
  }
  trajectory <- simulate_radial_diffusion(
    grid = grid,
    diffusion_um2_s = diffusion_um2_s,
    initial = scenario$parameters$initial_concentration_uM,
    duration_s = duration_s,
    reaction = function(time_s, concentration) pde_rate(concentration),
    inner_boundary = list(type = "reactive_sink",
                          value = capture_permeability_um_s),
    outer_boundary = list(type = "fixed_inward_flux", value = membrane_flux),
    save_times = unique(c(seq(0, duration_s, by = save_interval_s), duration_s))
  )
  concentrations <- trajectory$concentration_uM
  nuclear_boundary <- reactive_inner_boundary(
    concentrations[, 1], grid, diffusion_um2_s, capture_permeability_um_s
  )
  nuclear_area <- 4 * pi * nucleus_radius ^ 2
  membrane_area <- 4 * pi * cell_radius ^ 2
  capture_rate <- nuclear_boundary$flux_uM_um_s * nuclear_area
  pde_total_rate <- rowSums(pde_rate(concentrations) *
                              matrix(grid$volumes_um3, nrow(concentrations),
                                     length(grid$volumes_um3), byrow = TRUE))
  production_rate <- membrane_flux * membrane_area
  produced <- production_rate * duration_s
  cumulative_produced <- production_rate * trajectory$time_s
  cumulative_captured <- cumulative_trapezoid(trajectory$time_s, capture_rate)
  cumulative_hydrolyzed <- cumulative_trapezoid(trajectory$time_s, pde_total_rate)
  captured <- tail(cumulative_captured, 1)
  hydrolyzed <- tail(cumulative_hydrolyzed, 1)
  inventory <- as.numeric(concentrations %*% grid$volumes_um3)
  accumulated <- tail(inventory, 1) - inventory[1]
  fractions <- c(
    `Nuclear capture` = captured / produced,
    `PDE hydrolysis` = hydrolyzed / produced,
    `Net accumulation` = accumulated / produced
  )
  list(
    grid = grid,
    trajectory = trajectory,
    fate_trajectory = data.frame(
      time_s = trajectory$time_s,
      cumulative_nuclear_capture_fraction = ifelse(
        cumulative_produced > 0, cumulative_captured / cumulative_produced, NA_real_
      ),
      cumulative_pde_fraction = ifelse(
        cumulative_produced > 0, cumulative_hydrolyzed / cumulative_produced, NA_real_
      ),
      cumulative_accumulation_fraction = ifelse(
        cumulative_produced > 0, (inventory - inventory[1]) / cumulative_produced, NA_real_
      ),
      instantaneous_nuclear_capture_fraction = capture_rate / production_rate
    ),
    summary = data.frame(
      volume_multiplier = volume_multiplier,
      diffusion_um2_s = diffusion_um2_s,
      pde_multiplier = pde_multiplier,
      capture_permeability_um_s = capture_permeability_um_s,
      source_scaling = source_scaling,
      nuclear_capture_fraction = unname(fractions["Nuclear capture"]),
      pde_fraction = unname(fractions["PDE hydrolysis"]),
      accumulation_fraction = unname(fractions["Net accumulation"]),
      mass_balance_error = sum(fractions) - 1,
      cell_radius_um = cell_radius,
      nucleus_radius_um = nucleus_radius,
      membrane_flux_uM_um_s = membrane_flux
    )
  )
}
```

[&larr; Report catalogue](../index.html)

> **Status: literature-derived baseline + abstract extension.** Published
> PMVEC models motivate the geometry, cAMP transport, membrane production, and
> PDE parameters. The partially absorbing nuclear target and its size sweep
> are deliberate abstractions introduced here, not a reproduction of a
> published nuclear-capture model.

The exact data used in every plot can be downloaded from the project landing
page, and the complete report and model source is available in the
[repository](https://github.com/Richard-Beck/scaling_sandbox).

## Model and parameterization

A signal is produced uniformly at the plasma membrane, diffuses through a
spherical cytoplasmic shell, and faces two competing fates:

\[
\frac{\partial C}{\partial t}=D\nabla^2 C-
\frac{V_{\max}C}{K_m+C}.
\]

The default analysis holds total membrane production constant per unit cell
volume. Because membrane area scales as \(V^{2/3}\), its inward flux density is
\(J_{\rm mem}=\beta V^{1/3}\), where \(\beta\) is the 1x reference flux. Total
production therefore scales in direct proportion to cell volume.
In the reaction term, \(V_{\max}\) is the local maximum PDE hydrolysis rate per
unit cytoplasmic volume (uM/s); the realized local rate is
\(V_{\max}C/(K_m+C)\).
The nuclear envelope is a partially absorbing Robin boundary,
\(J_{\rm nuc}=P_{\rm nuc}C(R_{\rm nuc})\). Nuclear capture is counted as productive;
cytoplasmic PDE hydrolysis is lost. The main outcome is the integrated nuclear
capture divided by integrated membrane production.

The baseline geometry and cAMP parameters come from published PMVEC models:
\(R_{\rm cell}=9.34\) um, \(R_{\rm nuc}=5.26\) um, \(C_0=0.05\) uM,
\(\beta=0.3611\) uM um/s, \(V_{\max}=0.295\) uM/s, \(K_m=2\) uM, and
\(D=300,30,3\) um2/s. The membrane flux was derived in that work from AC
activity and the spherical geometry. [Feinstein et al. (2012)](https://pmc.ncbi.nlm.nih.gov/articles/PMC3311237/)
and a later [finite-element reproduction](https://pmc.ncbi.nlm.nih.gov/articles/PMC8773462/)
provide the biological and numerical basis. The latter assumes no transport
between cytoplasm and nucleus, so \(P_{\rm nuc}\) here is deliberately an abstract
productive-capture parameter, not an empirical cAMP permeability.

The baseline \(P_{\rm nuc}=0.03\) um/s is the geometric midpoint of a
0.003--0.3 um/s sensitivity range. Both radii scale with \(V^{1/3}\); the
1x-to-2x shell thickness therefore changes from 4.08 to 5.14 um.

At low concentration, \(k_{\rm eff}=V_{\max}/K_m=0.1475\) /s and the survival
length \(\sqrt{D/k_{\rm eff}}\) is 45.1, 14.3, or 4.51 um. This anticipates weak
transport limitation at high diffusion and a stronger opportunity for sink
competition as diffusion falls.

## Radial signal profiles at 300 seconds

```{r profiles, fig.width=9, fig.height=5}
base_runs <- list()
signal_profiles <- data.frame()
profile_boundaries <- data.frame()
signal_fate <- data.frame()
signal_fate_time <- data.frame()

for (diffusion in scenario$parameters$diffusion_um2_s) {
  for (volume_multiplier in c(1, 2)) {
    key <- paste(diffusion, volume_multiplier, sep = "_")
    result <- run_delivery_model(volume_multiplier, diffusion)
    base_runs[[key]] <- result
    volume_label <- paste0(volume_multiplier, "x volume")
    signal_profiles <- rbind(signal_profiles, data.frame(
      diffusion_um2_s = diffusion,
      volume = volume_label,
      radius_um = result$grid$centers_um,
      concentration_uM = result$trajectory$concentration_uM[
        nrow(result$trajectory$concentration_uM), ]
    ))
    profile_boundaries <- rbind(profile_boundaries,
      data.frame(diffusion_um2_s = diffusion, volume = volume_label,
                 boundary = "Nuclear envelope",
                 radius_um = result$summary$nucleus_radius_um),
      data.frame(diffusion_um2_s = diffusion, volume = volume_label,
                 boundary = "Plasma membrane",
                 radius_um = result$summary$cell_radius_um)
    )
    signal_fate <- rbind(signal_fate,
      data.frame(diffusion_um2_s = diffusion, volume = volume_label,
                 fate = "Nuclear capture",
                 fraction = result$summary$nuclear_capture_fraction),
      data.frame(diffusion_um2_s = diffusion, volume = volume_label,
                 fate = "PDE hydrolysis", fraction = result$summary$pde_fraction),
      data.frame(diffusion_um2_s = diffusion, volume = volume_label,
                 fate = "Net accumulation",
                 fraction = result$summary$accumulation_fraction)
    )
    signal_fate_time <- rbind(
      signal_fate_time,
      data.frame(diffusion_um2_s = diffusion, volume = volume_label,
                 time_s = result$fate_trajectory$time_s,
                 fate = "Nuclear capture",
                 fraction = result$fate_trajectory$cumulative_nuclear_capture_fraction),
      data.frame(diffusion_um2_s = diffusion, volume = volume_label,
                 time_s = result$fate_trajectory$time_s,
                 fate = "PDE hydrolysis",
                 fraction = result$fate_trajectory$cumulative_pde_fraction),
      data.frame(diffusion_um2_s = diffusion, volume = volume_label,
                 time_s = result$fate_trajectory$time_s,
                 fate = "Net accumulation",
                 fraction = result$fate_trajectory$cumulative_accumulation_fraction)
    )
  }
}

ggplot(signal_profiles,
       aes(radius_um, concentration_uM, colour = volume, linetype = volume)) +
  geom_line(linewidth = 0.8) +
  geom_vline(data = subset(profile_boundaries, boundary == "Nuclear envelope"),
             aes(xintercept = radius_um, colour = volume), linetype = "dotted",
             linewidth = 0.45, show.legend = FALSE) +
  geom_vline(data = subset(profile_boundaries, boundary == "Plasma membrane"),
             aes(xintercept = radius_um, colour = volume), linetype = "dashed",
             linewidth = 0.45, show.legend = FALSE) +
  facet_wrap(~ diffusion_um2_s, labeller = label_bquote(D == .(diffusion_um2_s))) +
  scale_colour_manual(values = volume_colours) +
  labs(title = "Membrane-to-nucleus signal profiles at 300 seconds",
       x = "Radius (um)", y = "Signal concentration (uM)", colour = "Cell size",
       linetype = "Cell size",
       caption = "Dotted: nuclear envelope. Dashed: plasma membrane.") +
  plot_theme
```

At \(D=300\) um2/s, the 1x profile spans about 1.685--1.688 uM and the 2x
profile 1.712--1.717 uM: both are nearly uniform. At \(D=3\) um2/s, the 1x
profile rises from about 1.559 uM at the nuclear side to 1.851 uM at the
membrane, while the 2x profile rises from 1.519 to 1.970 uM. Low diffusion thus
creates a visible transport gradient that is larger in the enlarged cell. The
high-diffusion result is especially informative: a strong size effect on
nuclear delivery survives even when cytoplasmic concentration is nearly
uniform. Long diffusion distance is therefore not its sole cause; lower
diffusion adds a genuine spatial transport penalty.

## Signal fate

```{r fate, fig.width=9, fig.height=8}
signal_fate_time$fate <- factor(
  signal_fate_time$fate,
  levels = c("Nuclear capture", "PDE hydrolysis", "Net accumulation")
)
ggplot(subset(signal_fate_time, time_s > 0),
       aes(time_s, fraction, colour = volume, linetype = volume)) +
  geom_line(linewidth = 0.8) +
  facet_grid(fate ~ diffusion_um2_s, scales = "free_y",
             labeller = labeller(diffusion_um2_s = function(x) {
               paste0("D = ", x, " um2/s")
             })) +
  scale_colour_manual(values = volume_colours) +
  scale_y_continuous(labels = scales::percent_format(accuracy = 1)) +
  labs(title = "Cumulative fate of membrane-produced signal",
       subtitle = "Each fate has its own vertical scale so nuclear capture remains visible",
       x = "Time (s)", y = "Fraction of cumulative production",
       colour = "Cell size", linetype = "Cell size") +
  plot_theme
```

The three fractions close the mass balance: capture plus hydrolysis plus the
net change in cytoplasmic inventory equals membrane production. The largest
absolute numerical closure error across these runs is
`r format(max(abs(vapply(base_runs, function(x) x$summary$mass_balance_error, numeric(1)))), digits = 2)`.
The separate row scales reveal the otherwise small productive-capture signal:
early production initially enters cytoplasmic inventory, PDE hydrolysis becomes
dominant over time, and cumulative nuclear capture gradually approaches its
300-s value.

At \(D=300\) um2/s, the 1x cell assigns 4.20% of production to nuclear capture,
91.94% to PDE hydrolysis, and 3.86% to net accumulation; the 2x values are
3.38%, 92.69%, and 3.93%. Doubling volume therefore reduces productive capture
by 19.5% relative. The corresponding reductions are 19.8% at \(D=30\) and
22.9% at \(D=3\) um2/s. Most of the size effect already exists in the
well-mixed limit: lowering diffusion by 100-fold increases the penalty only
from about 19% to 23%. In this parameterization, competition between the
area-localized nuclear target and volume-distributed PDE sink dominates, while
finite diffusion strengthens the effect. Mass-balance closure is better than
\(4\times10^{-5}\) in every baseline run.

## Nuclear delivery across cell sizes

```{r scaling, fig.width=8, fig.height=5}
delivery_scaling <- data.frame()
for (diffusion in scenario$parameters$diffusion_um2_s) {
  for (volume_multiplier in c(0.5, 1, 2, 4, 8)) {
    result <- if (volume_multiplier %in% c(1, 2)) {
      base_runs[[paste(diffusion, volume_multiplier, sep = "_")]]
    } else run_delivery_model(volume_multiplier, diffusion)
    delivery_scaling <- rbind(delivery_scaling, data.frame(
      diffusion_um2_s = factor(diffusion), volume_multiplier = volume_multiplier,
      nuclear_capture_fraction = result$summary$nuclear_capture_fraction
    ))
  }
}

ggplot(delivery_scaling,
       aes(volume_multiplier, nuclear_capture_fraction,
           colour = diffusion_um2_s, group = diffusion_um2_s)) +
  geom_line(linewidth = 0.8) + geom_point(size = 2) +
  scale_x_log10(breaks = c(0.5, 1, 2, 4, 8)) +
  scale_y_continuous(labels = scales::percent_format(accuracy = 1)) +
  scale_colour_manual(values = diffusion_colours) +
  labs(title = "Productive nuclear delivery declines with cell volume",
       x = "Cell-volume multiplier", y = "Nuclear capture / membrane production",
       colour = "D (um2/s)") + plot_theme
```

At \(D=300\) um2/s, nuclear capture declines monotonically from 5.19% to 4.20%,
3.38%, 2.71%, and 2.17% as volume increases from 0.5x through 1x, 2x, 4x, and
8x. The 8x cell therefore captures about 48% less signal per unit membrane
production than the 1x cell. At \(D=3\) um2/s, the 1x-to-8x decline is from
3.87% to 1.61%, about 58%. The effect is not peculiar to the 1x/2x comparison;
slower diffusion progressively steepens the scaling at large sizes.

## Sensitivity to the abstract nuclear capture parameter

```{r capture-sensitivity, fig.width=8, fig.height=5}
capture_sensitivity <- data.frame()
for (diffusion in scenario$parameters$diffusion_um2_s) {
  for (permeability in scenario$parameters$nuclear_capture_sweep_um_s) {
    one <- run_delivery_model(1, diffusion,
                              capture_permeability_um_s = permeability)
    two <- run_delivery_model(2, diffusion,
                              capture_permeability_um_s = permeability)
    capture_sensitivity <- rbind(capture_sensitivity, data.frame(
      diffusion_um2_s = factor(diffusion),
      capture_permeability_um_s = permeability,
      fraction_1x = one$summary$nuclear_capture_fraction,
      fraction_2x = two$summary$nuclear_capture_fraction,
      log2_size_effect = log2(two$summary$nuclear_capture_fraction /
                               one$summary$nuclear_capture_fraction)
    ))
  }
}

ggplot(capture_sensitivity,
       aes(capture_permeability_um_s, log2_size_effect,
           colour = diffusion_um2_s, group = diffusion_um2_s)) +
  geom_hline(yintercept = 0, colour = "grey60", linetype = "dashed") +
  geom_line(linewidth = 0.8) + geom_point(size = 1.8) +
  scale_x_log10() + scale_colour_manual(values = diffusion_colours) +
  labs(title = "The size effect across nuclear-capture efficiencies",
       x = "Nuclear capture permeability (um/s)",
       y = "log2[nuclear delivery at 2V / delivery at V]",
       colour = "D (um2/s)") + plot_theme
```

Across the full \(P_{\rm nuc}=0.003\)--0.3 um/s sweep, the 2x cell always has a
lower delivery fraction. At \(D=300\) um2/s, the penalty ranges from about 21%
at weak capture to 14% at strong capture; at \(D=3\), it ranges from about 24%
to 19%. Increasing capture efficiency raises absolute productive delivery and
partially buffers the size effect, but does not reverse it in the explored
range.

## Parameter regimes: diffusion and maximum PDE hydrolysis rate

```{r regime-heatmap, fig.width=8, fig.height=6}
regime_heatmap <- data.frame()
diffusion_grid <- 10 ^ seq(log10(3), log10(300), length.out = 9)
pde_grid <- 2 ^ seq(-2, 2, length.out = 9)
for (diffusion in diffusion_grid) {
  for (pde_multiplier in pde_grid) {
    one <- run_delivery_model(1, diffusion, pde_multiplier = pde_multiplier)
    two <- run_delivery_model(2, diffusion, pde_multiplier = pde_multiplier)
    regime_heatmap <- rbind(regime_heatmap, data.frame(
      diffusion_um2_s = diffusion, pde_multiplier = pde_multiplier,
      pde_vmax_uM_s = pde_multiplier * scenario$parameters$pde_vmax_uM_s,
      fraction_1x = one$summary$nuclear_capture_fraction,
      fraction_2x = two$summary$nuclear_capture_fraction,
      log2_size_effect = log2(two$summary$nuclear_capture_fraction /
                               one$summary$nuclear_capture_fraction)
    ))
  }
}

ggplot(regime_heatmap,
       aes(diffusion_um2_s, pde_vmax_uM_s, fill = log2_size_effect)) +
  geom_tile() +
  geom_hline(yintercept = scenario$parameters$pde_vmax_uM_s,
             colour = "black", linetype = "dashed", linewidth = 0.45) +
  scale_x_log10() + scale_y_log10() +
  scale_fill_gradient2(low = "#3B4CC0", mid = "white", high = "#B40426",
                       midpoint = 0) +
  labs(title = "Size effect across diffusion and PDE hydrolysis capacity",
       x = "Diffusion coefficient (um2/s)",
       y = "PDE Vmax (uM/s per cytoplasmic volume)",
       fill = "log2(2V / V)",
       caption = "Dashed line: baseline PDE Vmax = 0.295 uM/s.") + plot_theme
```

The vertical axis is the \(V_{\max}\) in the PDE term of the first equation:
it is the maximum local hydrolysis rate per cytoplasmic volume, not the
concentration-dependent realized rate. The scan covers 0.0738--1.18 uM/s,
equivalent to 0.25--4 times the baseline \(V_{\max}=0.295\) uM/s. All 81
diffusion--PDE combinations have a
negative \(\log_2(2V/V)\). The 2x capture fraction is 63--85% of the 1x value,
with no sign reversal in the grid. This supports robustness to the tested
diffusion and PDE activity region, not universality beyond it.

## Mechanistic control: remove the distributed sink

```{r pde-control, fig.width=9, fig.height=6}
pde_zero_control <- data.frame()
pde_zero_control_time <- data.frame()
pde_zero_duration_s <- 3000
for (diffusion in scenario$parameters$diffusion_um2_s) {
  for (volume_multiplier in c(1, 2)) {
    result <- run_delivery_model(
      volume_multiplier, diffusion, pde_multiplier = 0,
      duration_s = pde_zero_duration_s, save_interval_s = 5
    )
    volume_label <- paste0(volume_multiplier, "x volume")
    pde_zero_control <- rbind(pde_zero_control,
      data.frame(diffusion_um2_s = diffusion,
                 volume = volume_label,
                 fate = "Nuclear capture",
                 fraction = result$summary$nuclear_capture_fraction),
      data.frame(diffusion_um2_s = diffusion,
                 volume = volume_label,
                 fate = "Net accumulation",
                 fraction = result$summary$accumulation_fraction)
    )
    pde_zero_control_time <- rbind(
      pde_zero_control_time,
      data.frame(
        diffusion_um2_s = diffusion, volume = volume_label,
        time_s = result$fate_trajectory$time_s,
        metric = "Instantaneous nuclear capture / production",
        fraction = result$fate_trajectory$instantaneous_nuclear_capture_fraction
      ),
      data.frame(
        diffusion_um2_s = diffusion, volume = volume_label,
        time_s = result$fate_trajectory$time_s,
        metric = "Cumulative nuclear capture / production",
        fraction = result$fate_trajectory$cumulative_nuclear_capture_fraction
      )
    )
  }
}

ggplot(subset(pde_zero_control_time, time_s > 0),
       aes(time_s, fraction, colour = volume, linetype = volume)) +
  geom_hline(yintercept = 1, colour = "grey60", linetype = "dotted") +
  geom_vline(xintercept = 300, colour = "grey60", linetype = "dashed") +
  geom_line(linewidth = 0.8) +
  facet_grid(metric ~ diffusion_um2_s,
             labeller = labeller(diffusion_um2_s = function(x) {
               paste0("D = ", x, " um2/s")
             })) +
  scale_x_sqrt(breaks = c(0, 300, 1000, 3000)) +
  scale_colour_manual(values = volume_colours) +
  scale_y_continuous(labels = scales::percent_format(accuracy = 1)) +
  labs(title = "Without PDE, the finite-time size effect vanishes at steady state",
       subtitle = "Dashed vertical line: the original 300-s endpoint; dotted horizontal line: input-output balance",
       x = "Time (s; square-root scale)", y = "Nuclear capture fraction",
       colour = "Cell size", linetype = "Cell size") +
  plot_theme
```

The upper row shows the instantaneous capture rate divided by the production
rate; the lower row shows capture integrated from time zero divided by
cumulative production. At 300 s and \(D=300\) um2/s, instantaneous capture is
67.3% for 1x and 58.8% for 2x, while cumulative capture is only 39.8% and 33.7%.
The previously reported deficit is therefore a transient inventory effect. By
3000 s, instantaneous capture exceeds 99.98% in both sizes, demonstrating that
at steady state nuclear capture is the only outlet and must equal membrane
production. Cumulative capture is then 91.1% and 88.7% and converges more slowly
because it retains the history of early accumulation. The same behavior occurs
at lower diffusion. PDE therefore converts a timing/inventory difference into
persistent allocation to an irreversible competing sink.

## Source-scaling control

```{r source-control, fig.width=9, fig.height=5}
source_scaling_control <- data.frame()
for (source_scaling in c("constant production per unit volume",
                         "constant flux per unit membrane area")) {
  for (diffusion in scenario$parameters$diffusion_um2_s) {
    for (volume_multiplier in c(1, 2)) {
      result <- run_delivery_model(volume_multiplier, diffusion,
                                   source_scaling = source_scaling)
      source_scaling_control <- rbind(source_scaling_control, data.frame(
        source_scaling = source_scaling, diffusion_um2_s = factor(diffusion),
        volume = paste0(volume_multiplier, "x volume"),
        nuclear_capture_fraction = result$summary$nuclear_capture_fraction
      ))
    }
  }
}
source_scaling_control$source_scaling <- factor(
  source_scaling_control$source_scaling,
  levels = c("constant production per unit volume",
             "constant flux per unit membrane area")
)

ggplot(source_scaling_control,
       aes(volume, nuclear_capture_fraction, colour = diffusion_um2_s,
           group = diffusion_um2_s)) +
  geom_line(linewidth = 0.8) + geom_point(size = 2) +
  facet_wrap(~ source_scaling) +
  scale_y_continuous(labels = scales::percent_format(accuracy = 1)) +
  scale_colour_manual(values = diffusion_colours) +
  labs(title = "Source scaling changes the magnitude of the size effect",
       x = "Cell size", y = "Nuclear capture / membrane production",
       colour = "D (um2/s)") + plot_theme
```

The primary model makes total membrane production proportional to cell volume
by scaling flux density as \(V^{1/3}\). The control instead holds flux per unit
membrane area constant, so total production scales with membrane area. Source
scaling changes PDE loading as well as the surface-to-volume relationship, so
the comparison tests how strongly the size effect depends on this assumption.
Under the volume-normalized default, the 2x capture penalty is 19.5% at
\(D=300\) and 22.9% at \(D=3\) um2/s. Under constant flux per membrane area,
the corresponding penalties increase to 30.1% and 34.3%. The qualitative size
effect persists, but its magnitude depends materially on source scaling.

## Interpretation and limits

Together, the simulations separate two mechanisms. Enlargement produces a
modest finite-time transport/capture disadvantage even without PDE, but the
larger effect arises when a target scaling with nuclear surface area competes
with an irreversible sink distributed throughout cytoplasmic volume. Doubling
volume reduces nuclear capture by about 19% even at \(D=300\) um2/s, where
spatial gradients are negligible; lowering diffusion raises that penalty to
about 23%. The effect remains negative throughout the tested capture, PDE, and
diffusion ranges and becomes stronger when flux per membrane area, rather than
production per cell volume, is held constant. The model therefore demonstrates
a plausible mechanism by which cell
enlargement changes signalling allocation rather than merely diffusion time.

This is a deliberately minimal radial model. It assumes concentric spherical
geometry, uniform membrane production, uniform PDE, and a phenomenological
nuclear capture boundary. It does not represent cAMP buffering, PDE/AC
nanodomains, nuclear pores, or downstream biochemical amplification. The
capture sweep is therefore more informative than any single fitted-looking
value of \(P_{\rm nuc}\).

The same geometry could later support an ERK-like first-order model. Classic
[reaction--diffusion theory](https://pubmed.ncbi.nlm.nih.gov/10970807/) found
substantial diffusion control for membrane kinase opposed by cytoplasmic
phosphatase, and live-cell measurements place GFP-ERK1 diffusion near
6--8 um2/s ([live-cell FCS study](https://pmc.ncbi.nlm.nih.gov/articles/PMC2823437/)).
