---
title: "Microtubule generation and transport in 1x and 2x cells"
output:
  html_document:
    toc: true
    toc_float: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE,
                      fig.align = "center", dpi = 110)
library(ggplot2)
library(gridExtra)
source("../R/mt_network.R")
source("../R/mt_transport.R")
cache_path <- "output/mt_transport_results.rds"
if (!file.exists(cache_path)) {
  stop("Generate report data first: Rscript scripts/generate_mt_transport_results.R")
}
results <- readRDS(cache_path)
network_sets <- results$network_sets
network_summary <- results$network_summary
transport_summary <- results$transport_summary
junction_sensitivity <- results$junction_sensitivity
membrane_bins <- results$membrane_bins
membrane_metrics <- results$membrane_metrics
membrane_size_divergence <- results$membrane_size_divergence
variant_order <- names(network_sets)
size_order <- c("1x volume", "2x volume")
for (object_name in c("network_summary", "transport_summary", "junction_sensitivity",
                      "membrane_bins", "membrane_metrics")) {
  object <- get(object_name)
  object$variant <- factor(object$variant, variant_order)
  object$size <- factor(object$size, size_order)
  assign(object_name, object)
}
membrane_size_divergence$variant <- factor(membrane_size_divergence$variant,
                                            variant_order)
cargo_results <- do.call(rbind, lapply(variant_order, function(variant) {
  do.call(rbind, lapply(size_order, function(size) {
    transform(results$transport[[variant]][[size]], variant = variant, size = size)
  }))
}))
row.names(cargo_results) <- NULL
cargo_results$variant <- factor(cargo_results$variant, variant_order)
cargo_results$size <- factor(cargo_results$size, size_order)
size_colours <- c(`1x volume` = "#126782", `2x volume` = "#E07A2D")
plot_theme <- theme_minimal(base_size = 10) +
  theme(legend.position = "bottom", panel.grid.minor = element_blank())
```

[&larr; Report catalogue](../index.html)

> **Status: literature-derived network components + custom transport model.**
> Shariff et al. motivate persistent constrained MT growth; Li et al. motivate
> rebound/regeneration rules. Density matching, pruning variants, junction
> detection, cargo hopping, pause distributions, and the 1x/2x experiment are
> modelling assumptions implemented here.

# Question and source-model boundary

At matched microtubule polymer length per cytoplasmic volume, does doubling
cell volume alter network junction structure, cargo delivery time, deadline
success, or the spatial distribution of successful membrane delivery?

This is a comparative computational experiment, not a reproduction of any one
paper. Published work informs particular network-growth and junction ideas;
the assembled static-network transport system and its quantitative outputs are
custom to this repository.

## Network generators

All networks are generated in a flattened three-dimensional ellipsoid containing
an excluded nuclear ellipsoid and one perinuclear centrosome. The displayed
`xy` panels are projections of 3D paths, so MTs can populate the projected
nuclear shadow by passing above or below the nucleus.

Four generation protocols are compared:

1. **Unpruned Shariff:** 0.2-um persistent steps constrained to the cytosol.
2. **Integrated-density pruning:** ten passes remove up to 50% of MTs, weighted
   by mean Gaussian-smoothed density along each entire path, then refill.
3. **Distal-tip pruning:** the same procedure, scoring only plus-end density.
4. **Li rebound/regeneration:** blocked paths can relax their turn constraint;
   anti-hairpin checks reject pathological paths, and complete trajectories are
   retried rather than retained at truncated length.

The basic growth model follows
[Shariff, Murphy and Rohde (2010)](https://pmc.ncbi.nlm.nih.gov/articles/PMC2901542/),
and the rebound/regeneration rules follow
[Li et al. (2012)](https://pmc.ncbi.nlm.nih.gov/articles/PMC3508979/).
Within every generator, the 2x network is independently generated to match the
realized 1x polymer length per cytoplasmic volume. Pruning refills to that same
target after every pass.

```{r network-panels, fig.width=10.5, fig.height=16}
network_plots <- list()
plot_id <- 0L
for (variant in variant_order) for (size in size_order) {
  plot_id <- plot_id + 1L
  network_plots[[plot_id]] <- plot_mt_network(
    network_sets[[variant]][[size]], "xy", paste(variant, size, sep = " — "),
    max_filaments = 100, seed = 200 + plot_id)
}
do.call(grid.arrange, c(network_plots, ncol = 2))
```

Only a reproducible random subset of at most 100 MTs is drawn per panel. All
MTs are retained for density, junction and transport calculations.

```{r network-summary}
network_display <- network_summary[, c("variant", "size", "filament_count",
  "total_length_um", "achieved_density", "mean_realized_length_um",
  "junction_count", "mean_junction_multiplicity",
  "high_multiplicity_fraction", "bundle_count")]
network_display$total_length_um <- round(network_display$total_length_um, 1)
network_display$achieved_density <- round(network_display$achieved_density, 4)
network_display$mean_realized_length_um <- round(network_display$mean_realized_length_um, 2)
network_display$mean_junction_multiplicity <- round(network_display$mean_junction_multiplicity, 2)
network_display$high_multiplicity_fraction <- round(network_display$high_multiplicity_fraction, 3)
names(network_display) <- c("Generator", "Cell size", "MTs", "Polymer (um)",
  "Polymer density", "Mean MT length (um)", "Junctions",
  "Mean multiplicity", "Fraction >5 MTs", "Bundles")
knitr::kable(network_display)
```

## Three-dimensional junctions

An intersection is a close encounter between MT centrelines, not a projected
line crossing. Segment midpoints provide a fast spatial shortlist, after which
the closest points and minimum 3D separation are calculated. The default
encounter radius is 0.25 um. Pairwise encounters within 0.5 um are merged into
one junction, whose strength is the number of distinct MTs. The shared
centrosomal origin is excluded. Near-parallel MT pairs that remain close for at
least 1 um are recorded as bundles rather than junctions.

Each junction stores its position, participating MT identities, multiplicity,
minimum separation, crossing angles and arc-length position on every member MT.
The latter lets transport encounter junctions in route order.

The 3D definition is motivated by observations that intersecting tracks can
have axial separations from approximately 200 nm to below 1 um, and that cargo
behavior depends on separation and angle
([Gao et al., 2018](https://pmc.ncbi.nlm.nih.gov/articles/PMC6026372/)).
Multiplicity classes are `2`, `3–5`, and `>5` MTs, following
[Chen et al. (2023)](https://www.nature.com/articles/s41467-023-38452-2).

```{r junction-sensitivity, fig.width=10, fig.height=6}
ggplot(junction_sensitivity,
       aes(encounter_radius_um, junction_count, colour = size, group = size)) +
  geom_line(linewidth = 0.7) + geom_point(size = 2) +
  facet_wrap(~ variant, scales = "free_y") +
  scale_colour_manual(values = size_colours) +
  labs(title = "Junction count depends on the 3D encounter radius",
       x = "Encounter radius (um)", y = "Merged junctions", colour = "Cell size") +
  plot_theme
```

```{r junction-complexity, fig.width=9, fig.height=5}
ggplot(network_summary, aes(variant, high_multiplicity_fraction, fill = size)) +
  geom_col(position = position_dodge(width = 0.72), width = 0.68) +
  scale_fill_manual(values = size_colours) +
  scale_y_continuous(labels = scales::percent_format(accuracy = 1)) +
  labs(title = "Fraction of junctions containing more than five MTs",
       x = NULL, y = "High-multiplicity junction fraction", fill = "Cell size") +
  plot_theme + theme(axis.text.x = element_text(angle = 18, hjust = 1))
```

## Minimal cargo transport protocol

For each of 250 cargos per network:

1. Sample a source in a 0.75-um sphere around the perinuclear centrosome.
2. Capture the nearest MT if it is within 0.4 um.
3. Move plus-endward at 1 um/s.
4. Deliver upon coming within 0.3 um of the cell cortex.
5. At an interior MT end, search within 0.5 um for another MT, add a 1-s
   search cost, hop, and continue. If none is available, fail.
6. Fail if delivery exceeds 120 s.

At every junction, cargo pauses with probability 0.5. Delay depends only on
multiplicity: two-MT junctions have a short lognormal delay (median 1.5 s),
3–5-MT junctions draw uniformly from 3–10 s, and >5-MT junctions use a
heavy-tailed lognormal delay (median 20 s). These illustrative distributions
reflect the reported shift from approximately two-second passage at simple
junctions, through several-to-ten-second delays at intermediate junctions, to
occasional delays exceeding a minute at complex junctions
([Chen et al., 2023](https://www.nature.com/articles/s41467-023-38452-2)).
They are not fitted kinetic parameters.

The model has no dynein, motor switching, endosomal sorting or dynamic MT
turnover.

## Transport results

```{r transport-plots, fig.width=11, fig.height=9}
delivery_plot <- ggplot(subset(cargo_results, success),
  aes(size, delivery_time_s, fill = size)) +
  geom_boxplot(outlier.alpha = 0.25, width = 0.65) +
  facet_wrap(~ variant, scales = "free_y") +
  scale_fill_manual(values = size_colours, guide = "none") +
  labs(title = "Delivery time among successful cargos", x = NULL,
       y = "Delivery time (s)") + plot_theme
success_plot <- ggplot(transport_summary,
  aes(variant, success_probability, fill = size)) +
  geom_col(position = position_dodge(width = 0.72), width = 0.68) +
  scale_fill_manual(values = size_colours) +
  scale_y_continuous(limits = c(0, 1), labels = scales::percent_format()) +
  labs(title = "Delivery success before 120 s", x = NULL,
       y = "Success probability", fill = "Cell size") + plot_theme +
  theme(axis.text.x = element_text(angle = 18, hjust = 1))
hops_plot <- ggplot(transport_summary, aes(variant, mean_hops, fill = size)) +
  geom_col(position = position_dodge(width = 0.72), width = 0.68) +
  scale_fill_manual(values = size_colours) +
  labs(title = "MT-to-MT transfers", x = NULL, y = "Mean hops per cargo",
       fill = "Cell size") + plot_theme +
  theme(axis.text.x = element_text(angle = 18, hjust = 1))
inefficiency_plot <- ggplot(subset(cargo_results, success),
  aes(size, route_inefficiency, fill = size)) +
  geom_boxplot(outlier.alpha = 0.25, width = 0.65) +
  facet_wrap(~ variant, scales = "free_y") +
  scale_fill_manual(values = size_colours, guide = "none") +
  labs(title = "Route inefficiency among successful cargos", x = NULL,
       y = "Path length / straight-line distance") + plot_theme
grid.arrange(delivery_plot, success_plot, hops_plot, inefficiency_plot, ncol = 2)
```

```{r transport-table}
transport_display <- transport_summary[, c("variant", "size",
  "success_probability", "median_delivery_time_s", "mean_hops",
  "median_route_inefficiency")]
transport_display$success_probability <- round(transport_display$success_probability, 3)
transport_display$median_delivery_time_s <- round(transport_display$median_delivery_time_s, 2)
transport_display$mean_hops <- round(transport_display$mean_hops, 3)
transport_display$median_route_inefficiency <- round(transport_display$median_route_inefficiency, 3)
names(transport_display) <- c("Generator", "Cell size", "P(success)",
  "Median delivery time (s)", "Mean hops", "Median path/straight")
knitr::kable(transport_display)
```

Across all four generators, the 2x cell has longer median delivery times and
lower deadline success than its matched-density 1x counterpart. The enlarged
networks also contain more junctions. For the unpruned Shariff network, the
fraction of >5-MT junctions rises from about 5.9% to 12.0%; pruning attenuates
but does not eliminate that increase. These are outputs of this illustrative
static model, not biological estimates.

Route inefficiency changes less consistently than delivery time because the
ratio is conditional on successful cargo. Long, inefficient routes that miss
the deadline are excluded, so it must be interpreted alongside success.

## Membrane distribution of delivered cargo

Successful delivery sites are mapped onto the ellipsoidal cell membrane and
binned into 12 azimuth sectors by four equal-solid-angle latitude bands. The
reference distribution is uniform per unit **membrane area**, with correction
for the ellipsoid surface Jacobian. Thus, zero log2 enrichment means the cargo
occupies that patch as often as expected under uniform membrane delivery;
positive and negative values mark enriched and depleted patches.

```{r membrane-enrichment, fig.width=10.5, fig.height=10}
ggplot(membrane_bins,
       aes(azimuth_mid * 180 / pi, equal_area_latitude_mid,
           fill = log2_enrichment)) +
  geom_tile() +
  facet_grid(variant ~ size) +
  scale_fill_gradient2(low = "#2166AC", mid = "white", high = "#B2182B",
                       midpoint = 0, limits = c(-2, 2),
                       oob = scales::squish) +
  scale_x_continuous(breaks = c(-180, -90, 0, 90, 180)) +
  labs(title = "Delivered-cargo enrichment across the membrane",
       subtitle = "Successful deliveries; colour is relative to uniform membrane area",
       x = "Azimuth (degrees)", y = "sin(latitude)",
       fill = "log2 enrichment") +
  plot_theme
```

Total variation is the directly interpretable nonuniformity score: zero is a
uniform membrane distribution and one is maximal concentration. The
polarization resultant measures whether delivery is biased toward one side of
the cell (zero for balanced directions, one if all delivery is at one point).
Jensen--Shannon divergence quantifies how much the full delivery pattern changes
between matched-density 1x and 2x networks.

```{r membrane-metrics, fig.width=11, fig.height=4.8}
nonuniformity_plot <- ggplot(membrane_metrics,
  aes(variant, total_variation_from_uniform, fill = size)) +
  geom_col(position = position_dodge(width = 0.72), width = 0.68) +
  scale_fill_manual(values = size_colours) +
  scale_y_continuous(limits = c(0, 1)) +
  labs(title = "Departure from uniform membrane delivery",
       x = NULL, y = "Total variation from uniform", fill = "Cell size") +
  plot_theme + theme(axis.text.x = element_text(angle = 18, hjust = 1))

polarization_plot <- ggplot(membrane_metrics,
  aes(variant, polarization_resultant, fill = size)) +
  geom_col(position = position_dodge(width = 0.72), width = 0.68) +
  scale_fill_manual(values = size_colours) +
  scale_y_continuous(limits = c(0, 1)) +
  labs(title = "One-sided polarization of delivery sites",
       x = NULL, y = "Directional resultant", fill = "Cell size") +
  plot_theme + theme(axis.text.x = element_text(angle = 18, hjust = 1))

size_divergence_plot <- ggplot(membrane_size_divergence,
  aes(variant, js_between_sizes_bits, fill = variant)) +
  geom_col(width = 0.68, show.legend = FALSE) +
  labs(title = "Change in membrane pattern from 1x to 2x",
       x = NULL, y = "Jensen-Shannon divergence (bits)") +
  plot_theme + theme(axis.text.x = element_text(angle = 18, hjust = 1))

grid.arrange(nonuniformity_plot, polarization_plot, size_divergence_plot,
             ncol = 3)
```

```{r membrane-table}
membrane_display <- membrane_metrics[, c("variant", "size", "n_delivered",
  "total_variation_from_uniform", "js_from_uniform_bits",
  "polarization_resultant")]
membrane_display$total_variation_from_uniform <-
  round(membrane_display$total_variation_from_uniform, 3)
membrane_display$js_from_uniform_bits <- round(membrane_display$js_from_uniform_bits, 3)
membrane_display$polarization_resultant <- round(membrane_display$polarization_resultant, 3)
names(membrane_display) <- c("Generator", "Cell size", "Delivered",
  "TV from uniform", "JS from uniform (bits)", "Polarization")
knitr::kable(membrane_display)
```

These estimates are conditional on successful delivery and use only 250 cargo
trials per network. The half-count pseudocount stabilizes empty bins; replicate
networks and more cargos are required before interpreting fine spatial patches.

## Limitations and next checks

The largest uncertainties are the encounter radius, junction merging, pause
kinetics, capture/hop radii and cortex criterion. Junction-radius sensitivity is
shown above, but transport should also be rerun across those values. Replicate
network realizations are needed before treating 1x–2x differences as robust
generator-level effects. This comparison isolates static network geometry; it
does not model motor detachment kinetics, bidirectionality, MT turnover or
cargo-specific biology.
