---
title: "Cell-size scaling of oxygen transport: oxygen-replete versus oxygen-limited regimes"
output:
  html_document:
    toc: true
    toc_float: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE)
library(ggplot2)
source("../R/oxygen_model.R")
source("../R/oxygen_model_3d.R")
source("../scenarios/sedlack_single_cell_oxygen.R")
scenario <- sedlack_single_cell_oxygen_scenario()
cell_colours <- c(`1x volume` = "#0072B2", `2x volume` = "#D55E00")
report_theme <- theme_minimal(base_size = 11) + theme(legend.position = "bottom")
```

[&larr; Report catalogue](../index.html)

> **Status: literature-derived simplification + new size experiment.** Sedlack
> et al. supply the oxygen-transport motivation and parameter targets; this
> report replaces their reconstructed mitochondrial geometry with a uniform
> cytoplasmic sink. The 1x/2x comparison is performed here.

The exact data used in every plot can be downloaded from the project landing
page, and the complete report and model source is available in the
[repository](https://github.com/Richard-Beck/scaling_sandbox).

## Scope

This report implements a minimal 3D Cartesian analogue of Sedlack et al.
(2022). It retains the transport features relevant to cell-size scaling while
removing the reconstructed mitochondrial geometry:

- a 150-um-high medium box containing one cell in a repeating monolayer;
- a flattened half-ellipsoidal cell on an impermeable glass substrate;
- domain-specific oxygen diffusion and a finite-permeability plasma membrane;
- uniform intracellular Michaelis-Menten oxygen consumption; and
- fixed oxygen at the top, with no-flux bottom and lateral boundaries.

The model cannot predict a distinct mitochondrial pO2. Its purpose is to
separate cell-scale geometry and surface-area-to-volume effects from
subcellular mitochondrial localization.

## Equations and membrane coupling

At steady state,

\[
0=\nabla\cdot(D\nabla c)-R(c).
\]

The extracellular medium has `R = 0`. Inside the cell,

\[
R(c)=V_{\max,\mathrm{eff}}\frac{c}{K_m+c}.
\]

At each voxel face crossing the plasma membrane, the two half-voxel diffusion
resistances and the membrane resistance are placed in series. Because Sedlack
assigns equal aqueous oxygen solubility to medium and cytoplasm, the face flux
can be written

\[
J=\frac{c_{\rm medium}-c_{\rm cell}}
{\Delta_m/(2D_m)+1/P+\Delta_c/(2D_c)}.
\]

The same flux enters one voxel and leaves the other, so mass is conserved. The
large baseline permeability makes membrane resistance nearly negligible, but
retaining it makes that assumption testable.

## Geometry, boundaries, and parameters

The domain is

\[
x,y\in[-L/2,L/2],\qquad z\in[0,150\ \mu\mathrm m],
\]

and the cell occupies

\[
\frac{x^2}{a_x^2}+\frac{y^2}{a_y^2}+\frac{z^2}{h^2}\le1,
\qquad z\ge0.
\]

The baseline axes are `(30, 30, 12) um`. The top face is fixed at applied pO2,
the glass at `z = 0` is no flux, and lateral no-flux faces represent symmetry
between identical cells spaced by `L`.

```{r parameter-table, results='asis'}
parameter_table <- data.frame(
  quantity = c(
    "Medium diffusion", "Cytoplasmic diffusion", "Aqueous solubility",
    "Membrane permeability", "Mitochondrial-domain Vmax (extreme)",
    "Homogenized Vmax (3D calibrated baseline)", "Km", "Cell axes",
    "Medium height", "Calibration spacing", "Voxel size (x,y,z)"
  ),
  value = c(
    scenario$transport$diffusion_medium_um2_s,
    scenario$transport$diffusion_cytoplasm_um2_s,
    scenario$transport$solubility_medium_mol_m3_mmhg,
    scenario$transport$membrane_permeability_um_s,
    scenario$consumption$vmax_mito_mol_m3_s,
    scenario$consumption$vmax_effective_mol_m3_s,
    scenario$consumption$km_mol_m3,
    paste(scenario$geometry$cell_radii_um, collapse = ", "),
    scenario$geometry$medium_height_um,
    scenario$geometry$calibration_spacing_um,
    paste(c(scenario$geometry$grid_dx_um, scenario$geometry$grid_dx_um,
            scenario$geometry$grid_dz_um), collapse = ", ")
  ),
  units = c(
    "um^2/s", "um^2/s", "mol/m^3/mmHg", "um/s", "mol/m^3/s",
    "mol/m^3/s", "mol/m^3", "um", "um", "um", "um"
  )
)
knitr::kable(parameter_table, row.names = FALSE)
```

Sedlack's `10 mol/m^3/s` applies only inside explicitly reconstructed
mitochondria and was chosen as an extreme upper bound. Multiplying it by an
invented mitochondrial fraction is not a defensible baseline. The
homogenized rate is instead calibrated once in this 3D geometry. The former
`1 mol/m^3/s` homogenized setting remains available only as an extreme
sensitivity case. Source values and comparison data are from the
[Sedlack et al. open-access article](https://pmc.ncbi.nlm.nih.gov/articles/PMC9604273/).

## One-time 3D calibration

The calibration condition is 5% applied O2, 100-um cell spacing, and a target
mean cytoplasmic pO2 of 21.4 mmHg. Only `Vmax_eff` is fitted. It is then frozen
for all validation and size-scaling runs.

```{r calibrate-3d}
calibration_grid <- build_oxygen_grid_3d(
  spacing_um = scenario$geometry$calibration_spacing_um,
  medium_height_um = scenario$geometry$medium_height_um,
  cell_axes_um = scenario$geometry$cell_radii_um,
  dx_um = scenario$geometry$grid_dx_um,
  dz_um = scenario$geometry$grid_dz_um
)
calibration_transport <- assemble_oxygen_transport_3d(
  calibration_grid, scenario,
  scenario$boundary_conditions$screening_po2_mmhg
)
calibrated_vmax <- calibrate_oxygen_vmax_3d(
  calibration_grid, calibration_transport, scenario,
  target_mean_po2_mmhg = scenario$consumption$calibration_target_cytoplasm_po2_mmhg,
  bracket_mol_m3_s = c(0, 0.15)
)
scenario$consumption$vmax_effective_mol_m3_s <- calibrated_vmax
calibration_fit <- solve_oxygen_steady_3d(
  calibration_grid, calibration_transport, scenario,
  vmax_effective_mol_m3_s = calibrated_vmax
)
calibration_summary <- data.frame(
  calibrated_vmax_mol_m3_s = calibrated_vmax,
  target_mean_cytoplasmic_po2_mmhg = scenario$consumption$calibration_target_cytoplasm_po2_mmhg,
  fitted_mean_cytoplasmic_po2_mmhg = calibration_fit$mean_cell_po2_mmhg,
  fitted_total_ocr_fmol_s = calibration_fit$total_ocr_fmol_s,
  represented_cell_volume_um3 = calibration_grid$represented_cell_volume_um3,
  solver_converged = calibration_fit$converged
)
knitr::kable(calibration_summary, digits = 4)
```

The calibrated \(V_{\max,\mathrm{eff}}\) is 0.09869 mol/m3/s. The fitted mean
is 21.4001 mmHg against a 21.4-mmHg target, with a corresponding whole-cell OCR
of 2.242 fmol/s. This exact agreement is one fitted datum, not validation. The
value is specific to this geometry and discretization and should eventually be
checked against measured whole-cell OCR and a grid-refinement study.

## Validation against conditions not used for calibration

```{r validate-3d, fig.width=7, fig.height=5}
validation <- data.frame(
  spacing_um = c(70, 100, 200, 70),
  applied_po2_mmhg = c(38, 38, 38, 3.8),
  observed_mean_cytoplasmic_po2_mmhg = c(7.09, 21.4, 28.3, 0.23),
  condition = c("5% O2, 70 um", "5% O2, 100 um (calibration)",
                "5% O2, 200 um", "0.5% O2, 70 um")
)
validation$predicted_mean_cytoplasmic_po2_mmhg <- NA_real_
validation$predicted_ocr_fmol_s <- NA_real_
validation$point_type <- ifelse(grepl("calibration", validation$condition),
                                "Calibration", "Held out")

for (row in seq_len(nrow(validation))) {
  grid <- build_oxygen_grid_3d(
    validation$spacing_um[row], scenario$geometry$medium_height_um,
    scenario$geometry$cell_radii_um,
    dx_um = scenario$geometry$grid_dx_um,
    dz_um = scenario$geometry$grid_dz_um
  )
  transport <- assemble_oxygen_transport_3d(
    grid, scenario, validation$applied_po2_mmhg[row]
  )
  fit <- if (validation$applied_po2_mmhg[row] < 10) {
    solve_oxygen_continuation_3d(grid, transport, scenario, calibrated_vmax)
  } else {
    solve_oxygen_steady_3d(grid, transport, scenario, calibrated_vmax)
  }
  validation$predicted_mean_cytoplasmic_po2_mmhg[row] <- fit$mean_cell_po2_mmhg
  validation$predicted_ocr_fmol_s[row] <- fit$total_ocr_fmol_s
}

validation$residual_mmhg <- validation$predicted_mean_cytoplasmic_po2_mmhg -
  validation$observed_mean_cytoplasmic_po2_mmhg
validation$relative_error_percent <- 100 * validation$residual_mmhg /
  validation$observed_mean_cytoplasmic_po2_mmhg

ggplot(validation,
       aes(observed_mean_cytoplasmic_po2_mmhg,
           predicted_mean_cytoplasmic_po2_mmhg, label = condition)) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", colour = "grey50") +
  geom_point(aes(shape = point_type, colour = point_type), size = 3) +
  geom_text(nudge_y = 1.2, size = 3, check_overlap = TRUE) +
  coord_equal() +
  labs(title = "Validation of the frozen 3D calibration",
       x = "Sedlack mean cytoplasmic pO2 (mmHg)",
       y = "Homogenized-model mean cytoplasmic pO2 (mmHg)",
       shape = "Point type", colour = "Point type") +
  scale_colour_manual(values = c(Calibration = "#D55E00", `Held out` = "#0072B2")) +
  report_theme

knitr::kable(validation, digits = 3)
```

At 5% O2, the 70-um held-out condition is predicted at 8.97 rather than 7.09
mmHg (residual +1.88 mmHg, +26.6%), while the 200-um condition is predicted at
29.01 rather than 28.3 mmHg (+0.71 mmHg, +2.5%). At 0.5% O2 and 70-um spacing,
the model predicts only 0.0145 mmHg against 0.23 mmHg (about -94%). Thus the
model closely reproduces the 200-um held-out condition and gets the qualitative
5%-O2 spacing trend, but performs poorly under severe hypoxia. The low-oxygen
size simulations below are therefore a mechanistic demonstration of this
model's oxygen-limited regime, not a quantitatively validated prediction of
cellular pO2 or OCR under severe hypoxia. Possible missing effects include
mitochondrial localization, detailed geometry, the square-array medium field,
and resolution near anoxic regions.

## 1x versus 2x cell volume

For the size comparison, all three half-ellipsoid axes scale by `V^(1/3)`.
Consumption capacity per cell volume remains fixed at the one calibrated value,
so maximum total OCR scales with cell volume. Cell spacing stays at 100 um.
The baseline 38-mmHg medium boundary is compared with a hypoxic boundary of
3.8 mmHg—10% of the baseline oxygen concentration.

```{r size-scaling-3d, fig.width=10, fig.height=5}
volume_conditions <- c(`1x volume` = 1, `2x volume` = 2)
oxygen_conditions <- c(`Baseline: 38 mmHg` = 38, `Hypoxia: 3.8 mmHg` = 3.8)
slice_data <- data.frame()
size_summary <- data.frame()
po2_distribution_summary <- data.frame()

for (oxygen_name in names(oxygen_conditions)) {
  for (volume_name in names(volume_conditions)) {
    scale <- volume_conditions[[volume_name]] ^ (1 / 3)
    axes <- scenario$geometry$cell_radii_um * scale
    grid <- build_oxygen_grid_3d(
      spacing_um = 100, medium_height_um = scenario$geometry$medium_height_um,
      cell_axes_um = axes, dx_um = scenario$geometry$grid_dx_um,
      dz_um = scenario$geometry$grid_dz_um
    )
    transport <- assemble_oxygen_transport_3d(
      grid, scenario, oxygen_conditions[[oxygen_name]]
    )
    fit <- if (oxygen_conditions[[oxygen_name]] < 10) {
      solve_oxygen_continuation_3d(grid, transport, scenario, calibrated_vmax)
    } else {
      solve_oxygen_steady_3d(grid, transport, scenario, calibrated_vmax)
    }
    y_slice <- grid$y_um[which.min(abs(grid$y_um))]
    keep <- grid$coordinates$y_um == y_slice
    slice_data <- rbind(
      slice_data,
      data.frame(
        oxygen_condition = oxygen_name,
        volume = volume_name,
        x_um = grid$coordinates$x_um[keep],
        z_um = grid$coordinates$z_um[keep],
        region = ifelse(grid$is_cell[keep], "Cell", "Medium"),
        po2_mmhg = fit$po2_mmhg[keep],
        consumption_mol_m3_s = ifelse(
          grid$is_cell[keep], fit$consumption_mol_m3_s[keep], NA_real_
        )
      )
    )
    size_summary <- rbind(
      size_summary,
      data.frame(
        oxygen_condition = oxygen_name,
        volume = volume_name,
        represented_cell_volume_um3 = grid$represented_cell_volume_um3,
        mean_cytoplasmic_po2_mmhg = fit$mean_cell_po2_mmhg,
        consumption_per_volume_mol_m3_s = fit$mean_cell_consumption_mol_m3_s,
        total_ocr_fmol_s = fit$total_ocr_fmol_s,
        converged = fit$converged
      )
    )
    cell_po2 <- fit$po2_mmhg[grid$is_cell]
    km_po2_mmhg <- scenario$consumption$km_mol_m3 /
      scenario$transport$solubility_medium_mol_m3_mmhg
    analytic_volume_um3 <- 2 * pi / 3 * prod(axes)
    po2_distribution_summary <- rbind(
      po2_distribution_summary,
      data.frame(
        oxygen_condition = oxygen_name, volume = volume_name,
        min_po2_mmhg = min(cell_po2),
        q05_po2_mmhg = unname(quantile(cell_po2, 0.05)),
        median_po2_mmhg = median(cell_po2),
        q95_po2_mmhg = unname(quantile(cell_po2, 0.95)),
        max_po2_mmhg = max(cell_po2),
        fraction_voxels_below_km = mean(cell_po2 < km_po2_mmhg),
        km_po2_mmhg = km_po2_mmhg,
        analytic_cell_volume_um3 = analytic_volume_um3,
        represented_cell_volume_um3 = grid$represented_cell_volume_um3,
        represented_volume_error_percent = 100 *
          (grid$represented_cell_volume_um3 / analytic_volume_um3 - 1)
      )
    )
  }
}

kinetics_curve <- data.frame(
  series = "Michaelis-Menten curve",
  po2_mmhg = exp(seq(log(0.001), log(50), length.out = 300))
)
kinetics_curve$relative_rate <- kinetics_curve$po2_mmhg /
  (km_po2_mmhg + kinetics_curve$po2_mmhg)
kinetics_curve$oxygen_condition <- NA_character_
kinetics_curve$volume <- NA_character_
kinetics_points <- data.frame(
  series = "Size-scaling means",
  po2_mmhg = size_summary$mean_cytoplasmic_po2_mmhg,
  relative_rate = size_summary$mean_cytoplasmic_po2_mmhg /
    (km_po2_mmhg + size_summary$mean_cytoplasmic_po2_mmhg),
  oxygen_condition = size_summary$oxygen_condition,
  volume = size_summary$volume
)
kinetics <- rbind(kinetics_curve, kinetics_points)

ggplot(slice_data, aes(x_um, z_um, fill = po2_mmhg)) +
  geom_raster() +
  geom_contour(data = slice_data,
               aes(x = x_um, y = z_um, z = as.numeric(region == "Cell")),
               inherit.aes = FALSE, breaks = 0.5,
               colour = "white", linewidth = 0.4) +
  facet_grid(oxygen_condition ~ volume) +
  coord_equal() +
  scale_fill_viridis_c(option = "C") +
  labs(title = "Central 3D-model slice: intra- and extracellular oxygen",
       x = "x (um)", y = "z above glass (um)", fill = "pO2\n(mmHg)",
       caption = "White contour: voxelized plasma membrane. Fill scales are shared across panels.") +
  report_theme
```

## Oxygen kinetics explain when the gradient matters

```{r oxygen-kinetics, fig.width=8, fig.height=5}
ggplot() +
  geom_line(data = kinetics_curve, aes(po2_mmhg, relative_rate), linewidth = 0.9) +
  geom_vline(xintercept = km_po2_mmhg, linetype = "dashed", colour = "grey50") +
  geom_point(data = kinetics_points,
             aes(po2_mmhg, relative_rate, colour = oxygen_condition, shape = volume),
             size = 3) +
  scale_x_log10() +
  scale_y_continuous(labels = scales::percent_format(accuracy = 1)) +
  labs(title = "Oxygen-consumption kinetics and simulated mean pO2",
       x = "Mean cytoplasmic pO2 (mmHg, log scale)",
       y = expression(R/V[max]), colour = "Medium oxygen", shape = "Cell size",
       caption = "Dashed line: Michaelis-Menten half-saturation pO2.") +
  report_theme
```

The half-saturation concentration corresponds to about 0.0775 mmHg. Baseline
mean pO2 is 21.40 mmHg (about 276x this scale) in the 1x cell and 7.35 mmHg
(95x) in the 2x cell. The hypoxic means are instead 0.0342 mmHg (0.44x) and
0.0147 mmHg (0.19x). This kinetic position, rather than gradient magnitude
alone, explains why the baseline and hypoxic functional responses differ.

```{r size-scaling-consumption-3d, fig.width=10, fig.height=4}
ggplot(slice_data, aes(x_um, z_um, fill = consumption_mol_m3_s)) +
  geom_raster() +
  geom_contour(data = slice_data,
               aes(x = x_um, y = z_um, z = as.numeric(region == "Cell")),
               inherit.aes = FALSE, breaks = 0.5,
               colour = "white", linewidth = 0.4) +
  facet_grid(oxygen_condition ~ volume) +
  coord_equal() +
  scale_fill_viridis_c(option = "B", na.value = "grey92") +
  labs(title = "Uniform-field intracellular oxygen consumption",
       x = "x (um)", y = "z above glass (um)",
       fill = "Consumption\n(mol/m3/s)",
       caption = "Grey: extracellular medium, where consumption is zero.") +
  report_theme

ggplot(size_summary,
       aes(volume, consumption_per_volume_mol_m3_s,
           colour = oxygen_condition, group = oxygen_condition)) +
  geom_line(linewidth = 0.8) +
  geom_point(size = 2.5) +
  labs(title = "Integrated oxygen consumption per cell volume",
       x = "Cell size", y = "Consumption (mol/m3/s)",
       colour = "Medium oxygen") +
  report_theme

ggplot(size_summary,
       aes(volume, total_ocr_fmol_s,
           colour = oxygen_condition, group = oxygen_condition)) +
  geom_line(linewidth = 0.8) +
  geom_point(size = 2.5) +
  labs(title = "Total whole-cell oxygen consumption",
       x = "Cell size", y = "Total OCR (fmol/s)",
       colour = "Medium oxygen") +
  report_theme

knitr::kable(size_summary, digits = 4)
```

Under the baseline boundary, represented cell volume increases from 22,800 to
46,800 um3 (+105%; voxelization makes this 2.053x rather than exactly 2x).
Mean cytoplasmic pO2 falls from 21.40 to 7.35 mmHg (-66%), yet consumption per
volume falls only from 0.09833 to 0.09753 mol/m3/s (-0.82%) and OCR rises from
2.242 to 4.564 fmol/s (+104%). Both means remain orders of magnitude above the
kinetic half-saturation scale. A large concentration gradient is therefore not
automatically a functional limitation; its location on the kinetic response
curve matters. Full-volume distribution statistics, including the fraction of
intracellular voxels below \(K_m\), are exported with the plotting data.

Under the 3.8-mmHg boundary, mean pO2 falls from 0.0342 to 0.0147 mmHg (-57%)
and consumption per volume from 0.02384 to 0.01288 mol/m3/s (-46%). OCR rises
only from 0.5436 to 0.6027 fmol/s (+10.9%) despite roughly doubling consuming
volume. Once oxygen falls below half-saturation, the geometric deficit becomes
functionally consequential: added consuming volume is nearly offset by lower
respiration per unit volume. Because the corresponding low-O2 validation case
is severely over-depleted, this is a mechanistic regime result, not a
quantitatively validated severe-hypoxia prediction.

## Numerical limitations and next checks

This is a deliberately minimal voxel model. The current 10 x 10 x 3 um grid
staircases the cell surface. The analytic 1x half-ellipsoid is about 22,619 um3
versus 22,800 um3 represented (+0.8%); the nominal 2x analytic volume is about
45,239 um3 versus 46,800 um3 represented (+3.5%). Thus the simulated comparison
is 2.053x represented voxel volume. The calibration belongs to this
discretization, and small effects such as the -0.82% baseline change in
consumption per volume should not be overinterpreted before grid refinement.

The next biological checks are independent whole-cell OCR calibration,
alternative OCR-versus-volume scaling, substrate oxygen permeability, and
comparison of the homogenized sink with a localized-sink sensitivity model.
