---
title: "Cell-size scaling under imposed flux: spatial gradients without lost throughput"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE)
library(ggplot2)
source("../R/scaling.R")
source("../R/diffusion.R")
source("../R/plotting.R")
source("../scenarios/external_supply.R")
source("../scenarios/nuclear_processing.R")

volume_multipliers <- c(`1x volume` = 1, `2x volume` = 2)
volume_colours <- c(`1x volume` = "#0072B2", `2x volume` = "#D55E00")
volume_linetypes <- c(`1x volume` = "solid", `2x volume` = "dashed")
plot_theme <- theme_minimal(base_size = 11) + theme(legend.position = "bottom")
```

The exact data used in every plot can be downloaded from the project landing
page, and the complete report and model source is available in the
[repository](https://github.com/Richard-Beck/scaling_sandbox).

## External boundary: fixed input per cell volume

Total external influx is proportional to cell volume. Consequently, a larger
cell needs a larger flux density at its membrane (`J = input × volume / area`).
Dotted vertical lines mark the plasma membrane, making the distinct 1× and 2×
radii directly auditable.

```{r external-profiles, fig.width=10, fig.height=8}
external <- external_supply_scenario()
external_profiles <- data.frame()
external_boundaries <- data.frame()
external_consumption_summary <- data.frame()

for (consumer_name in names(external$parameters$consumers)) {
  consumer <- external$parameters$consumers[[consumer_name]]
  for (rate_name in names(external$parameters$input_rates_uM_s)) {
    rate <- external$parameters$input_rates_uM_s[[rate_name]]
    for (volume_name in names(volume_multipliers)) {
      geometry <- spherical_geometry(external$geometry$cell_radius_um,
                                     volume_multipliers[[volume_name]])
      grid <- radial_grid(geometry$radius_um, n_cells = 120)
      surface_flux <- rate * geometry$volume_um3 / geometry$area_um2
      reaction <- if (identical(consumer_name, "michaelis_menten")) {
        function(time_s, concentration) {
          consumer$vmax_uM_s * concentration / (consumer$km_uM + concentration)
        }
      } else NULL
      concentration <- simulate_radial_diffusion(
        grid, external$parameters$diffusion_um2_s, initial = 0, duration_s = 500,
        loss_s = if (identical(consumer_name, "first_order")) consumer$loss_s else 0,
        reaction = reaction,
        outer_boundary = list(type = "fixed_inward_flux", value = surface_flux)
      )
      consumption <- if (identical(consumer_name, "first_order")) {
        consumer$loss_s * concentration
      } else {
        consumer$vmax_uM_s * concentration / (consumer$km_uM + concentration)
      }
      external_profiles <- rbind(
        external_profiles,
        data.frame(consumer = consumer$label, input = tools::toTitleCase(rate_name),
                   volume = volume_name, radius_um = grid$centers_um,
                   quantity = "Concentration (uM)", value = concentration),
        data.frame(consumer = consumer$label, input = tools::toTitleCase(rate_name),
                   volume = volume_name, radius_um = grid$centers_um,
                   quantity = "Consumption (uM/s)", value = consumption)
      )
      external_boundaries <- rbind(
        external_boundaries,
        data.frame(consumer = consumer$label, input = tools::toTitleCase(rate_name),
                   volume = volume_name, cell_radius_um = geometry$radius_um)
      )
      external_consumption_summary <- rbind(
        external_consumption_summary,
        data.frame(consumer = consumer$label, input = tools::toTitleCase(rate_name),
                   volume = volume_name,
                   consumption_per_volume_uM_s = volume_weighted_mean(consumption, grid))
      )
    }
  }
}

plot_external_consumer <- function(consumer_label) {
  ggplot(subset(external_profiles, consumer == consumer_label),
         aes(radius_um, value, colour = volume, linetype = volume)) +
    geom_line(linewidth = 0.8) +
    geom_vline(data = subset(external_boundaries, consumer == consumer_label),
               aes(xintercept = cell_radius_um, colour = volume),
               linetype = "dotted", linewidth = 0.5, show.legend = FALSE) +
    facet_grid(quantity ~ input, scales = "free_y") +
    scale_colour_manual(values = volume_colours) +
    scale_linetype_manual(values = volume_linetypes) +
    labs(title = consumer_label, x = "Radius (um)", y = NULL,
         colour = "Cell size", linetype = "Cell size",
         caption = "Dotted lines: plasma-membrane position.") +
    plot_theme
}

plot_external_consumer("First-order consumption")
```

```{r external-saturable-profiles, fig.width=10, fig.height=8}
plot_external_consumer("Saturable consumption")
```

With first-order consumption, doubling cell volume leaves the volume-averaged
concentration unchanged: 1.00 uM at high input and 0.20 uM at low input. The
radial range nevertheless grows by about 58%. At high input it widens from
approximately 0.975--1.016 uM in the 1x cell to 0.961--1.026 uM in the 2x cell;
at both sizes, integrated consumption remains 0.050 uM/s per cell volume. At
low input, it likewise remains 0.010 uM/s per volume. Doubling volume therefore
does not reduce throughput in this flux-clamped model. Instead, the larger cell
pays a spatial cost: a larger membrane-to-centre concentration difference is
required to sustain the same consumption per unit volume. The approximately
58% increase is the expected \(2^{2/3}\simeq1.59\) scaling of the
diffusion-associated concentration difference.

The saturable sink responds differently in space. At high input, the range
widens from 0.0299--0.0662 uM to 0.0219--0.0755 uM (about 47%), while integrated
consumption remains exactly 0.050 uM/s per volume. At low input, the respective
ranges are 0.00243--0.00830 uM and 0.00158--0.00958 uM, with consumption fixed
at 0.010 uM/s per volume. The enlarged-cell centre operates at lower local
concentration and reaction rate while the membrane-adjacent region operates at
higher values; this spatial redistribution lets the integrated sink match the
imposed source. This does not show that saturable consumption is immune to cell
size: it shows that, while the imposed source remains below total sink capacity,
the concentration field can reorganize to accommodate that flux.

### Interpretation and limits of the fixed-flux assumption

At steady state, integrated consumption per cell volume equals the imposed
input per cell volume: there is no accumulation or alternative fate, and the
selected sinks retain sufficient capacity. If the volume-normalized input is
\(q\), the required membrane flux density is \(J=qR/3\); characteristic
diffusive concentration differences scale as \(JR/D=qR^2/(3D)\).

The exterior-shell plots use the same far-field concentration `C_bulk` for
both cell sizes. They do **not** make the far-field concentration size
dependent. Rather, for a no-consumption spherical exterior,
`C_bulk - C_surface = JR/D_external = qR^2/(3D_external)`: maintaining the
specified flux requires a larger depletion near the membrane as cells grow or
external diffusion slows. Conversely, if surface concentration were fixed,
the far-field concentration needed to sustain this flux would increase with
\(R^2/D_{\rm ext}\). These parameter choices remain feasible, but the analytic
relation implies that sufficiently large \(R\), low \(D_{\rm ext}\), or high
imposed flux would eventually require an unphysical negative surface
concentration.

The plasma membrane is presently a flux-clamp boundary: the prescribed flux
is imposed on the cell side and removed from the external shell, conserving
mass but allowing a concentration jump across the membrane. It has no
explicit thickness, permeability, partition coefficient, or transporter
kinetics. This is a useful limiting case for strongly regulated or active
uptake, but it is not a passive-membrane model. With fixed transporter density,
capacity would usually scale with membrane area, not cell volume. The separate
Sedlack-inspired oxygen report introduces a permeability-coupled membrane
formulation in which flux is an outcome rather than an input.

```{r external-integrated-consumption, fig.width=8, fig.height=4}
ggplot(external_consumption_summary,
       aes(input, consumption_per_volume_uM_s, colour = volume, linetype = volume,
           group = volume)) +
  geom_line(linewidth = 0.8) +
  geom_point(size = 2) +
  facet_wrap(~ consumer) +
  scale_colour_manual(values = volume_colours) +
  scale_linetype_manual(values = volume_linetypes) +
  labs(title = "Integrated cellular consumption: external-boundary scenarios",
       x = "Input rate per cell volume", y = "Consumption per cell volume (uM/s)",
       colour = "Cell size", linetype = "Cell size") +
  plot_theme
```

The overlap of the 1x and 2x curves is a mass-balance/control result, not an
independent biological finding. At numerical steady state, with no other fate
for substrate, integrated consumption must equal imposed influx. The
informative size effect is therefore the spatial concentration field rather
than total throughput.

## Cell plus exterior medium

The next figure shows the cell interior and an extracellular shell. Flux is
fixed at the membrane and the external medium has no consumption. The outer
shell uses the spherical far-field condition \(C \to C_{\rm bulk}\) and
\(dC/dr \to 0\) as \(r \to \infty\), implemented at a finite radius with the corresponding
Robin condition.

```{r interior-exterior, fig.width=10, fig.height=5}
rate <- external$parameters$input_rates_uM_s[["high"]]
# Keep intracellular diffusion fixed. Only the extracellular medium changes.
media <- list(
  `External diffusion slower` = c(inside = 20, outside = 5),
  `External diffusion faster` = c(inside = 20, outside = 50)
)
combined_profiles <- data.frame()
medium_boundaries <- data.frame()
medium_consumption_summary <- data.frame()

for (medium_name in names(media)) {
  for (volume_name in names(volume_multipliers)) {
    geometry <- spherical_geometry(external$geometry$cell_radius_um,
                                   volume_multipliers[[volume_name]])
    flux <- rate * geometry$volume_um3 / geometry$area_um2
    inside_grid <- radial_grid(geometry$radius_um, 120)
    inside <- simulate_radial_diffusion(
      inside_grid, media[[medium_name]][["inside"]], 0, 500,
      loss_s = external$parameters$consumers$first_order$loss_s,
      outer_boundary = list(type = "fixed_inward_flux", value = flux)
    )
    outside_grid <- radial_grid(external$parameters$external_medium_radius_um,
                                180, inner_radius_um = geometry$radius_um)
    outside <- simulate_radial_diffusion(
      outside_grid, media[[medium_name]][["outside"]],
      initial = external$parameters$external_bulk_concentration_uM, duration_s = 10000,
      inner_boundary = list(type = "fixed_outward_flux", value = -flux),
      outer_boundary = list(type = "infinite_reservoir",
                            value = external$parameters$external_bulk_concentration_uM)
    )
    combined_profiles <- rbind(
      combined_profiles,
      data.frame(medium = medium_name, volume = volume_name, region = "Cell interior",
                 radius_um = inside_grid$centers_um, concentration_uM = inside),
      data.frame(medium = medium_name, volume = volume_name, region = "External medium",
                 radius_um = outside_grid$centers_um, concentration_uM = outside)
    )
    medium_boundaries <- rbind(
      medium_boundaries,
      data.frame(medium = medium_name, volume = volume_name,
                 cell_radius_um = geometry$radius_um)
    )
    medium_consumption_summary <- rbind(
      medium_consumption_summary,
      data.frame(medium = medium_name, volume = volume_name,
                 consumption_per_volume_uM_s = volume_weighted_mean(
                   external$parameters$consumers$first_order$loss_s * inside, inside_grid
                 ))
    )
  }
}

ggplot(combined_profiles,
       aes(radius_um, concentration_uM, colour = volume, linetype = volume,
           group = interaction(volume, region))) +
  geom_line(linewidth = 0.8) +
  geom_vline(data = medium_boundaries,
             aes(xintercept = cell_radius_um, colour = volume),
             linetype = "dotted", linewidth = 0.5, show.legend = FALSE) +
  facet_wrap(~ medium) +
  scale_colour_manual(values = volume_colours) +
  scale_linetype_manual(values = volume_linetypes) +
  labs(title = "Internal and external concentration gradients",
       x = "Radius from cell centre (um)", y = "Concentration (uM)",
       colour = "Cell size", linetype = "Cell size",
       caption = "Dotted lines: plasma membrane. Each line breaks at the membrane.") +
  plot_theme
```

At high input, the concentration immediately outside the membrane is about
0.967 uM (1x) and 0.948 uM (2x) when \(D_{\rm ext}=50\) um2/s, but only 0.675
and 0.481 uM when \(D_{\rm ext}=5\) um2/s. Despite this large extracellular
difference, intracellular profiles are effectively identical between media,
and consumption remains 0.050 uM/s per volume for both sizes. Slower external
transport therefore amplifies the depletion required to support the imposed
flux, especially in the larger cell. Because influx is prescribed rather than
concentration-dependent, that depletion cannot feed back onto uptake. This
exercise identifies the extracellular concentration cost of maintaining a
volume-scaled flux; it does not model extracellular diffusion limitation of
uptake.

```{r interior-exterior-integrated-consumption, fig.width=7, fig.height=4}
ggplot(medium_consumption_summary,
       aes(volume, consumption_per_volume_uM_s, colour = medium, group = medium)) +
  geom_line(linewidth = 0.8) +
  geom_point(size = 2) +
  labs(title = "Integrated cellular consumption: exterior-medium scenarios",
       x = "Cell size", y = "Consumption per cell volume (uM/s)",
       colour = "External diffusion condition") +
  plot_theme
```

## Internal boundary: nuclear export with two cytoplasmic sinks

Nuclear export supplies the diffusing species at the nuclear envelope. It is
removed through two distinct cytoplasmic processes: saturable consumption
`Vmax × C / (Km + C)` and first-order degradation `k × C`. The plots below
show the final state after the prescribed degradation-rate change. Dotted
vertical lines mark the nuclear envelope and dashed lines mark the plasma
membrane. Their locations scale with `V^(1/3)`.

```{r nuclear-processing, fig.width=8, fig.height=5}
nuclear <- nuclear_processing_scenario()
cytoplasmic_profiles <- data.frame()
cytoplasmic_boundaries <- data.frame()
cytoplasmic_consumption_summary <- data.frame()
cytoplasmic_sink_summary <- data.frame()
cytoplasmic_sink_profiles <- data.frame()
for (volume_name in names(volume_multipliers)) {
  whole_cell <- spherical_geometry(nuclear$geometry$cell_radius_um,
                                  volume_multipliers[[volume_name]])
  nuclear_radius <- whole_cell$radius_um * nuclear$geometry$nucleus_radius_fraction
  grid <- radial_grid(whole_cell$radius_um, 120, inner_radius_um = nuclear_radius)
  cytoplasm_volume <- sum(grid$volumes_um3)
  nuclear_area <- 4 * pi * nuclear_radius ^ 2
  export_flux <- nuclear$parameters$source_rate * cytoplasm_volume / nuclear_area
  degradation <- function(time_s) {
    ifelse(time_s < nuclear$parameters$degradation_transition_s,
           nuclear$parameters$initial_degradation_s,
           nuclear$parameters$final_degradation_s)
  }
  concentration <- simulate_radial_diffusion(
    grid, nuclear$parameters$diffusion_um2_s, 0, 200,
    loss_s = degradation,
    reaction = function(time_s, concentration) {
      nuclear$parameters$consumption_vmax_uM_s * concentration /
        (nuclear$parameters$consumption_km_uM + concentration)
    },
    inner_boundary = list(type = "fixed_outward_flux", value = export_flux)
  )
  consumption_rate <- nuclear$parameters$consumption_vmax_uM_s * concentration /
    (nuclear$parameters$consumption_km_uM + concentration)
  degradation_rate <- nuclear$parameters$final_degradation_s * concentration
  cytoplasmic_profiles <- rbind(
    cytoplasmic_profiles,
    data.frame(volume = volume_name, radius_um = grid$centers_um,
               concentration_uM = concentration)
  )
  cytoplasmic_boundaries <- rbind(
    cytoplasmic_boundaries,
    data.frame(volume = volume_name, boundary = "Nuclear envelope", radius_um = nuclear_radius),
    data.frame(volume = volume_name, boundary = "Plasma membrane", radius_um = whole_cell$radius_um)
  )
  cytoplasmic_consumption_summary <- rbind(
    cytoplasmic_consumption_summary,
    data.frame(
      volume = volume_name,
      consumption_per_volume_uM_s = volume_weighted_mean(consumption_rate, grid)
    )
  )
  saturable_mean <- volume_weighted_mean(consumption_rate, grid)
  degradation_mean <- volume_weighted_mean(degradation_rate, grid)
  cytoplasmic_sink_summary <- rbind(
    cytoplasmic_sink_summary,
    data.frame(volume = volume_name,
               process = c("Saturable consumption", "First-order degradation"),
               removal_per_volume_uM_s = c(saturable_mean, degradation_mean)),
    data.frame(volume = volume_name, process = "Total removal",
               removal_per_volume_uM_s = saturable_mean + degradation_mean)
  )
  cytoplasmic_sink_profiles <- rbind(
    cytoplasmic_sink_profiles,
    data.frame(volume = volume_name, radius_um = grid$centers_um,
               process = "Saturable consumption", rate_uM_s = consumption_rate),
    data.frame(volume = volume_name, radius_um = grid$centers_um,
               process = "First-order degradation", rate_uM_s = degradation_rate)
  )
}

ggplot(cytoplasmic_profiles,
       aes(radius_um, concentration_uM, colour = volume, linetype = volume)) +
  geom_line(linewidth = 0.8) +
  geom_vline(data = subset(cytoplasmic_boundaries, boundary == "Nuclear envelope"),
             aes(xintercept = radius_um, colour = volume),
             linetype = "dotted", linewidth = 0.5, show.legend = FALSE) +
  geom_vline(data = subset(cytoplasmic_boundaries, boundary == "Plasma membrane"),
             aes(xintercept = radius_um, colour = volume),
             linetype = "dashed", linewidth = 0.5, show.legend = FALSE) +
  scale_colour_manual(values = volume_colours) +
  scale_linetype_manual(values = volume_linetypes) +
  labs(title = "Final cytoplasmic concentration", x = "Radius (um)",
       y = "Concentration (uM)", colour = "Cell size", linetype = "Cell size",
       caption = "Dotted: nuclear envelope. Dashed: plasma membrane.") +
  plot_theme
```

```{r nuclear-sink-profiles, fig.width=8, fig.height=5}
ggplot(cytoplasmic_sink_profiles,
       aes(radius_um, rate_uM_s, colour = volume, linetype = volume)) +
  geom_line(linewidth = 0.8) +
  geom_vline(data = subset(cytoplasmic_boundaries, boundary == "Nuclear envelope"),
             aes(xintercept = radius_um, colour = volume),
             linetype = "dotted", linewidth = 0.5, show.legend = FALSE) +
  geom_vline(data = subset(cytoplasmic_boundaries, boundary == "Plasma membrane"),
             aes(xintercept = radius_um, colour = volume),
             linetype = "dashed", linewidth = 0.5, show.legend = FALSE) +
  facet_wrap(~ process, scales = "free_y") +
  scale_colour_manual(values = volume_colours) +
  scale_linetype_manual(values = volume_linetypes) +
  labs(title = "Separate cytoplasmic sink-rate profiles", x = "Radius (um)",
       y = "Removal rate (uM/s)", colour = "Cell size", linetype = "Cell size",
       caption = "Dotted: nuclear envelope. Dashed: plasma membrane.") +
  plot_theme
```

```{r nuclear-integrated-consumption, fig.width=7, fig.height=4}
ggplot(cytoplasmic_sink_summary,
       aes(volume, removal_per_volume_uM_s, colour = process, group = process)) +
  geom_line(linewidth = 0.8) +
  geom_point(size = 2) +
  labs(title = "Integrated cytoplasmic removal at the final state",
       x = "Cell size", y = "Removal per cytoplasmic volume (uM/s)",
       colour = "Removal term",
       caption = "Total removal is the sum of saturable consumption and first-order degradation.") +
  plot_theme

knitr::kable(cytoplasmic_sink_summary, digits = 6)
```

Nuclear radius grows from 4.50 to 5.67 um and cell radius from 10.0 to 12.60
um. Concentration beside the nuclear envelope rises from about 0.0778 to 0.0972
uM (+25%), whereas concentration near the plasma membrane falls from 0.0384 to
0.0356 uM. The nuclear-to-membrane difference consequently grows from roughly
0.0394 to 0.0616 uM, about 56%, although the volume-averaged concentration
changes by only about 1.8%.

Enlargement substantially steepens the profile but produces essentially no
loss of total processing under the assumed volume-proportional source and sink
capacity. Saturable processing changes from 0.028198 to 0.028166 uM/s per
cytoplasmic volume (about -0.11%); first-order degradation changes from about
0.001802 to 0.001834 uM/s. Total removal is therefore approximately 0.0300
uM/s per cytoplasmic volume in both cells, matching the normalized source.
This is a geometrical scaling effect that creates spatial heterogeneity without
an obvious global functional deficit. The degradation switch at 75 s is a
numerical protocol for reaching the final removal regime; no biological claim
is made about that transient here.
