---
title: "Cell-size scaling: boundary-flux scenarios"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(ggplot2)
source("../R/scaling.R")
source("../R/diffusion.R")
source("../R/plotting.R")
source("../scenarios/external_supply.R")
source("../scenarios/nuclear_processing.R")

volume_multipliers <- c(`1x volume` = 1, `2x volume` = 2)
volume_colours <- c(`1x volume` = "#0072B2", `2x volume` = "#D55E00")
volume_linetypes <- c(`1x volume` = "solid", `2x volume` = "dashed")
plot_theme <- theme_minimal(base_size = 11) + theme(legend.position = "bottom")
```

## External boundary: fixed input per cell volume

Total external influx is proportional to cell volume. Consequently, a larger
cell needs a larger flux density at its membrane (`J = input × volume / area`).
Dotted vertical lines mark the plasma membrane, making the distinct 1× and 2×
radii directly auditable.

```{r external-profiles, fig.width=10, fig.height=8}
external <- external_supply_scenario()
external_profiles <- data.frame()
external_boundaries <- data.frame()
external_consumption_summary <- data.frame()

for (consumer_name in names(external$parameters$consumers)) {
  consumer <- external$parameters$consumers[[consumer_name]]
  for (rate_name in names(external$parameters$input_rates_uM_s)) {
    rate <- external$parameters$input_rates_uM_s[[rate_name]]
    for (volume_name in names(volume_multipliers)) {
      geometry <- spherical_geometry(external$geometry$cell_radius_um,
                                     volume_multipliers[[volume_name]])
      grid <- radial_grid(geometry$radius_um, n_cells = 120)
      surface_flux <- rate * geometry$volume_um3 / geometry$area_um2
      reaction <- if (identical(consumer_name, "michaelis_menten")) {
        function(time_s, concentration) {
          consumer$vmax_uM_s * concentration / (consumer$km_uM + concentration)
        }
      } else NULL
      concentration <- simulate_radial_diffusion(
        grid, external$parameters$diffusion_um2_s, initial = 0, duration_s = 500,
        loss_s = if (identical(consumer_name, "first_order")) consumer$loss_s else 0,
        reaction = reaction,
        outer_boundary = list(type = "fixed_inward_flux", value = surface_flux)
      )
      consumption <- if (identical(consumer_name, "first_order")) {
        consumer$loss_s * concentration
      } else {
        consumer$vmax_uM_s * concentration / (consumer$km_uM + concentration)
      }
      external_profiles <- rbind(
        external_profiles,
        data.frame(consumer = consumer$label, input = tools::toTitleCase(rate_name),
                   volume = volume_name, radius_um = grid$centers_um,
                   quantity = "Concentration (uM)", value = concentration),
        data.frame(consumer = consumer$label, input = tools::toTitleCase(rate_name),
                   volume = volume_name, radius_um = grid$centers_um,
                   quantity = "Consumption (uM/s)", value = consumption)
      )
      external_boundaries <- rbind(
        external_boundaries,
        data.frame(consumer = consumer$label, input = tools::toTitleCase(rate_name),
                   volume = volume_name, cell_radius_um = geometry$radius_um)
      )
      external_consumption_summary <- rbind(
        external_consumption_summary,
        data.frame(consumer = consumer$label, input = tools::toTitleCase(rate_name),
                   volume = volume_name,
                   consumption_per_volume_uM_s = volume_weighted_mean(consumption, grid))
      )
    }
  }
}

plot_external_consumer <- function(consumer_label) {
  ggplot(subset(external_profiles, consumer == consumer_label),
         aes(radius_um, value, colour = volume, linetype = volume)) +
    geom_line(linewidth = 0.8) +
    geom_vline(data = subset(external_boundaries, consumer == consumer_label),
               aes(xintercept = cell_radius_um, colour = volume),
               linetype = "dotted", linewidth = 0.5, show.legend = FALSE) +
    facet_grid(quantity ~ input, scales = "free_y") +
    scale_colour_manual(values = volume_colours) +
    scale_linetype_manual(values = volume_linetypes) +
    labs(title = consumer_label, x = "Radius (um)", y = NULL,
         colour = "Cell size", linetype = "Cell size",
         caption = "Dotted lines: plasma-membrane position.") +
    plot_theme
}

plot_external_consumer("First-order consumption")
```

```{r external-saturable-profiles, fig.width=10, fig.height=8}
plot_external_consumer("Saturable consumption")
```

The first figure uses a linear first-order sink; the second uses a saturable
Michaelis-Menten sink. Both retain the same boundary-input convention, making
the different concentration and consumption gradients attributable to the
reaction term rather than to a changed flux assumption.

### Interpretation and limits of the fixed-flux assumption

At steady state, integrated consumption per cell volume equals the imposed
input per cell volume in these scenarios: there is no accumulation, no export,
and the selected sinks retain sufficient capacity. The broader concentration
range in the larger cell is still meaningful. If the volume-normalized input
is `q`, the required membrane flux density is `J = qR/3`; characteristic
diffusive concentration differences therefore scale approximately as
`JR/D = qR^2/(3D)`. Doubling volume increases this geometric term by
`2^(2/3)`, even while total consumption per volume stays fixed.

The exterior-shell plots use the same far-field concentration `C_bulk` for
both cell sizes. They do **not** make the far-field concentration size
dependent. Rather, for a no-consumption spherical exterior,
`C_bulk - C_surface = JR/D_external = qR^2/(3D_external)`: maintaining the
specified flux requires a larger depletion near the membrane as cells grow or
external diffusion slows. Conversely, if surface concentration were fixed,
the far-field concentration needed to sustain this flux would increase with
`R^2/D_external`. A specified flux can therefore become infeasible when it
would require a negative surface concentration.

The plasma membrane is presently a flux-clamp boundary: the prescribed flux
is imposed on the cell side and removed from the external shell, conserving
mass but allowing a concentration jump across the membrane. It has no
explicit thickness, permeability, partition coefficient, or transporter
kinetics. This is a useful limiting case for strongly regulated or active
uptake, but it is not a passive-membrane model. With fixed transporter density,
capacity would usually scale with membrane area, not cell volume. The separate
Sedlack-inspired oxygen report introduces a permeability-coupled membrane
formulation in which flux is an outcome rather than an input.

```{r external-integrated-consumption, fig.width=8, fig.height=4}
ggplot(external_consumption_summary,
       aes(input, consumption_per_volume_uM_s, colour = volume, linetype = volume,
           group = volume)) +
  geom_line(linewidth = 0.8) +
  geom_point(size = 2) +
  facet_wrap(~ consumer) +
  scale_colour_manual(values = volume_colours) +
  scale_linetype_manual(values = volume_linetypes) +
  labs(title = "Integrated cellular consumption: external-boundary scenarios",
       x = "Input rate per cell volume", y = "Consumption per cell volume (uM/s)",
       colour = "Cell size", linetype = "Cell size") +
  plot_theme
```

## Cell plus exterior medium

The next figure shows the cell interior and an extracellular shell. Flux is
fixed at the membrane and the external medium has no consumption. The outer
shell uses the spherical far-field condition `C -> C_bulk` and `dC/dr -> 0`
as `r -> infinity`, implemented at a finite radius with the corresponding
Robin condition.

```{r interior-exterior, fig.width=10, fig.height=5}
rate <- external$parameters$input_rates_uM_s[["high"]]
# Keep intracellular diffusion fixed. Only the extracellular medium changes.
media <- list(
  `External diffusion slower` = c(inside = 20, outside = 5),
  `External diffusion faster` = c(inside = 20, outside = 50)
)
combined_profiles <- data.frame()
medium_boundaries <- data.frame()
medium_consumption_summary <- data.frame()

for (medium_name in names(media)) {
  for (volume_name in names(volume_multipliers)) {
    geometry <- spherical_geometry(external$geometry$cell_radius_um,
                                   volume_multipliers[[volume_name]])
    flux <- rate * geometry$volume_um3 / geometry$area_um2
    inside_grid <- radial_grid(geometry$radius_um, 120)
    inside <- simulate_radial_diffusion(
      inside_grid, media[[medium_name]][["inside"]], 0, 500,
      loss_s = external$parameters$consumers$first_order$loss_s,
      outer_boundary = list(type = "fixed_inward_flux", value = flux)
    )
    outside_grid <- radial_grid(external$parameters$external_medium_radius_um,
                                180, inner_radius_um = geometry$radius_um)
    outside <- simulate_radial_diffusion(
      outside_grid, media[[medium_name]][["outside"]],
      initial = external$parameters$external_bulk_concentration_uM, duration_s = 10000,
      inner_boundary = list(type = "fixed_outward_flux", value = -flux),
      outer_boundary = list(type = "infinite_reservoir",
                            value = external$parameters$external_bulk_concentration_uM)
    )
    combined_profiles <- rbind(
      combined_profiles,
      data.frame(medium = medium_name, volume = volume_name, region = "Cell interior",
                 radius_um = inside_grid$centers_um, concentration_uM = inside),
      data.frame(medium = medium_name, volume = volume_name, region = "External medium",
                 radius_um = outside_grid$centers_um, concentration_uM = outside)
    )
    medium_boundaries <- rbind(
      medium_boundaries,
      data.frame(medium = medium_name, volume = volume_name,
                 cell_radius_um = geometry$radius_um)
    )
    medium_consumption_summary <- rbind(
      medium_consumption_summary,
      data.frame(medium = medium_name, volume = volume_name,
                 consumption_per_volume_uM_s = volume_weighted_mean(
                   external$parameters$consumers$first_order$loss_s * inside, inside_grid
                 ))
    )
  }
}

ggplot(combined_profiles,
       aes(radius_um, concentration_uM, colour = volume, linetype = volume,
           group = interaction(volume, region))) +
  geom_line(linewidth = 0.8) +
  geom_vline(data = medium_boundaries,
             aes(xintercept = cell_radius_um, colour = volume),
             linetype = "dotted", linewidth = 0.5, show.legend = FALSE) +
  facet_wrap(~ medium) +
  scale_colour_manual(values = volume_colours) +
  scale_linetype_manual(values = volume_linetypes) +
  labs(title = "Internal and external concentration gradients",
       x = "Radius from cell centre (um)", y = "Concentration (uM)",
       colour = "Cell size", linetype = "Cell size",
       caption = "Dotted lines: plasma membrane. Each line breaks at the membrane.") +
  plot_theme
```

```{r interior-exterior-integrated-consumption, fig.width=7, fig.height=4}
ggplot(medium_consumption_summary,
       aes(volume, consumption_per_volume_uM_s, colour = medium, group = medium)) +
  geom_line(linewidth = 0.8) +
  geom_point(size = 2) +
  labs(title = "Integrated cellular consumption: exterior-medium scenarios",
       x = "Cell size", y = "Consumption per cell volume (uM/s)",
       colour = "External diffusion condition") +
  plot_theme
```

## Internal boundary: nuclear export with two cytoplasmic sinks

Nuclear export supplies the diffusing species at the nuclear envelope. It is
removed through two distinct cytoplasmic processes: saturable consumption
`Vmax × C / (Km + C)` and first-order degradation `k × C`. The plots below
show the final state after the prescribed degradation-rate change. Dotted
vertical lines mark the nuclear envelope and dashed lines mark the plasma
membrane. Their locations scale with `V^(1/3)`.

```{r nuclear-processing, fig.width=8, fig.height=5}
nuclear <- nuclear_processing_scenario()
cytoplasmic_profiles <- data.frame()
cytoplasmic_boundaries <- data.frame()
cytoplasmic_consumption_summary <- data.frame()
cytoplasmic_sink_profiles <- data.frame()
for (volume_name in names(volume_multipliers)) {
  whole_cell <- spherical_geometry(nuclear$geometry$cell_radius_um,
                                  volume_multipliers[[volume_name]])
  nuclear_radius <- whole_cell$radius_um * nuclear$geometry$nucleus_radius_fraction
  grid <- radial_grid(whole_cell$radius_um, 120, inner_radius_um = nuclear_radius)
  cytoplasm_volume <- sum(grid$volumes_um3)
  nuclear_area <- 4 * pi * nuclear_radius ^ 2
  export_flux <- nuclear$parameters$source_rate * cytoplasm_volume / nuclear_area
  degradation <- function(time_s) {
    ifelse(time_s < nuclear$parameters$degradation_transition_s,
           nuclear$parameters$initial_degradation_s,
           nuclear$parameters$final_degradation_s)
  }
  concentration <- simulate_radial_diffusion(
    grid, nuclear$parameters$diffusion_um2_s, 0, 200,
    loss_s = degradation,
    reaction = function(time_s, concentration) {
      nuclear$parameters$consumption_vmax_uM_s * concentration /
        (nuclear$parameters$consumption_km_uM + concentration)
    },
    inner_boundary = list(type = "fixed_outward_flux", value = export_flux)
  )
  consumption_rate <- nuclear$parameters$consumption_vmax_uM_s * concentration /
    (nuclear$parameters$consumption_km_uM + concentration)
  degradation_rate <- nuclear$parameters$final_degradation_s * concentration
  cytoplasmic_profiles <- rbind(
    cytoplasmic_profiles,
    data.frame(volume = volume_name, radius_um = grid$centers_um,
               concentration_uM = concentration)
  )
  cytoplasmic_boundaries <- rbind(
    cytoplasmic_boundaries,
    data.frame(volume = volume_name, boundary = "Nuclear envelope", radius_um = nuclear_radius),
    data.frame(volume = volume_name, boundary = "Plasma membrane", radius_um = whole_cell$radius_um)
  )
  cytoplasmic_consumption_summary <- rbind(
    cytoplasmic_consumption_summary,
    data.frame(
      volume = volume_name,
      consumption_per_volume_uM_s = volume_weighted_mean(consumption_rate, grid)
    )
  )
  cytoplasmic_sink_profiles <- rbind(
    cytoplasmic_sink_profiles,
    data.frame(volume = volume_name, radius_um = grid$centers_um,
               process = "Saturable consumption", rate_uM_s = consumption_rate),
    data.frame(volume = volume_name, radius_um = grid$centers_um,
               process = "First-order degradation", rate_uM_s = degradation_rate)
  )
}

ggplot(cytoplasmic_profiles,
       aes(radius_um, concentration_uM, colour = volume, linetype = volume)) +
  geom_line(linewidth = 0.8) +
  geom_vline(data = subset(cytoplasmic_boundaries, boundary == "Nuclear envelope"),
             aes(xintercept = radius_um, colour = volume),
             linetype = "dotted", linewidth = 0.5, show.legend = FALSE) +
  geom_vline(data = subset(cytoplasmic_boundaries, boundary == "Plasma membrane"),
             aes(xintercept = radius_um, colour = volume),
             linetype = "dashed", linewidth = 0.5, show.legend = FALSE) +
  scale_colour_manual(values = volume_colours) +
  scale_linetype_manual(values = volume_linetypes) +
  labs(title = "Final cytoplasmic concentration", x = "Radius (um)",
       y = "Concentration (uM)", colour = "Cell size", linetype = "Cell size",
       caption = "Dotted: nuclear envelope. Dashed: plasma membrane.") +
  plot_theme
```

```{r nuclear-sink-profiles, fig.width=8, fig.height=5}
ggplot(cytoplasmic_sink_profiles,
       aes(radius_um, rate_uM_s, colour = volume, linetype = volume)) +
  geom_line(linewidth = 0.8) +
  geom_vline(data = subset(cytoplasmic_boundaries, boundary == "Nuclear envelope"),
             aes(xintercept = radius_um, colour = volume),
             linetype = "dotted", linewidth = 0.5, show.legend = FALSE) +
  geom_vline(data = subset(cytoplasmic_boundaries, boundary == "Plasma membrane"),
             aes(xintercept = radius_um, colour = volume),
             linetype = "dashed", linewidth = 0.5, show.legend = FALSE) +
  facet_wrap(~ process, scales = "free_y") +
  scale_colour_manual(values = volume_colours) +
  scale_linetype_manual(values = volume_linetypes) +
  labs(title = "Separate cytoplasmic sink-rate profiles", x = "Radius (um)",
       y = "Removal rate (uM/s)", colour = "Cell size", linetype = "Cell size",
       caption = "Dotted: nuclear envelope. Dashed: plasma membrane.") +
  plot_theme
```

```{r nuclear-integrated-consumption, fig.width=6, fig.height=4}
ggplot(cytoplasmic_consumption_summary,
       aes(volume, consumption_per_volume_uM_s, group = 1)) +
  geom_line(linewidth = 0.8, colour = "#009E73") +
  geom_point(size = 2, colour = "#009E73") +
  labs(title = "Integrated saturable cytoplasmic consumption",
       x = "Cell size", y = "Consumption per cytoplasmic volume (uM/s)",
       caption = "Includes only the saturable consumption term; first-order degradation is excluded.") +
  plot_theme
```
