---
title: "Cell size and wave-pinning polarity"
output:
  html_document:
    toc: true
    toc_float: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE)

library(ggplot2)
library(deSolve)

volume_colours <- c(`1x volume` = "#0072B2", `2x volume` = "#D55E00")
plot_theme <- theme_minimal(base_size = 11) +
  theme(
    legend.position = "bottom",
    panel.grid.minor = element_blank()
  )
```

[&larr; Report catalogue](../index.html)

> **Status: published minimal model + new size sweep.** The reaction--diffusion
> mechanism follows Mori, Jilkine, and Edelstein-Keshet; varying domain size
> under fixed kinetic assumptions is the experiment performed here.

The exact data used in every plot are exported to `docs/data/` when this report
is rendered inside the project repository. The complete report source is
intended to live in `reports/`.

# Question

Can changing cell size alone alter the ability of an otherwise unchanged
intracellular polarity network to generate and maintain a polarized state?

This report uses the minimal **wave-pinning** reaction--diffusion model
introduced by Mori, Jilkine, and Edelstein-Keshet. The model represents a
generic Rho-family GTPase that cycles between a slowly diffusing active,
membrane-associated state and a rapidly diffusing inactive, cytosolic state.
Positive feedback in activation creates bistable local kinetics; conservation
of the total GTPase pool and the large diffusion-rate difference allow an
activation wave to advance and then stall, leaving a persistent front/back
pattern.

The model is useful here because cell size enters directly through the spatial
domain. We therefore change **only domain length**, keeping concentrations,
reaction rates, diffusion coefficients, and the initiating stimulus fixed.

This is a deliberately stripped-down scaling test. It is not a model of
whole-cell migration.

# Model

We use a one-dimensional domain \(0 < x < L\), interpreted as a strip across a
cell diameter. The active form \(a(x,t)\) diffuses slowly and the inactive form
\(b(x,t)\) diffuses rapidly:

\[
\frac{\partial a}{\partial t}
=
D_a\frac{\partial^2 a}{\partial x^2}
+
f(a,b;x,t)
\]

\[
\frac{\partial b}{\partial t}
=
D_b\frac{\partial^2 b}{\partial x^2}
-
f(a,b;x,t).
\]

The interconversion term is

\[
f(a,b;x,t)
=
b\left[
k_0
+
\gamma\frac{a^2}{K^2+a^2}
+
k_s(x,t)
\right]
-
\delta a.
\]

The Hill term gives positive feedback: active GTPase promotes further
activation. The transient stimulus \(k_s(x,t)\) locally increases activation at
the left edge of the cell but neither creates nor destroys total GTPase.
No-flux boundary conditions are imposed at both ends, so the spatially
integrated \(a+b\) pool is conserved.

The parameter set below follows the commonly used dimensional wave-pinning
regime reported in the original Mori model and subsequent deterministic
implementations.

```{r parameters}
pars <- list(
  D_a = 0.1,        # um^2 / s
  D_b = 10,         # um^2 / s
  k0 = 0.067,       # 1 / s
  gamma = 1,        # 1 / s
  K = 1,            # uM
  delta = 1,        # 1 / s
  a0 = 0.2683,      # uM
  b0 = 2,           # uM
  stimulus_rate = 10,      # 1 / s
  stimulus_start = 50,     # s
  stimulus_end = 70,       # s
  stimulus_width_um = 0.4, # um
  baseline_length_um = 10, # um
  end_time_s = 300
)

parameter_table <- data.frame(
  Parameter = c(
    "D_a", "D_b", "k0", "gamma", "K", "delta",
    "a(x,0)", "b(x,0)", "stimulus", "stimulus window",
    "stimulated region", "baseline L"
  ),
  Value = c(
    "0.1 um^2/s", "10 um^2/s", "0.067 /s", "1 /s", "1 uM", "1 /s",
    "0.2683 uM", "2 uM", "10 /s", "50-70 s",
    "0-0.4 um", "10 um"
  )
)

knitr::kable(parameter_table, align = c("l", "l"))
```

## Numerical implementation

The PDE is discretized by a cell-centred finite-volume method. Zero diffusive
flux is imposed at the two boundaries, and the resulting method-of-lines
system is integrated with `deSolve::lsodes`. The finite-volume stencil is
conservative: diffusion redistributes material between neighboring bins but
does not change the total amount.

```{r solver}
laplacian_no_flux <- function(y, dx) {
  n <- length(y)
  out <- numeric(n)

  out[1] <- (y[2] - y[1]) / dx^2
  out[n] <- (y[n - 1] - y[n]) / dx^2

  if (n > 2) {
    out[2:(n - 1)] <-
      (y[1:(n - 2)] - 2 * y[2:(n - 1)] + y[3:n]) / dx^2
  }

  out
}

simulate_wave_pinning <- function(
    length_um,
    parameters = pars,
    n_cells = 81,
    output_times = seq(0, parameters$end_time_s, by = 1)) {

  dx <- length_um / n_cells
  x <- (seq_len(n_cells) - 0.5) * dx

  state0 <- c(
    setNames(rep(parameters$a0, n_cells), sprintf("a_%03d", seq_len(n_cells))),
    setNames(rep(parameters$b0, n_cells), sprintf("b_%03d", seq_len(n_cells)))
  )

  rhs <- function(time, state, controls) {
    a <- state[seq_len(n_cells)]
    b <- state[n_cells + seq_len(n_cells)]

    stimulus <- if (controls$stimulus_on) {
      ifelse(x <= parameters$stimulus_width_um,
             parameters$stimulus_rate, 0)
    } else {
      rep(0, n_cells)
    }

    hill <- a^2 / (parameters$K^2 + a^2)
    reaction <-
      b * (parameters$k0 + parameters$gamma * hill + stimulus) -
      parameters$delta * a

    da <-
      parameters$D_a * laplacian_no_flux(a, dx) +
      reaction

    db <-
      parameters$D_b * laplacian_no_flux(b, dx) -
      reaction

    list(c(da, db))
  }

  run_stage <- function(state, times, stimulus_on) {
    times <- sort(unique(times))
    if (length(times) < 2) {
      stop("Each integration stage must contain at least two time points.")
    }

    ode(
      y = state,
      times = times,
      func = rhs,
      parms = list(stimulus_on = stimulus_on),
      method = "lsodes",
      rtol = 1e-7,
      atol = 1e-9
    )
  }

  add_bounds <- function(values, lower, upper) {
    sort(unique(c(lower, values[values > lower & values < upper], upper)))
  }

  t0 <- 0
  ts <- parameters$stimulus_start
  te <- parameters$stimulus_end
  tf <- max(output_times)

  times_1 <- add_bounds(output_times, t0, ts)
  stage_1 <- run_stage(state0, times_1, FALSE)

  state_1 <- as.numeric(stage_1[nrow(stage_1), -1])
  names(state_1) <- names(state0)

  times_2 <- add_bounds(output_times, ts, te)
  stage_2 <- run_stage(state_1, times_2, TRUE)

  state_2 <- as.numeric(stage_2[nrow(stage_2), -1])
  names(state_2) <- names(state0)

  times_3 <- add_bounds(output_times, te, tf)
  stage_3 <- run_stage(state_2, times_3, FALSE)

  combined <- rbind(
    stage_1,
    stage_2[-1, , drop = FALSE],
    stage_3[-1, , drop = FALSE]
  )

  combined <- as.data.frame(combined)
  combined <- combined[combined$time %in% output_times, , drop = FALSE]

  a_cols <- grep("^a_", names(combined))
  b_cols <- grep("^b_", names(combined))

  total_mean <- rowMeans(combined[, a_cols, drop = FALSE]) +
    rowMeans(combined[, b_cols, drop = FALSE])

  mass_error <- max(abs(total_mean - total_mean[1]))

  list(
    length_um = length_um,
    x_um = x,
    dx_um = dx,
    result = combined,
    a_cols = a_cols,
    b_cols = b_cols,
    mass_error = mass_error
  )
}

profile_data <- function(simulation, times = NULL) {
  result <- simulation$result
  if (!is.null(times)) {
    result <- result[result$time %in% times, , drop = FALSE]
  }

  do.call(
    rbind,
    lapply(seq_len(nrow(result)), function(i) {
      data.frame(
        time_s = result$time[i],
        x_um = simulation$x_um,
        active_uM = as.numeric(result[i, simulation$a_cols]),
        inactive_uM = as.numeric(result[i, simulation$b_cols])
      )
    })
  )
}

polarity_score <- function(simulation, front_fraction = 0.2) {
  result <- simulation$result
  final_a <- as.numeric(result[nrow(result), simulation$a_cols])
  x <- simulation$x_um
  L <- simulation$length_um

  front <- mean(final_a[x <= front_fraction * L])
  back <- mean(final_a[x >= (1 - front_fraction) * L])

  (front - back) / (front + back)
}
```

# 1. Wave pinning in the reference 10-um cell

The first simulation is simply a sanity check that the implementation exhibits
the intended behavior. The cell starts spatially uniform. A local activation
pulse is applied from 50 to 70 s over the leftmost 0.4 um. Active GTPase rises
locally, the high-activity region advances, and after the stimulus is removed
the front stalls rather than relaxing back to a homogeneous state.

```{r reference-simulation}
reference_times <- seq(0, pars$end_time_s, by = 1)

reference_sim <- simulate_wave_pinning(
  length_um = pars$baseline_length_um,
  output_times = reference_times
)

stopifnot(reference_sim$mass_error < 1e-5)

display_times <- c(0, 50, 55, 70, 100, 200, 300)
reference_profiles <- profile_data(reference_sim, display_times)

reference_profiles$time_label <- factor(
  paste0(reference_profiles$time_s, " s"),
  levels = paste0(display_times, " s")
)
```

```{r reference-plot, fig.width=9, fig.height=6}
ggplot(reference_profiles, aes(x_um, active_uM)) +
  geom_line(linewidth = 0.85) +
  facet_wrap(~ time_label, ncol = 4) +
  labs(
    title = "A transient local stimulus produces a pinned polarity front",
    subtitle = "Active membrane-associated GTPase",
    x = "Position across cell (um)",
    y = "Active GTPase (uM)"
  ) +
  plot_theme
```

The important feature is persistence after the stimulus disappears. The
stimulus ends at 70 s, but the final state at 300 s remains strongly
asymmetric. The front/back polarity score in the reference simulation is
`r sprintf("%.3f", polarity_score(reference_sim))`.

This is the wave-pinning mechanism: local positive feedback launches an
activation front, while depletion and rapid redistribution of the inactive
pool prevent the front from consuming the whole domain.

# 2. Doubling cell volume

For geometrically similar cells, length scales with the cube root of volume:

\[
L(V) = L_0 V^{1/3}.
\]

A 2x-volume cell is therefore only \(2^{1/3} \approx 1.26\) times longer along
each linear dimension. Starting from the 10-um reference cell gives a
2x-volume domain of approximately 12.6 um.

Nothing else changes. Initial concentrations, total concentration per unit
volume, kinetic parameters, diffusion coefficients, and the physical 0.4-um
stimulus width are identical.

```{r volume-comparison}
volume_multipliers <- c(`1x volume` = 1, `2x volume` = 2)

comparison_sims <- lapply(
  volume_multipliers,
  function(v) {
    simulate_wave_pinning(
      length_um = pars$baseline_length_um * v^(1 / 3),
      output_times = c(0, pars$stimulus_start, pars$stimulus_end, pars$end_time_s)
    )
  }
)

comparison_profiles <- do.call(
  rbind,
  lapply(names(comparison_sims), function(label) {
    sim <- comparison_sims[[label]]
    final <- profile_data(sim, pars$end_time_s)
    final$volume <- label
    final$cell_length_um <- sim$length_um
    final$relative_position <- final$x_um / sim$length_um
    final
  })
)

comparison_scores <- data.frame(
  volume = names(comparison_sims),
  volume_multiplier = unname(volume_multipliers),
  cell_length_um = vapply(comparison_sims, function(x) x$length_um, numeric(1)),
  polarity_score = vapply(comparison_sims, polarity_score, numeric(1))
)
```

```{r volume-comparison-plot, fig.width=8, fig.height=5}
ggplot(
  comparison_profiles,
  aes(x_um, active_uM, colour = volume, linetype = volume)
) +
  geom_line(linewidth = 0.9) +
  scale_colour_manual(values = volume_colours) +
  scale_linetype_manual(values = c(`1x volume` = "solid", `2x volume` = "dashed")) +
  labs(
    title = "Final pinned polarity profile: 1x versus 2x cell volume",
    x = "Physical position across cell (um)",
    y = "Active GTPase (uM)",
    colour = "Cell size",
    linetype = "Cell size"
  ) +
  plot_theme
```

The 1x cell has a final polarity score of
`r sprintf("%.3f", comparison_scores$polarity_score[comparison_scores$volume == "1x volume"])`;
the 2x-volume cell has a score of
`r sprintf("%.3f", comparison_scores$polarity_score[comparison_scores$volume == "2x volume"])`.

Under this published parameter regime, **both cells are comfortably
polarizable**. Doubling volume does not qualitatively switch the system into a
new state. The larger domain slightly strengthens the spatial separation, but
the effect is modest compared with the existence of wave pinning itself.

That is an important null result for the WGD scaling question. A simple
1.26-fold increase in linear size is not automatically sufficient to disrupt
or radically reorganize this canonical polarity circuit.

# 3. Size sweep

The more informative scaling exercise is to ask where changing domain size
actually becomes consequential. We therefore sweep cell volume over more than
two orders of magnitude while holding the molecular system fixed.

The plotted score compares the mean active GTPase in the front 20% of the cell
with that in the back 20%:

\[
P =
\frac{\bar a_{\rm front} - \bar a_{\rm back}}
     {\bar a_{\rm front} + \bar a_{\rm back}}.
\]

A score near zero is unpolarized; larger positive values indicate stronger
front/back segregation.

```{r size-sweep}
sweep_volumes <- sort(unique(c(
  exp(seq(log(0.01), log(4), length.out = 34)),
  1,
  2
)))

sweep_rows <- lapply(sweep_volumes, function(v) {
  L <- pars$baseline_length_um * v^(1 / 3)

  sim <- simulate_wave_pinning(
    length_um = L,
    output_times = c(0, pars$stimulus_start, pars$stimulus_end, pars$end_time_s)
  )

  data.frame(
    volume_multiplier = v,
    cell_length_um = L,
    polarity_score = polarity_score(sim),
    mass_error = sim$mass_error
  )
})

size_sweep <- do.call(rbind, sweep_rows)

stopifnot(max(size_sweep$mass_error) < 1e-5)
```

```{r size-sweep-plot, fig.width=8, fig.height=5}
ggplot(size_sweep, aes(volume_multiplier, polarity_score)) +
  geom_line(linewidth = 0.85) +
  geom_point(size = 1.8) +
  geom_vline(
    xintercept = c(1, 2),
    linetype = c("solid", "dashed"),
    linewidth = 0.55
  ) +
  scale_x_log10(
    breaks = c(0.01, 0.03, 0.1, 0.3, 1, 2, 4),
    labels = c("0.01x", "0.03x", "0.1x", "0.3x", "1x", "2x", "4x")
  ) +
  labs(
    title = "Fixed polarity kinetics impose a cell-size dependence",
    subtitle = "Vertical lines mark the 1x and 2x-volume cells",
    x = "Cell-volume multiplier",
    y = "Final front/back polarity score"
  ) +
  plot_theme
```

The sweep exposes a regime in which sufficiently small domains cannot sustain
the same spatial separation. The reason is straightforward: diffusion operates
over a characteristic time proportional to \(L^2/D\), whereas the local
reaction timescale is fixed. Shrinking the cell therefore makes membrane
diffusion increasingly effective at erasing spatial differences before the
reaction network can maintain them.

Equivalently, increasing volume from 1x to 2x increases the characteristic
membrane diffusion time across the cell by

\[
\left(2^{1/3}\right)^2 = 2^{2/3} \approx 1.59.
\]

Thus cell size changes the relative balance between spatial mixing and local
reaction kinetics even though no molecular parameter itself is changed.

This general mechanism is consistent with later work showing that several
mass-conserved polarity networks, including wave pinning, possess a critical
polarizable system size when their reaction and diffusion parameters fail to
scale with cell size.

# What this does -- and does not -- imply for WGD

This minimal example establishes a useful physical principle:

> A fixed reaction--diffusion polarity circuit is not scale invariant. Changing
> cell size changes the ratio of diffusion and reaction timescales and can,
> in principle, move the same molecular network between weakly polarized and
> strongly polarized regimes.

However, the specific 1x-to-2x WGD-like comparison in this parameterization is
not dramatic. Both the 10-um and 12.6-um domains lie well inside the polarized
regime. The model therefore **does not by itself support a claim that WGD
should strongly impair polarity or generate multiple fronts**.

To make that connection biologically, at least one additional ingredient would
be needed, for example:

- a polarity network whose physiological operating point lies closer to a
  size-dependent bifurcation;
- explicit membrane-to-cytosol geometry and surface-to-volume scaling;
- multiple interacting polarity species;
- coupling to actin, protrusions, or cell shape;
- or experimental evidence that a relevant polarity length scale fails to
  increase with WGD-associated cell size.

Those extensions are deliberately outside this sandbox demo.

# Numerical and biological limitations

1. **One-dimensional geometry.** The model is a strip across a static cell, not
   a deforming 2D or 3D membrane.
2. **No mechanics or migration.** A polarized signaling profile is not itself a
   migration trajectory.
3. **No explicit membrane/bulk geometry.** Active and inactive species are
   represented as 1D concentration fields with different diffusion
   coefficients rather than as a true membrane--cytosol bulk/surface PDE.
4. **Concentration homeostasis is assumed.** Initial concentrations are held
   fixed as size changes, so total molecular abundance scales with the
   unmodeled cross-sectional size of the cell.
5. **The stimulus is fixed in physical width.** The leftmost 0.4 um is
   stimulated at every cell size. This isolates domain-size scaling but is only
   one possible biological assumption.
6. **Parameters are illustrative, not fitted to WGD cells.** The purpose is to
   expose the scaling mechanism, not to predict a particular cancer cell line.

# Export plot data

```{r export-data, include=FALSE}
data_dir <- "../docs/data"

if (dir.exists("../docs")) {
  dir.create(data_dir, recursive = TRUE, showWarnings = FALSE)

  write.csv(
    reference_profiles,
    file.path(data_dir, "wave_pinning_timecourse.csv"),
    row.names = FALSE
  )

  write.csv(
    comparison_profiles,
    file.path(data_dir, "wave_pinning_volume_comparison.csv"),
    row.names = FALSE
  )

  write.csv(
    size_sweep,
    file.path(data_dir, "wave_pinning_size_sweep.csv"),
    row.names = FALSE
  )
}
```

# References

Mori Y, Jilkine A, Edelstein-Keshet L. **Wave-Pinning and Cell Polarity from a
Bistable Reaction-Diffusion System.** *Biophysical Journal* 94:3684-3697
(2008). DOI: 10.1529/biophysj.107.120824.

Mori Y, Jilkine A, Edelstein-Keshet L. **Asymptotic and Bifurcation Analysis
of Wave-Pinning in a Reaction-Diffusion Model for Cell Polarization.**
*SIAM Journal on Applied Mathematics* 71:1401-1427 (2011).
DOI: 10.1137/10079118X.

Walther GR, Marée AFM, Edelstein-Keshet L, Grieneisen VA.
**Deterministic Versus Stochastic Cell Polarisation Through Wave-Pinning.**
*Bulletin of Mathematical Biology* 74:2570-2599 (2012).
DOI: 10.1007/s11538-012-9766-5. The explicit localized transient stimulus,
initial concentrations, domain length, and dimensional parameter values used
here follow the deterministic implementation reported in this study.

Hubatsch L, Peglion F, Reich JD, et al. **A cell size threshold limits cell
polarity and asymmetric division potential.** *Nature Physics* 15:1078-1085
(2019). DOI: 10.1038/s41567-019-0601-x.
