---
title: "An Eroumé--Marée polarity model under cell-size scaling"
subtitle: "Exact VCML reaction-system port with a new 0.5x/1x/2x geometry experiment"
output:
  html_document:
    toc: true
    toc_float: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE,
                      fig.align = "center", dpi = 110)
library(ggplot2)
library(gridExtra)
source("../R/wgd_spatial_analysis.R")
snapshot <- read_wgd_snapshot("assets/wgd_spatial/v1")
validate_wgd_snapshot_values(snapshot)
summary_table <- snapshot$summary
metrics <- snapshot$dynamics
conservation <- snapshot$conservation
boundary_profiles <- snapshot$boundary
longitudinal_profiles <- snapshot$longitudinal
spatial_fields <- snapshot$spatial
geometry_colours <- wgd_geometry_colours()
plot_theme <- theme_minimal(base_size = 11) +
  theme(panel.grid.minor = element_blank(), legend.position = "bottom",
        plot.title = element_text(face = "bold"))
```

[&larr; Report catalogue](../index.html)

> **Status: published VCell model ported + WGD-oriented size extension.** The
> nine-PDE reaction system and parameters are from the Eroumé et al. public
> VCell model; changing cell size while holding concentrations and kinetics
> fixed is a new experiment in this repository.

The exact plot-level data used here are exported to `docs/data/` during the
site build. The completed numerical experiment is stored as a small,
checksummed reporting snapshot. Rebuilding this page validates and reads that
snapshot; it does not compile the numerical kernel or execute the PDE solver.

# Question

Does cell enlargement alone change the dynamics or stability of a spatial
Rac/Cdc42/Rho polarity circuit when concentrations, kinetic constants,
diffusion coefficients, shape, and relative stimulus position are otherwise
held fixed?

# Model provenance: what is published and what is new

Marée et al. (2012) developed the parent polarity model: active and inactive
Cdc42, Rac, and Rho interact with PIP, PIP2, and PIP3 feedback and, in the full
model, cytoskeletal and cell-shape dynamics. That paper is the foundational
lineage; it is not separately reimplemented here.

Eroumé et al. (2021) implemented the Marée signalling system in Virtual Cell
to study polarization on fixed cell shapes. Because shape was prescribed,
they omitted the cytoskeletal module. Their extended fixed-shape model has
nine PDE variables: active and inactive Cdc42, Rac, and Rho, plus PIP, PIP2,
and PIP3. Eroumé et al. state that its parameter values are those of Marée et
al. and released the VCell models publicly, including
`Kerbai_PLoSone_2021_teardrop_polarization_extended`.

For this repository, that Eroumé VCML model was obtained and parsed, and its
reaction system was ported without redesigning the biochemical network. The
new contribution is the controlled size experiment below: geometrically
matched 0.5x, 1x, and 2x implied volumes at two aspect ratios. Eroumé et al.
studied shape; they did **not** report this size-scaling experiment.

# Experimental design

Six simulations cross two teardrop aspect ratios with three implied 3D volume
classes. Linear dimensions scale with the cube root of volume. The original
VCML cell mask is stretched onto an 80 by 80 mesh, while physical diffusion
coefficients and reaction parameters remain unchanged.

```{r design-table}
knitr::kable(
  summary_table[c("geometry", "size", "length_um", "width_um", "aspect_ratio")],
  col.names = c("Geometry", "Volume", "Length (um)", "Width (um)", "Aspect ratio"),
  digits = 2, row.names = FALSE
)
```

The system was integrated from 0 to 200 s with a 0.02 s timestep. A spatial
input is present through 10 s and then becomes uniform. Thus later structure
tests whether the network maintains or reorganizes the induced polarity.

# Polarization dynamics

```{r dynamics, fig.width=11, fig.height=7}
contrast_long <- rbind(
  data.frame(metrics[c("geometry", "time")], species = "Cdc42",
             contrast = metrics$cdc42_contrast),
  data.frame(metrics[c("geometry", "time")], species = "Rac",
             contrast = metrics$rac_contrast)
)
p_contrast <- ggplot(contrast_long,
  aes(time, contrast, colour = geometry)) +
  geom_vline(xintercept = 10, linetype = 3, colour = "grey45") +
  geom_line(linewidth = 0.8) + facet_wrap(~species, ncol = 1, scales = "free_y") +
  scale_colour_manual(values = geometry_colours) +
  labs(title = "Front--back polarization dynamics",
       subtitle = "Dotted line: spatial stimulus becomes uniform",
       x = "Time (s)", y = "(front - back) / (front + back)", colour = NULL) +
  plot_theme

p_separation <- ggplot(metrics,
  aes(time, rho_separation, colour = geometry)) +
  geom_vline(xintercept = 10, linetype = 3, colour = "grey45") +
  geom_line(linewidth = 0.8) + scale_colour_manual(values = geometry_colours) +
  labs(title = "Cdc42--Rho spatial separation", x = "Time (s)",
       y = "Centroid separation (cell lengths)", colour = NULL) + plot_theme

gridExtra::grid.arrange(p_contrast, p_separation, ncol = 2, widths = c(1.15, 1))
```

The larger cells establish polarity later. More importantly, the two 2x-volume
geometries retain multiple Cdc42 peaks for most of the post-stimulus interval,
whereas the smaller cells settle into one peak.

```{r result-table}
display_summary <- summary_table[c(
  "geometry", "time_to_90pct_cdc42_polarity_s",
  "final_cdc42_front_back", "final_rho_separation_cell_lengths",
  "cdc42_multi_peak_duration_s", "unique_state_stable"
)]
knitr::kable(display_summary,
  col.names = c("Geometry", "90% polarity (s)", "Final Cdc42 front:back",
                "Cdc42--Rho separation", "Multi-peak duration (s)",
                "Stable unique state"),
  digits = 3, row.names = FALSE)
```

# Spatial organization

```{r final-fields, fig.width=11, fig.height=11}
final_fields <- subset(spatial_fields, time == 200)
final_fields$species <- factor(final_fields$species,
  levels = c("cdc42_act", "Rac_act", "Rho_act"),
  labels = c("Cdc42 active", "Rac active", "Rho active"))
ggplot(final_fields, aes(x, y, fill = value)) +
  geom_tile(aes(width = dx, height = dy)) + coord_fixed() +
  facet_grid(geometry ~ species) + scale_fill_viridis_c(option = "magma") +
  labs(title = "Spatial fields at 200 s", x = "x (um)", y = "y (um)",
       fill = "uM") + plot_theme + theme(legend.position = "right")
```

```{r longitudinal, fig.width=12, fig.height=7}
profile_plot_data <- subset(longitudinal_profiles, time %in% c(10, 50, 200))
profile_plot_data$species <- factor(profile_plot_data$species,
  levels = c("cdc42_act", "Rac_act", "Rho_act"),
  labels = c("Cdc42 active", "Rac active", "Rho active"))
ggplot(profile_plot_data,
  aes(relative_x, value, colour = geometry)) +
  geom_line(linewidth = 0.7) + facet_grid(species ~ time, scales = "free_y") +
  scale_colour_manual(values = geometry_colours) +
  labs(title = "Long-axis concentration profiles",
       x = "Relative position from rear to front", y = "Mean concentration (uM)",
       colour = NULL) + plot_theme
```

# Numerical checks and interpretation

The maximum relative drift in each conserved GTPase pool is below
`r format(max(conservation$max_relative_drift), scientific = TRUE, digits = 2)`.
This makes numerical mass loss an implausible explanation for the size-dependent
states.

Within this model, enlargement does more than slow the approach to the same
solution: it changes the spatial state selected after stimulation. Both tested
2x-volume geometries retain two late Cdc42 peaks and fail the predefined stable
unique-state criterion. Because this is a model-port experiment, that result is
a mechanistic prediction rather than direct evidence that genome-doubled cells
must behave this way in vivo.

# Limitations

This is a fixed-shape, two-dimensional numerical experiment interpreted using
implied three-dimensional volume scaling. It omits the cytoskeletal and shape
feedback present in the full Marée model, and it does not model genome doubling
itself, gene dosage, growth history, mechanics, or parameter adaptation. The
tested teardrop geometries and stimulus protocol span only a small part of the
possible shape and parameter space. Multiple late peaks are therefore a model
prediction under the stated controls, not evidence that enlarged or
genome-doubled cells generally become multipolar.

# Reproducibility boundary

The report snapshot schema is version `r snapshot$manifest$schema_version` and
was generated at `r snapshot$manifest$generated_utc`. Its VCML source SHA-256 is
`r snapshot$manifest$source_vcml_sha256`. Full solver checkpoints are excluded
from the repository because they are hundreds of megabytes; the committed
snapshot contains all values needed to reproduce the plots and tables above.

# References

Marée AFM, Grieneisen VA, Edelstein-Keshet L (2012). How Cells Integrate
Complex Stimuli: The Effect of Feedback from Phosphoinositides and Cell Shape
on Cell Polarization and Motility. *PLoS Computational Biology* 8(3): e1002402.
[doi:10.1371/journal.pcbi.1002402](https://doi.org/10.1371/journal.pcbi.1002402).

Eroumé K, Vasilevich A, Vermeulen S, de Boer J, Carlier A (2021). On the
influence of cell shape on dynamic reaction-diffusion polarization patterns.
*PLoS ONE* 16(3): e0248293.
[doi:10.1371/journal.pone.0248293](https://doi.org/10.1371/journal.pone.0248293).

De Belly H, Gallén AF, Strickland E, et al. (2026). Long-range mutual
activation establishes Rho and Rac polarity during cell migration. *Nature
Cell Biology* 28:1244--1257.
[doi:10.1038/s41556-026-01965-1](https://doi.org/10.1038/s41556-026-01965-1).
This mechanochemical alternative has been evaluated conceptually but is not
implemented here.
