---
title: "An Eroumé--Marée polarity model under cell-size scaling"
subtitle: "Exact VCML reaction-system port with a new 0.5x/1x/2x geometry experiment"
output:
  html_document:
    toc: true
    toc_float: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE,
                      fig.align = "center", dpi = 110)
library(ggplot2)
library(gridExtra)
source("../R/wgd_spatial_analysis.R")
snapshot <- read_wgd_snapshot("assets/wgd_spatial/v1")
validate_wgd_snapshot_values(snapshot)
summary_table <- snapshot$summary
metrics <- snapshot$dynamics
conservation <- snapshot$conservation
boundary_profiles <- snapshot$boundary
longitudinal_profiles <- snapshot$longitudinal
spatial_fields <- snapshot$spatial
geometry_colours <- wgd_geometry_colours()
plot_theme <- theme_minimal(base_size = 11) +
  theme(panel.grid.minor = element_blank(), legend.position = "bottom",
        plot.title = element_text(face = "bold"))
```

[&larr; Report catalogue](../index.html)

> **Status: published VCell model ported + WGD-oriented size extension.** The
> nine-PDE reaction system and parameters are from the Eroumé et al. public
> VCell model; changing cell size while holding concentrations and kinetics
> fixed is a new experiment in this repository.

The exact plot-level data used here are exported to `docs/data/` during the
site build. The completed numerical experiment is stored as a small,
checksummed reporting snapshot. Rebuilding this page validates and reads that
snapshot; it does not compile the numerical kernel or execute the PDE solver.

# Question and answer

Does cell enlargement alone cause a spatial Rac/Cdc42/Rho polarity circuit to
establish more concentration peaks, particularly along the cell boundary, when
concentrations, kinetic constants, diffusion coefficients, shape, and relative
stimulus position are otherwise held fixed?

**In these simulations, yes--but the boundary result is partly transient and
shape-dependent.** The 0.5x- and 1x-volume cells never exceed one longitudinal
Cdc42 peak after the stimulus, whereas both 2x cells reach three, end with two,
and remain multi-peaked for 166--168 s. Along the boundary-adjacent trace, both
2x cells also establish a third local Cdc42 maximum at 50 and 100 s, compared
with two flanking maxima in the smaller cells. At 200 s, that extra boundary
maximum remains in the aspect-ratio 1.00 cell but has merged away in the
aspect-ratio 1.29 cell. Enlargement therefore permits and prolongs additional
spatial domains; it does not guarantee a permanently larger number of membrane
boundary peaks in every shape.

# Model provenance: what is published and what is new

Marée et al. (2012) developed the parent polarity model: active and inactive
Cdc42, Rac, and Rho interact with PIP, PIP2, and PIP3 feedback and, in the full
model, cytoskeletal and cell-shape dynamics. That paper is the foundational
lineage; it is not separately reimplemented here.

Eroumé et al. (2021) implemented the Marée signalling system in Virtual Cell
to study polarization on fixed cell shapes. Because shape was prescribed,
they omitted the cytoskeletal module. Their extended fixed-shape model has
nine PDE variables: active and inactive Cdc42, Rac, and Rho, plus PIP, PIP2,
and PIP3. Eroumé et al. state that its parameter values are those of Marée et
al. and released the VCell models publicly, including
`Kerbai_PLoSone_2021_teardrop_polarization_extended`.

For this repository, that Eroumé VCML model was obtained and parsed, and its
reaction system was ported without redesigning the biochemical network. The
new contribution is the controlled size experiment below: geometrically
matched 0.5x, 1x, and 2x implied volumes at two aspect ratios. Eroumé et al.
studied shape; they did **not** report this size-scaling experiment.

# Experimental design

Six simulations cross two teardrop aspect ratios with three implied 3D volume
classes. Linear dimensions scale with the cube root of volume. The original
VCML cell mask is stretched onto an 80 by 80 mesh, while physical diffusion
coefficients and reaction parameters remain unchanged.

```{r design-table}
knitr::kable(
  summary_table[c("geometry", "size", "length_um", "width_um", "aspect_ratio")],
  col.names = c("Geometry", "Volume", "Length (um)", "Width (um)", "Aspect ratio"),
  digits = 2, row.names = FALSE
)
```

The system was integrated from 0 to 200 s with a 0.02 s timestep. A spatial
input is present through 10 s and then becomes uniform. Thus later structure
tests whether the network maintains or reorganizes the induced polarity.

# Polarization dynamics

```{r dynamics, fig.width=11, fig.height=7}
contrast_long <- rbind(
  data.frame(metrics[c("geometry", "time")], species = "Cdc42",
             contrast = metrics$cdc42_contrast),
  data.frame(metrics[c("geometry", "time")], species = "Rac",
             contrast = metrics$rac_contrast)
)
p_contrast <- ggplot(contrast_long,
  aes(time, contrast, colour = geometry)) +
  geom_vline(xintercept = 10, linetype = 3, colour = "grey45") +
  geom_line(linewidth = 0.8) + facet_wrap(~species, ncol = 1, scales = "free_y") +
  scale_colour_manual(values = geometry_colours) +
  labs(title = "Front--back polarization dynamics",
       subtitle = "Dotted line: spatial stimulus becomes uniform",
       x = "Time (s)", y = "(front - back) / (front + back)", colour = NULL) +
  plot_theme

p_separation <- ggplot(metrics,
  aes(time, rho_separation, colour = geometry)) +
  geom_vline(xintercept = 10, linetype = 3, colour = "grey45") +
  geom_line(linewidth = 0.8) + scale_colour_manual(values = geometry_colours) +
  labs(title = "Cdc42--Rho spatial separation", x = "Time (s)",
       y = "Centroid separation (cell lengths)", colour = NULL) + plot_theme

gridExtra::grid.arrange(p_contrast, p_separation, ncol = 2, widths = c(1.15, 1))
```

The larger cells establish polarity later: the 90% time increases from
20--23 s in the 0.5x and 1x cells to 25 s in both 2x cells. More importantly,
the two 2x-volume geometries retain multiple *longitudinal* Cdc42 peaks for
most of the post-stimulus interval, whereas the smaller cells never establish
more than one. These peak counts are computed from the concentration averaged
across each transverse slice; the boundary-specific evidence is shown in the
next section.

```{r result-table}
display_summary <- summary_table[c(
  "geometry", "time_to_90pct_cdc42_polarity_s",
  "max_cdc42_peaks_after_stimulus", "final_cdc42_peaks",
  "cdc42_multi_peak_duration_s", "unique_state_stable"
)]
knitr::kable(display_summary,
  col.names = c("Geometry", "90% polarity (s)",
                "Maximum longitudinal peaks", "Final longitudinal peaks",
                "Multi-peak duration (s)", "Stable unique state"),
  digits = 3, row.names = FALSE)
```

# Spatial organization

```{r final-fields, fig.width=11, fig.height=11}
final_fields <- subset(spatial_fields, time == 200)
final_fields$species <- factor(final_fields$species,
  levels = c("cdc42_act", "Rac_act", "Rho_act"),
  labels = c("Cdc42 active", "Rac active", "Rho active"))
ggplot(final_fields, aes(x, y, fill = value)) +
  geom_tile(aes(width = dx, height = dy)) + coord_fixed() +
  facet_grid(geometry ~ species) + scale_fill_viridis_c(option = "magma") +
  labs(title = "Spatial fields at 200 s", x = "x (um)", y = "y (um)",
       fill = "uM") + plot_theme + theme(legend.position = "right")
```

## Concentration along the membrane boundary

The next plot is the direct test of the membrane-boundary part of the
question. It follows active Cdc42 in the grid cells immediately adjacent to
the membrane. Relative perimeter 0 is the rear tip, the trace proceeds
clockwise, and the front tip is near 0.5. These are boundary-adjacent samples
of a spatial concentration field, not a separately modelled membrane-bound
species.

```{r membrane-boundary, fig.width=12, fig.height=6.5}
boundary_cdc42 <- subset(
  boundary_profiles,
  species == "cdc42_act" & time %in% c(50, 100, 200)
)
boundary_cdc42$size <- factor(boundary_cdc42$size,
  levels = c("0.5x", "1x", "2x"))
boundary_cdc42$time <- factor(boundary_cdc42$time,
  levels = c(50, 100, 200), labels = c("50 s", "100 s", "200 s"))
size_colours <- c(`0.5x` = "#2166AC", `1x` = "#7B3294", `2x` = "#D73027")

ggplot(boundary_cdc42,
  aes(relative_perimeter, concentration_uM, colour = size)) +
  geom_vline(xintercept = 0.5, linetype = 3, colour = "grey60") +
  geom_line(linewidth = 0.85) +
  facet_grid(pair ~ time) +
  scale_colour_manual(values = size_colours, drop = FALSE) +
  scale_x_continuous(breaks = seq(0, 1, by = 0.25)) +
  labs(title = "Active Cdc42 concentration along the cell perimeter",
       subtitle = "Dotted line marks the front tip; distance begins at the rear and proceeds clockwise",
       x = "Relative distance along perimeter", y = "Boundary-adjacent concentration (uM)",
       colour = "Relative volume") + plot_theme
```

The smaller cells show two perimeter maxima, one on either flank of the broad
front domain. At 50 and 100 s, the 2x cells show an additional maximum at the
front tip: three boundary-adjacent maxima rather than two. The extra maximum
persists through 200 s for aspect ratio 1.00. For aspect ratio 1.29 it merges
into the front domain between 100 and 150 s, leaving two boundary maxima at
200 s even though the whole-cell longitudinal profile remains multi-peaked.
Thus enlargement clearly increases the *capacity* and lifetime of additional
peaks, but final boundary peak number also depends on shape.

```{r longitudinal, fig.width=12, fig.height=7}
profile_plot_data <- subset(longitudinal_profiles, time %in% c(10, 50, 200))
profile_plot_data$species <- factor(profile_plot_data$species,
  levels = c("cdc42_act", "Rac_act", "Rho_act"),
  labels = c("Cdc42 active", "Rac active", "Rho active"))
ggplot(profile_plot_data,
  aes(relative_x, value, colour = geometry)) +
  geom_line(linewidth = 0.7) + facet_grid(species ~ time, scales = "free_y") +
  scale_colour_manual(values = geometry_colours) +
  labs(title = "Long-axis concentration profiles",
       x = "Relative position from rear to front", y = "Mean concentration (uM)",
       colour = NULL) + plot_theme
```

# Numerical checks and interpretation

The maximum relative drift in each conserved GTPase pool is below
`r format(max(conservation$max_relative_drift), scientific = TRUE, digits = 2)`.
This makes numerical mass loss an implausible explanation for the size-dependent
states.

Within this model, enlargement does more than slow the approach to the same
solution: it changes the number and persistence of the spatial domains selected
after stimulation. Both tested 2x-volume geometries reach three longitudinal
Cdc42 peaks, end with two, and fail the predefined stable unique-state
criterion; no smaller cell exceeds one longitudinal peak. On the boundary,
both larger cells transiently establish a third Cdc42 maximum, but only one
shape retains it to 200 s. The conclusion is therefore that added cell size
supports extra polarity domains, including extra boundary structure, while
shape influences whether a boundary peak persists. Because this is a
model-port experiment, that result is a mechanistic prediction rather than
direct evidence that genome-doubled cells must behave this way in vivo.

# Limitations

This is a fixed-shape, two-dimensional numerical experiment interpreted using
implied three-dimensional volume scaling. It omits the cytoskeletal and shape
feedback present in the full Marée model, and it does not model genome doubling
itself, gene dosage, growth history, mechanics, or parameter adaptation. The
tested teardrop geometries and stimulus protocol span only a small part of the
possible shape and parameter space. The peak-count result also depends on the
stated operational definitions: longitudinal peaks are counted after transverse
averaging, while the membrane plot samples boundary-adjacent grid cells and
shows local perimeter maxima. Multiple late peaks are therefore a model
prediction under the stated controls, not evidence that enlarged or
genome-doubled cells generally become multipolar.

# Reproducibility boundary

The report snapshot schema is version `r snapshot$manifest$schema_version` and
was generated at `r snapshot$manifest$generated_utc`. Its VCML source SHA-256 is
`r snapshot$manifest$source_vcml_sha256`. Full solver checkpoints are excluded
from the repository because they are hundreds of megabytes; the committed
snapshot contains all values needed to reproduce the plots and tables above.

# References

Marée AFM, Grieneisen VA, Edelstein-Keshet L (2012). How Cells Integrate
Complex Stimuli: The Effect of Feedback from Phosphoinositides and Cell Shape
on Cell Polarization and Motility. *PLoS Computational Biology* 8(3): e1002402.
[doi:10.1371/journal.pcbi.1002402](https://doi.org/10.1371/journal.pcbi.1002402).

Eroumé K, Vasilevich A, Vermeulen S, de Boer J, Carlier A (2021). On the
influence of cell shape on dynamic reaction-diffusion polarization patterns.
*PLoS ONE* 16(3): e0248293.
[doi:10.1371/journal.pone.0248293](https://doi.org/10.1371/journal.pone.0248293).

De Belly H, Gallén AF, Strickland E, et al. (2026). Long-range mutual
activation establishes Rho and Rac polarity during cell migration. *Nature
Cell Biology* 28:1244--1257.
[doi:10.1038/s41556-026-01965-1](https://doi.org/10.1038/s41556-026-01965-1).
This mechanochemical alternative has been evaluated conceptually but is not
implemented here.
