---
title: "Cell size and spatial Rac/Cdc42 polarity after genome doubling"
subtitle: "Matched-volume reaction--diffusion experiments in two cell shapes"
output:
  html_document:
    toc: true
    toc_float: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE,
                      fig.align = "center", dpi = 110)
library(ggplot2)
library(gridExtra)
source("../R/wgd_spatial_analysis.R")
snapshot <- read_wgd_snapshot("assets/wgd_spatial/v1")
validate_wgd_snapshot_values(snapshot)
summary_table <- snapshot$summary
metrics <- snapshot$dynamics
conservation <- snapshot$conservation
boundary_profiles <- snapshot$boundary
longitudinal_profiles <- snapshot$longitudinal
spatial_fields <- snapshot$spatial
geometry_colours <- wgd_geometry_colours()
plot_theme <- theme_minimal(base_size = 11) +
  theme(panel.grid.minor = element_blank(), legend.position = "bottom",
        plot.title = element_text(face = "bold"))
```

The exact plot-level data used here are exported to `docs/data/` during the
site build. The completed numerical experiment is stored as a small,
checksummed reporting snapshot. Rebuilding this page validates and reads that
snapshot; it does not compile the numerical kernel or execute the PDE solver.

# Question

Does cell enlargement alone change the dynamics or stability of a spatial
Rac/Cdc42/Rho polarity circuit when concentrations, kinetic constants,
diffusion coefficients, shape, and relative stimulus position are otherwise
held fixed?

# Experimental design

Six simulations cross two teardrop aspect ratios with three implied 3D volume
classes. Linear dimensions scale with the cube root of volume. The original
VCML cell mask is stretched onto an 80 by 80 mesh, while physical diffusion
coefficients and reaction parameters remain unchanged.

```{r design-table}
knitr::kable(
  summary_table[c("geometry", "size", "length_um", "width_um", "aspect_ratio")],
  col.names = c("Geometry", "Volume", "Length (um)", "Width (um)", "Aspect ratio"),
  digits = 2, row.names = FALSE
)
```

The system was integrated from 0 to 200 s with a 0.02 s timestep. A spatial
input is present through 10 s and then becomes uniform. Thus later structure
tests whether the network maintains or reorganizes the induced polarity.

# Polarization dynamics

```{r dynamics, fig.width=11, fig.height=7}
contrast_long <- rbind(
  data.frame(metrics[c("geometry", "time")], species = "Cdc42",
             contrast = metrics$cdc42_contrast),
  data.frame(metrics[c("geometry", "time")], species = "Rac",
             contrast = metrics$rac_contrast)
)
p_contrast <- ggplot(contrast_long,
  aes(time, contrast, colour = geometry)) +
  geom_vline(xintercept = 10, linetype = 3, colour = "grey45") +
  geom_line(linewidth = 0.8) + facet_wrap(~species, ncol = 1, scales = "free_y") +
  scale_colour_manual(values = geometry_colours) +
  labs(title = "Front--back polarization dynamics",
       subtitle = "Dotted line: spatial stimulus becomes uniform",
       x = "Time (s)", y = "(front - back) / (front + back)", colour = NULL) +
  plot_theme

p_separation <- ggplot(metrics,
  aes(time, rho_separation, colour = geometry)) +
  geom_vline(xintercept = 10, linetype = 3, colour = "grey45") +
  geom_line(linewidth = 0.8) + scale_colour_manual(values = geometry_colours) +
  labs(title = "Cdc42--Rho spatial separation", x = "Time (s)",
       y = "Centroid separation (cell lengths)", colour = NULL) + plot_theme

gridExtra::grid.arrange(p_contrast, p_separation, ncol = 2, widths = c(1.15, 1))
```

The larger cells establish polarity later. More importantly, the two 2x-volume
geometries retain multiple Cdc42 peaks for most of the post-stimulus interval,
whereas the smaller cells settle into one peak.

```{r result-table}
display_summary <- summary_table[c(
  "geometry", "time_to_90pct_cdc42_polarity_s",
  "final_cdc42_front_back", "final_rho_separation_cell_lengths",
  "cdc42_multi_peak_duration_s", "unique_state_stable"
)]
knitr::kable(display_summary,
  col.names = c("Geometry", "90% polarity (s)", "Final Cdc42 front:back",
                "Cdc42--Rho separation", "Multi-peak duration (s)",
                "Stable unique state"),
  digits = 3, row.names = FALSE)
```

# Spatial organization

```{r final-fields, fig.width=11, fig.height=11}
final_fields <- subset(spatial_fields, time == 200)
final_fields$species <- factor(final_fields$species,
  levels = c("cdc42_act", "Rac_act", "Rho_act"),
  labels = c("Cdc42 active", "Rac active", "Rho active"))
ggplot(final_fields, aes(x, y, fill = value)) +
  geom_tile(aes(width = dx, height = dy)) + coord_fixed() +
  facet_grid(geometry ~ species) + scale_fill_viridis_c(option = "magma") +
  labs(title = "Spatial fields at 200 s", x = "x (um)", y = "y (um)",
       fill = "uM") + plot_theme + theme(legend.position = "right")
```

```{r longitudinal, fig.width=12, fig.height=7}
profile_plot_data <- subset(longitudinal_profiles, time %in% c(10, 50, 200))
profile_plot_data$species <- factor(profile_plot_data$species,
  levels = c("cdc42_act", "Rac_act", "Rho_act"),
  labels = c("Cdc42 active", "Rac active", "Rho active"))
ggplot(profile_plot_data,
  aes(relative_x, value, colour = geometry)) +
  geom_line(linewidth = 0.7) + facet_grid(species ~ time, scales = "free_y") +
  scale_colour_manual(values = geometry_colours) +
  labs(title = "Long-axis concentration profiles",
       x = "Relative position from rear to front", y = "Mean concentration (uM)",
       colour = NULL) + plot_theme
```

# Numerical checks and interpretation

The maximum relative drift in each conserved GTPase pool is below
`r format(max(conservation$max_relative_drift), scientific = TRUE, digits = 2)`.
This makes numerical mass loss an implausible explanation for the size-dependent
states.

Within this model, enlargement does more than slow the approach to the same
solution: it changes the spatial state selected after stimulation. Both tested
2x-volume geometries retain two late Cdc42 peaks and fail the predefined stable
unique-state criterion. Because this is a model-port experiment, that result is
a mechanistic prediction rather than direct evidence that genome-doubled cells
must behave this way in vivo.

# Reproducibility boundary

The report snapshot schema is version `r snapshot$manifest$schema_version` and
was generated at `r snapshot$manifest$generated_utc`. Its VCML source SHA-256 is
`r snapshot$manifest$source_vcml_sha256`. Full solver checkpoints are excluded
from the repository because they are hundreds of megabytes; the committed
snapshot contains all values needed to reproduce the plots and tables above.
