---
title: "Cell enlargement can desensitize a Rac/Cdc42 polarity circuit"
subtitle: "A Holmes et al. (2012) membrane-cytosol scaling experiment"
output:
  html_document:
    toc: true
    toc_float: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = FALSE,
  message = FALSE,
  warning = FALSE,
  fig.align = "center"
)

library(ggplot2)
library(deSolve)

volume_colours <- c(`1x volume` = "#0072B2", `2x volume` = "#D55E00")

plot_theme <- theme_minimal(base_size = 11) +
  theme(
    legend.position = "bottom",
    panel.grid.minor = element_blank()
  )
```

[&larr; Report catalogue](../index.html)

> **Status: published model adapted + new enlargement experiment.** The
> Rac/Cdc42/Rho circuit and membrane-accessibility mechanism come from Holmes
> et al.; geometrically similar 1x/2x enlargement is tested here.

The exact data used in every plot are exported to `docs/data/` when this report
is rendered inside the project repository.

# Question

If a mammalian cell doubles in volume while maintaining the concentrations of
its polarity proteins, can geometry alone make it harder to establish a
Rac/Cdc42 front?

This report adapts the experimentally motivated model of:

> Holmes WR, Lin B, Levchenko A, Edelstein-Keshet L. **Modelling Cell
> Polarization Driven by Synthetic Spatially Graded Rac Activation.**
> *PLoS Computational Biology* 8:e1002366 (2012).
> DOI: 10.1371/journal.pcbi.1002366.

The original study was built around HeLa cells in microfluidic channels. A
linear rapamycin gradient recruited the Rac GEF Tiam1 to the plasma membrane,
allowing Rac activation to be stimulated directly. The experiments showed
threshold-like polarization, and the model was constructed to reproduce that
behavior.

Holmes et al. also introduced a geometric effect that is particularly useful
for cell-size scaling: inactive Rac, Cdc42, and Rho exchange between the
cytosol and plasma membrane, but GEFs can activate only the membrane-bound
fraction. Their model therefore makes signaling sensitivity depend on the
ratio of membrane area to cytoplasmic volume.

The original paper considered **flattening at approximately fixed volume**:
surface-area:volume increased and polarization became easier. Here we ask the
reverse scaling question relevant to cell enlargement:

> For geometrically similar cells, does the fall in surface-area:volume at
> larger volume reduce membrane accessibility enough to change the response to
> the same Rac-GEF gradient?

This is an extension of the Holmes model, not an experiment performed in the
paper.

# 1. The Holmes GTPase model

Holmes et al. represent each GTPase with a slowly diffusing active
membrane-bound form \(G\) and a composite inactive form \(G_i\), which combines
inactive membrane-bound and cytosolic material. After assuming that
membrane/cytosol exchange is fast, their reduced equations are

\[
\frac{\partial G}{\partial t}
=
I_G\,\bar c\,\frac{G_i}{G_t}
-
\delta_G G
+
D_m \frac{\partial^2 G}{\partial x^2},
\]

\[
\frac{\partial G_i}{\partial t}
=
-
I_G\,\bar c\,\frac{G_i}{G_t}
+
\delta_G G
+
D_{mc}\frac{\partial^2 G_i}{\partial x^2}.
\]

Here \(G_t\) is the total level of that GTPase and

\[
\bar c=\frac{c}{c_0}
\]

is the membrane-accessibility factor relative to a baseline geometry.

The underlying fraction of inactive GTPase that is membrane bound is

\[
c
=
\frac{k_{\rm on}}
{k_{\rm on} + (V/L)k_{\rm off}},
\]

where \(V/L\) represents the cross-sectional volume associated with a unit
length of the one-dimensional cell representation.

This term is the important part for the current scaling exercise. The GEF
activation term is multiplied by \(\bar c\), because GEFs can access only the
membrane-bound fraction of the inactive pool.

Holmes et al. report that geometry-dependent changes in the effective inactive
diffusion coefficient have relatively little effect on the polarity regimes,
whereas changing the membrane-accessible fraction has a substantial effect.

## Model 4: Cdc42, Rac, and Rho

We use the GTPase layer of their final Model 4, without phosphoinositide
feedback. This matches the geometry analysis in their Figure 8, for which
\(f_1=0\).

The activation rates are

\[
I_C
=
\frac{\hat I_C}
{1+(\rho/a_1)^n},
\]

\[
I_R
=
\hat I_{R1}
+
aC
+
S(x),
\]

\[
I_\rho
=
\frac{\hat I_\rho}
{1+(R/a_2)^n},
\]

where \(C\), \(R\), and \(\rho\) denote active Cdc42, Rac, and Rho
respectively.

Thus Rho inhibits Cdc42, Cdc42 promotes Rac, and Rac inhibits Rho. This
mutually coupled circuit has bistable regimes in which a sufficiently
asymmetric Rac-GEF stimulus can push one end of the cell across a threshold and
generate a polarized state.

The signal is represented as a spatial Rac-GEF gradient,

\[
S(x)
=
s_{\rm mean}
+
s_{\rm grad}
\left(2x/L-1\right),
\]

so `s_mean` controls the background stimulus while `s_grad` controls the
front/back asymmetry. The mean signal is held fixed when cell size changes.

# 2. Published parameters

The core parameters below come from Table 1 of Holmes et al. The Cdc42-to-Rac
feedback strength is set to \(a=0.55~{\rm s^{-1}}\), the value used in their
Figure 8 geometry analysis rather than the Table 1 default of 0.65.

```{r parameter-table}
pars <- list(
  L0_um = 20,
  Ct = 2.4,
  Rt = 7.5,
  rhot = 3.1,
  IC_hat = 2.95,
  IR1_hat = 0.2,
  Irho_hat = 6.6,
  a1 = 1.25,
  a2 = 1,
  n = 3,
  a = 0.55,
  delta_C = 1,
  delta_R = 1,
  delta_rho = 1,
  Dm = 0.1,
  Dmc0 = 50
)

parameter_table <- data.frame(
  Parameter = c(
    "Baseline cell length L0",
    "Total Cdc42, Rac, Rho",
    "I_C, I_R1, I_rho",
    "a1, a2",
    "Hill coefficient n",
    "Cdc42 -> Rac feedback a",
    "GAP inactivation rates",
    "Active membrane diffusion Dm",
    "Baseline composite inactive diffusion Dmc"
  ),
  Value = c(
    "20 um",
    "2.4, 7.5, 3.1 uM",
    "2.95, 0.2, 6.6 uM/s",
    "1.25, 1 uM",
    "3",
    "0.55 /s",
    "1 /s",
    "0.1 um^2/s",
    "50 um^2/s"
  ),
  Source = c(
    rep("Holmes Table 1", 5),
    "Holmes Figure 8",
    rep("Holmes Table 1", 3)
  )
)

knitr::kable(parameter_table, align = c("l", "l", "l"))
```

The authors explicitly state that \(k_{\rm on}\), \(k_{\rm off}\), cytosolic
diffusion, and the baseline membrane fraction were not sufficiently determined
to assign unique numerical values. They therefore varied the composite
quantity \(\bar c\) directly.

That uncertainty is retained here. We do **not** assume that there is one
known membrane-bound fraction of Rac/Cdc42/Rho. Instead we repeat the scaling
exercise for baseline membrane fractions ranging from 0.1 to 0.8.

# 3. Mapping 1x -> 2x volume onto membrane accessibility

For geometrically similar cells,

\[
L(v)=L_0 v^{1/3},
\]

and

\[
\frac{V}{L}\propto v^{2/3}.
\]

Let \(c_0\) be the membrane-bound fraction in the 1x cell. Eliminating the
unknown ratio \(k_{\rm off}/k_{\rm on}\) gives

\[
c(v)
=
\frac{1}
{1+
\left(\frac{1-c_0}{c_0}\right)v^{2/3}
}.
\]

Therefore

\[
\bar c(v)=\frac{c(v)}{c_0}.
\]

This is the WGD-oriented extrapolation. Holmes et al. derived the same
membrane/cytosol partitioning relationship, but used it for a cell that
elongated and flattened at approximately fixed volume. Here the geometry is
instead scaled isotropically with volume.

```{r geometry-functions}
membrane_fraction <- function(volume_multiplier, baseline_fraction) {
  1 / (
    1 +
      ((1 - baseline_fraction) / baseline_fraction) *
      volume_multiplier^(2 / 3)
  )
}

accessibility_ratio <- function(volume_multiplier, baseline_fraction) {
  membrane_fraction(volume_multiplier, baseline_fraction) / baseline_fraction
}

cytosolic_diffusion_from_baseline <- function(baseline_fraction) {
  (
    pars$Dmc0 -
      baseline_fraction * pars$Dm
  ) / (1 - baseline_fraction)
}

inactive_diffusion <- function(volume_multiplier, baseline_fraction) {
  c_now <- membrane_fraction(volume_multiplier, baseline_fraction)
  Dc <- cytosolic_diffusion_from_baseline(baseline_fraction)

  c_now * pars$Dm + (1 - c_now) * Dc
}

baseline_fractions <- c(0.1, 0.2, 0.5, 0.8)
volume_grid <- exp(seq(log(0.5), log(4), length.out = 100))

geometry_scaling <- do.call(
  rbind,
  lapply(baseline_fractions, function(c0) {
    data.frame(
      volume_multiplier = volume_grid,
      baseline_fraction = factor(c0),
      membrane_fraction = membrane_fraction(volume_grid, c0),
      accessibility_ratio = accessibility_ratio(volume_grid, c0)
    )
  })
)
```

```{r geometry-scaling-plot, fig.width=8, fig.height=5}
ggplot(
  geometry_scaling,
  aes(
    volume_multiplier,
    accessibility_ratio,
    colour = baseline_fraction
  )
) +
  geom_line(linewidth = 0.9) +
  geom_vline(xintercept = c(1, 2), linetype = c("solid", "dashed")) +
  scale_x_log10(
    breaks = c(0.5, 1, 2, 4),
    labels = c("0.5x", "1x", "2x", "4x")
  ) +
  labs(
    title = "Larger cells expose less of the inactive GTPase pool to membrane GEFs",
    subtitle = "Holmes membrane/cytosol partitioning under isotropic cell scaling",
    x = "Cell-volume multiplier",
    y = "Membrane accessibility relative to 1x",
    colour = "1x membrane\nfraction"
  ) +
  plot_theme
```

For a 2x-volume cell the linear dimension increases by only
\(2^{1/3}\approx1.26\), but the cross-sectional term \(V/L\) increases by
\(2^{2/3}\approx1.59\). The resulting fall in membrane accessibility depends
on the unknown baseline partitioning.

```{r geometry-summary}
geometry_summary <- data.frame(
  baseline_membrane_fraction = baseline_fractions,
  one_x_accessibility = 1,
  two_x_membrane_fraction =
    vapply(
      baseline_fractions,
      function(c0) membrane_fraction(2, c0),
      numeric(1)
    ),
  two_x_relative_accessibility =
    vapply(
      baseline_fractions,
      function(c0) accessibility_ratio(2, c0),
      numeric(1)
    )
)

knitr::kable(geometry_summary, digits = 3)
```

Across this deliberately broad range of baseline membrane fractions, doubling
cell volume reduces the GEF-accessible inactive pool per total GTPase pool by
approximately
`r sprintf("%.0f", 100 * (1 - max(geometry_summary$two_x_relative_accessibility)))`--
`r sprintf("%.0f", 100 * (1 - min(geometry_summary$two_x_relative_accessibility)))`%.

Unlike the Buttenschön dilution example, this effect remains even if the
**total concentrations of Cdc42, Rac, and Rho are maintained perfectly** as
the cell enlarges.

# 4. Numerical implementation

We discretize the one-dimensional cell with a cell-centred finite-difference
grid and no-flux boundaries. The method-of-lines system is integrated with
`deSolve::lsodes`.

The original paper used an implicit-diffusion, explicit-reaction scheme. Thus
the equations and parameterization are taken from Holmes et al., but the
numerical integrator is not identical.

```{r solver-functions}
laplacian_no_flux <- function(u, dx) {
  n <- length(u)
  ans <- numeric(n)

  ans[1] <- (u[2] - u[1]) / dx^2
  ans[n] <- (u[n - 1] - u[n]) / dx^2

  if (n > 2) {
    ans[2:(n - 1)] <-
      (
        u[1:(n - 2)] -
          2 * u[2:(n - 1)] +
          u[3:n]
      ) / dx^2
  }

  ans
}

reaction_terms <- function(C, Rac, Rho, Ci, Raci, Rhoi,
                           accessibility, signal) {
  I_C <-
    pars$IC_hat /
    (1 + (Rho / pars$a1)^pars$n)

  I_Rac <-
    pars$IR1_hat +
    pars$a * C +
    signal

  I_Rho <-
    pars$Irho_hat /
    (1 + (Rac / pars$a2)^pars$n)

  f_C <-
    I_C * accessibility * Ci / pars$Ct -
    pars$delta_C * C

  f_Rac <-
    I_Rac * accessibility * Raci / pars$Rt -
    pars$delta_R * Rac

  f_Rho <-
    I_Rho * accessibility * Rhoi / pars$rhot -
    pars$delta_rho * Rho

  list(C = f_C, Rac = f_Rac, Rho = f_Rho)
}

hss_cache <- new.env(parent = emptyenv())

relax_homogeneous_state <- function(accessibility, duration_s = 300) {
  key <- sprintf("%.8f", accessibility)

  if (exists(key, envir = hss_cache, inherits = FALSE)) {
    return(get(key, envir = hss_cache, inherits = FALSE))
  }

  initial <- c(
    C = 0.1,
    Rac = 0.1,
    Rho = 0.1,
    Ci = pars$Ct - 0.1,
    Raci = pars$Rt - 0.1,
    Rhoi = pars$rhot - 0.1
  )

  rhs <- function(time_s, state, parameters) {
    rxn <- reaction_terms(
      state[["C"]],
      state[["Rac"]],
      state[["Rho"]],
      state[["Ci"]],
      state[["Raci"]],
      state[["Rhoi"]],
      accessibility = accessibility,
      signal = 0
    )

    list(c(
      C = rxn$C,
      Rac = rxn$Rac,
      Rho = rxn$Rho,
      Ci = -rxn$C,
      Raci = -rxn$Rac,
      Rhoi = -rxn$Rho
    ))
  }

  out <- ode(
    y = initial,
    times = c(0, duration_s),
    func = rhs,
    parms = NULL,
    method = "lsodes",
    rtol = 1e-9,
    atol = 1e-11
  )

  result <- as.numeric(out[nrow(out), -1])
  names(result) <- names(initial)

  assign(key, result, envir = hss_cache)
  result
}

simulate_holmes <- function(
    length_um,
    accessibility,
    D_inactive,
    signal_mean = 0.3,
    signal_half_range = 0.1,
    duration_s = 300,
    n_cells = 41,
    save_times = NULL) {

  dx <- length_um / n_cells
  x <- (seq_len(n_cells) - 0.5) * dx

  signal <-
    signal_mean +
    signal_half_range * (2 * x / length_um - 1)

  if (any(signal < 0)) {
    stop("Signal became negative; increase signal_mean or reduce signal_half_range.")
  }

  hss <- relax_homogeneous_state(accessibility)

  state0 <- c(
    setNames(rep(hss[["C"]], n_cells), paste0("C_", seq_len(n_cells))),
    setNames(rep(hss[["Rac"]], n_cells), paste0("Rac_", seq_len(n_cells))),
    setNames(rep(hss[["Rho"]], n_cells), paste0("Rho_", seq_len(n_cells))),
    setNames(rep(hss[["Ci"]], n_cells), paste0("Ci_", seq_len(n_cells))),
    setNames(rep(hss[["Raci"]], n_cells), paste0("Raci_", seq_len(n_cells))),
    setNames(rep(hss[["Rhoi"]], n_cells), paste0("Rhoi_", seq_len(n_cells)))
  )

  rhs <- function(time_s, state, parameters) {
    idx <- seq_len(n_cells)

    C <- state[idx]
    Rac <- state[n_cells + idx]
    Rho <- state[2 * n_cells + idx]
    Ci <- state[3 * n_cells + idx]
    Raci <- state[4 * n_cells + idx]
    Rhoi <- state[5 * n_cells + idx]

    rxn <- reaction_terms(
      C, Rac, Rho, Ci, Raci, Rhoi,
      accessibility = accessibility,
      signal = signal
    )

    dC <-
      pars$Dm * laplacian_no_flux(C, dx) +
      rxn$C

    dRac <-
      pars$Dm * laplacian_no_flux(Rac, dx) +
      rxn$Rac

    dRho <-
      pars$Dm * laplacian_no_flux(Rho, dx) +
      rxn$Rho

    dCi <-
      D_inactive * laplacian_no_flux(Ci, dx) -
      rxn$C

    dRaci <-
      D_inactive * laplacian_no_flux(Raci, dx) -
      rxn$Rac

    dRhoi <-
      D_inactive * laplacian_no_flux(Rhoi, dx) -
      rxn$Rho

    list(c(dC, dRac, dRho, dCi, dRaci, dRhoi))
  }

  if (is.null(save_times)) {
    save_times <- c(0, duration_s)
  }

  save_times <- sort(unique(c(0, save_times, duration_s)))

  out <- as.data.frame(
    ode(
      y = state0,
      times = save_times,
      func = rhs,
      parms = NULL,
      method = "lsodes",
      rtol = 2e-6,
      atol = 1e-8,
      maxsteps = 100000
    )
  )

  final <- out[nrow(out), -1]
  idx <- seq_len(n_cells)

  C <- as.numeric(final[idx])
  Rac <- as.numeric(final[n_cells + idx])
  Rho <- as.numeric(final[2 * n_cells + idx])
  Ci <- as.numeric(final[3 * n_cells + idx])
  Raci <- as.numeric(final[4 * n_cells + idx])
  Rhoi <- as.numeric(final[5 * n_cells + idx])

  contrast <- function(z) {
    (max(z) - min(z)) / (max(z) + min(z) + 1e-12)
  }

  list(
    x_um = x,
    signal = signal,
    output = out,
    profiles = data.frame(
      x_um = rep(x, 3),
      species = rep(c("Cdc42", "Rac", "Rho"), each = n_cells),
      active = c(C, Rac, Rho)
    ),
    summary = data.frame(
      length_um = length_um,
      accessibility = accessibility,
      D_inactive = D_inactive,
      signal_mean = signal_mean,
      signal_half_range = signal_half_range,
      Cdc42_contrast = contrast(C),
      Rac_contrast = contrast(Rac),
      Rho_contrast = contrast(Rho),
      Rac_min = min(Rac),
      Rac_max = max(Rac),
      Cdc42_min = min(C),
      Cdc42_max = max(C),
      Rho_min = min(Rho),
      Rho_max = max(Rho),
      mean_Cdc42_total = mean(C + Ci),
      mean_Rac_total = mean(Rac + Raci),
      mean_Rho_total = mean(Rho + Rhoi)
    )
  )
}
```

# 5. Decomposing the 2x-volume effect

Before scanning stimulus thresholds, it is useful to separate two consequences
of cell enlargement:

1. **longer physical domain**, which changes diffusion distance;
2. **lower membrane accessibility**, which changes how much inactive GTPase the
   GEF reactions can use.

For this decomposition we use a baseline membrane-bound inactive fraction of
0.2. This is an illustrative value, not a measured Holmes parameter.

The signal mean is 0.3 model concentration/time units and the half-range of the
gradient is 0.1. Thus the same signal runs from 0.2 to 0.4 across every model
cell.

```{r decomposition-runs}
default_c0 <- 0.2
L1 <- pars$L0_um
L2 <- pars$L0_um * 2^(1 / 3)

cbar_2 <- accessibility_ratio(2, default_c0)
Dmc_2 <- inactive_diffusion(2, default_c0)

decomposition_conditions <- list(
  `1x baseline` = list(
    length_um = L1,
    accessibility = 1,
    D_inactive = pars$Dmc0
  ),
  `2x: length only` = list(
    length_um = L2,
    accessibility = 1,
    D_inactive = pars$Dmc0
  ),
  `2x: accessibility only` = list(
    length_um = L1,
    accessibility = cbar_2,
    D_inactive = Dmc_2
  ),
  `2x: full scaling` = list(
    length_um = L2,
    accessibility = cbar_2,
    D_inactive = Dmc_2
  )
)

decomposition_results <- lapply(
  decomposition_conditions,
  function(z) {
    simulate_holmes(
      length_um = z$length_um,
      accessibility = z$accessibility,
      D_inactive = z$D_inactive,
      signal_mean = 0.3,
      signal_half_range = 0.1,
      duration_s = 300,
      n_cells = 61
    )
  }
)

decomposition_summary <- do.call(
  rbind,
  lapply(names(decomposition_results), function(label) {
    x <- decomposition_results[[label]]$summary
    x$condition <- label
    x
  })
)

decomposition_summary$condition <- factor(
  decomposition_summary$condition,
  levels = names(decomposition_conditions)
)
```

```{r decomposition-plot, fig.width=8, fig.height=4.5}
ggplot(
  decomposition_summary,
  aes(condition, Rac_contrast)
) +
  geom_col(width = 0.7) +
  geom_hline(yintercept = 0.4, linetype = "dotted") +
  labs(
    title = "The size effect comes from membrane accessibility, not diffusion distance",
    subtitle = "Same Rac-GEF gradient in every condition; dotted line is an operational strong-polarity threshold",
    x = NULL,
    y = "Final Rac front/back contrast"
  ) +
  plot_theme +
  theme(axis.text.x = element_text(angle = 20, hjust = 1))
```

In this example, increasing linear cell size by \(2^{1/3}\) **without**
changing membrane accessibility has little effect on the Rac response. In
contrast, applying the Holmes membrane/cytosol partitioning change largely
removes the amplified Rac front.

The 1x Rac contrast is
`r sprintf("%.2f", decomposition_summary$Rac_contrast[decomposition_summary$condition == "1x baseline"])`,
whereas the full 2x-volume analogue is
`r sprintf("%.2f", decomposition_summary$Rac_contrast[decomposition_summary$condition == "2x: full scaling"])`.

This reproduces the qualitative conclusion of Holmes et al. from the opposite
direction: changing geometry matters primarily because it changes the fraction
of inactive GTPase that membrane GEFs can access.

# 6. What do the actual polarity profiles look like?

```{r example-profiles}
profile_cases <- c("1x baseline", "2x: full scaling")

example_profiles <- do.call(
  rbind,
  lapply(profile_cases, function(label) {
    x <- decomposition_results[[label]]$profiles
    x$volume <- ifelse(label == "1x baseline", "1x volume", "2x volume")
    x$relative_position <-
      x$x_um /
      decomposition_results[[label]]$summary$length_um
    x
  })
)

example_profiles$volume <- factor(
  example_profiles$volume,
  levels = c("1x volume", "2x volume")
)
```

```{r example-profiles-plot, fig.width=9, fig.height=6}
ggplot(
  example_profiles,
  aes(
    relative_position,
    active,
    colour = volume,
    linetype = volume
  )
) +
  geom_line(linewidth = 0.9) +
  facet_wrap(~ species, scales = "free_y", ncol = 1) +
  scale_colour_manual(values = volume_colours) +
  scale_linetype_manual(
    values = c(`1x volume` = "solid", `2x volume` = "dashed")
  ) +
  labs(
    title = "Same Rac-GEF gradient, different polarity response",
    subtitle = "Cdc42 and Rac mark the front; Rho becomes complementary",
    x = "Relative position across cell",
    y = "Active GTPase",
    colour = "Cell size",
    linetype = "Cell size"
  ) +
  plot_theme
```

The 1x cell exhibits nonlinear amplification of the imposed Rac-GEF gradient:
Rac and Cdc42 accumulate toward the stimulated end while Rho becomes
complementary.

The 2x-volume analogue receives the **same mean signal and the same fractional
gradient across the cell**, and total Cdc42/Rac/Rho concentrations are
unchanged. Nevertheless, the lower membrane-accessible fraction leaves the
system much closer to a weak graded response.

This is the qualitative effect that was missing from the simpler size-only
wave-pinning example.

# 7. Does enlargement shift the polarization threshold?

The most useful readout is not one arbitrarily selected stimulus. We therefore
scan the strength of the Rac-GEF gradient while keeping its mean fixed at 0.3.

For each run we quantify final Rac asymmetry as

\[
P_R
=
\frac{R_{\max}-R_{\min}}
{R_{\max}+R_{\min}}.
\]

This is a continuous response measure. For a compact summary we also define
\(P_R\ge0.4\) as an **operational** strong-polarity threshold. The value 0.4 is
not from Holmes et al.; it is only used to compare the response curves.

```{r gradient-sweep}
gradient_values <- seq(0, 0.30, by = 0.02)

sweep_rows <- list()
row_index <- 1

for (c0 in baseline_fractions) {
  for (volume_name in names(c(`1x volume` = 1, `2x volume` = 2))) {
    v <- c(`1x volume` = 1, `2x volume` = 2)[[volume_name]]

    L <- pars$L0_um * v^(1 / 3)
    cb <- accessibility_ratio(v, c0)
    D_inactive <- inactive_diffusion(v, c0)

    for (g in gradient_values) {
      run <- simulate_holmes(
        length_um = L,
        accessibility = cb,
        D_inactive = D_inactive,
        signal_mean = 0.3,
        signal_half_range = g,
        duration_s = 300,
        n_cells = 31
      )

      sweep_rows[[row_index]] <- data.frame(
        baseline_membrane_fraction = c0,
        volume = volume_name,
        volume_multiplier = v,
        signal_half_range = g,
        Rac_contrast = run$summary$Rac_contrast
      )

      row_index <- row_index + 1
    }
  }
}

gradient_sweep <- do.call(rbind, sweep_rows)

gradient_sweep$volume <- factor(
  gradient_sweep$volume,
  levels = c("1x volume", "2x volume")
)
```

```{r gradient-sweep-plot, fig.width=9, fig.height=7}
ggplot(
  gradient_sweep,
  aes(
    signal_half_range,
    Rac_contrast,
    colour = volume,
    linetype = volume
  )
) +
  geom_hline(yintercept = 0.4, linetype = "dotted") +
  geom_line(linewidth = 0.85) +
  geom_point(size = 1.5) +
  facet_wrap(
    ~ baseline_membrane_fraction,
    labeller = labeller(
      baseline_membrane_fraction =
        function(x) paste0("1x membrane fraction = ", x)
    )
  ) +
  scale_colour_manual(values = volume_colours) +
  scale_linetype_manual(
    values = c(`1x volume` = "solid", `2x volume` = "dashed")
  ) +
  labs(
    title = "Cell enlargement shifts the Rac-gradient response curve",
    subtitle = "Mean Rac-GEF stimulation is fixed; only gradient strength is varied",
    x = "Rac-GEF gradient half-range",
    y = "Final Rac contrast",
    colour = "Cell size",
    linetype = "Cell size"
  ) +
  plot_theme
```

```{r threshold-summary}
strong_threshold <- 0.4

threshold_summary <- do.call(
  rbind,
  lapply(
    split(
      gradient_sweep,
      interaction(
        gradient_sweep$baseline_membrane_fraction,
        gradient_sweep$volume,
        drop = TRUE
      )
    ),
    function(x) {
      x <- x[order(x$signal_half_range), ]

      crossing <- x$signal_half_range[
        x$Rac_contrast >= strong_threshold
      ]

      data.frame(
        baseline_membrane_fraction =
          x$baseline_membrane_fraction[1],
        volume = as.character(x$volume[1]),
        threshold_gradient =
          if (length(crossing) == 0) NA_real_ else min(crossing)
      )
    }
  )
)

threshold_wide <- merge(
  subset(
    threshold_summary,
    volume == "1x volume",
    select = c(
      baseline_membrane_fraction,
      threshold_gradient
    )
  ),
  subset(
    threshold_summary,
    volume == "2x volume",
    select = c(
      baseline_membrane_fraction,
      threshold_gradient
    )
  ),
  by = "baseline_membrane_fraction",
  suffixes = c("_1x", "_2x")
)

threshold_wide$fold_increase <-
  threshold_wide$threshold_gradient_2x /
  threshold_wide$threshold_gradient_1x

knitr::kable(threshold_wide, digits = 2)
```

Within this model regime, the size effect is not confined to one specially
chosen membrane fraction. Across the tested baseline partitioning values, the
2x-volume analogue requires a substantially larger imposed Rac-GEF asymmetry
to reach the same strong-polarity criterion.

For baseline membrane fractions where both thresholds are reached within the
scan, the required gradient increases by approximately
`r sprintf("%.1f", min(threshold_wide$fold_increase, na.rm = TRUE))`--
`r sprintf("%.1f", max(threshold_wide$fold_increase, na.rm = TRUE))` fold.

The exact numbers should **not** be interpreted as quantitative predictions for
WGD cells. The baseline membrane-bound fraction is unknown, and the Holmes
stimulus units are not calibrated here to an experimental rapamycin
concentration. The robust result is the direction and potential magnitude of
the threshold shift.

# 8. Why this is more interesting than the previous size examples

This exercise has three useful properties.

First, the effect survives **concentration homeostasis**. Total Cdc42, Rac, and
Rho concentrations are identical in the 1x and 2x simulations. The larger cell
is not being penalized by arbitrarily holding protein amount fixed.

Second, the dominant mechanism is a genuine **surface/volume consequence**.
For a larger geometrically similar cell, more inactive GTPase resides in the
cytoplasmic reservoir relative to the membrane, reducing the fraction that
membrane GEFs can activate.

Third, the output is migration-relevant. The model does not merely change
concentration slightly: it can move the same signaling network from a strongly
amplified front/back response toward a weak graded response, or equivalently
raise the Rac-GEF gradient needed to polarize.

A concise interpretation is therefore:

> **If the membrane/cytosol partitioning logic of Holmes et al. applies across
> WGD-associated cell sizes, maintaining total Rac/Cdc42/Rho concentration is
> not sufficient to preserve polarity sensitivity. Larger cells can contain
> proportionally more inaccessible cytosolic GTPase and require a stronger
> spatial cue to establish the same front/back state.**

# 9. What is published and what is extrapolated?

## Directly from Holmes et al.

- Rac, Cdc42, and Rho are represented as active membrane forms plus composite
  inactive membrane/cytosol pools.
- GEF activation acts only on the membrane-accessible fraction of inactive
  GTPase.
- The membrane fraction depends on cell geometry through
  \(c=k_{\rm on}/[k_{\rm on}+(V/L)k_{\rm off}]\).
- Increasing membrane accessibility sensitizes the polarity system.
- In their HeLa geometry, cell elongation/flattening raises surface-area:volume
  and can remove hysteresis and stabilize polarization.
- Their Figure 8 geometry analysis uses Model 4 without PI feedback and
  \(a=0.55~{\rm s^{-1}}\).

## Added here

- The cell is enlarged isotropically from 1x to 2x volume.
- Total GTPase **concentrations**, rather than total molecular amounts, are held
  fixed between the two established cell states.
- The unknown baseline membrane fraction is treated as a sensitivity parameter.
- A fixed-mean Rac-GEF gradient is applied to both sizes.
- Rac contrast of 0.4 is used as an operational strong-polarity threshold.

These distinctions matter. The report demonstrates a plausible consequence of
the published geometry mechanism, not a quantitatively validated prediction
that WGD doubles or triples the Rac-gradient threshold in a real cancer cell.

# 10. Limitations

1. **One-dimensional cell.** The Holmes model was designed for HeLa cells
   confined in narrow channels. A freely migrating mammalian cell has a much
   more complex 2D/3D shape.
2. **Unknown membrane cycling constants.** This is the largest uncertainty.
   Holmes et al. explicitly did not assign unique values to \(k_{\rm on}\),
   \(k_{\rm off}\), or the baseline membrane fraction.
3. **No phosphoinositide feedback here.** We deliberately set \(f_1=0\) to
   isolate the same geometry mechanism emphasized in their Figure 8.
4. **No cytoskeletal or adhesion feedback.** Rac/Cdc42/Rho polarity is treated
   as the endpoint rather than being coupled to actual migration.
5. **Isotropic WGD scaling is a null geometry.** Real WGD-associated size
   changes need not preserve shape.
6. **No experimental WGD calibration.** Protein partitioning, GEF/GAP
   abundance, membrane density, and Rac/Cdc42 activity have not been fitted to
   the relevant high- and low-ploidy cell lines.

# Export plot data

```{r export-data, include=FALSE}
data_dir <- "../docs/data"

if (dir.exists("../docs")) {
  dir.create(data_dir, recursive = TRUE, showWarnings = FALSE)

  write.csv(
    geometry_scaling,
    file.path(data_dir, "holmes_geometry_accessibility.csv"),
    row.names = FALSE
  )

  write.csv(
    geometry_summary,
    file.path(data_dir, "holmes_geometry_summary.csv"),
    row.names = FALSE
  )

  write.csv(
    decomposition_summary,
    file.path(data_dir, "holmes_size_effect_decomposition.csv"),
    row.names = FALSE
  )

  write.csv(
    example_profiles,
    file.path(data_dir, "holmes_example_polarity_profiles.csv"),
    row.names = FALSE
  )

  write.csv(
    gradient_sweep,
    file.path(data_dir, "holmes_gradient_response_sweep.csv"),
    row.names = FALSE
  )

  write.csv(
    threshold_wide,
    file.path(data_dir, "holmes_gradient_thresholds.csv"),
    row.names = FALSE
  )
}
```

# References

Holmes WR, Lin B, Levchenko A, Edelstein-Keshet L.
**Modelling Cell Polarization Driven by Synthetic Spatially Graded Rac
Activation.** *PLoS Computational Biology* 8:e1002366 (2012).
DOI: 10.1371/journal.pcbi.1002366.

Lin B, Holmes WR, Wang CJ, et al.
**Synthetic spatially graded Rac activation drives cell polarization and
movement.** *Proceedings of the National Academy of Sciences USA*
109:E3668-E3677 (2012). DOI: 10.1073/pnas.1210295109.

Jilkine A, Marée AFM, Edelstein-Keshet L.
**Mathematical model for spatial segregation of the Rho-family GTPases based
on inhibitory crosstalk.** *Bulletin of Mathematical Biology*
69:1943-1978 (2007).

Mori Y, Jilkine A, Edelstein-Keshet L.
**Wave-pinning and cell polarity from a bistable reaction-diffusion system.**
*Biophysical Journal* 94:3684-3697 (2008).
